### 2024-11-13 code/R/TimeVaryingDetection_mrmr.Rmd
initial commit

---
title: "TimeVaryingDetection_mrmr"
output: html_document
date: "2024-11-12"
---

**Here, I make a custom version of mrmr functons, aiming to get time-varying detection probability estimates. I updated the twostate.stan model to this end, and am now customizing mrmr's fit_model() function to use my updated stan model**

```{r}

library(cmdstanr)
library(mrmr)

```

**Write custom version of mrmr::fit_model()**

Change only line 53. Tell function to pull from my TVD_twostate.stan file

```{r}

# Custom fit function using the updated Stan model file
fit_model_custom <- function(
  data,
  seed = NULL,
  refresh = NULL,
  init = NULL,
  save_latent_dynamics = FALSE,
  output_dir = NULL,
  output_basename = NULL,
  sig_figs = NULL,
  chains = 4,
  parallel_chains = getOption("mc.cores", 1),
  chain_ids = seq_len(chains),
  threads_per_chain = NULL,
  opencl_ids = NULL,
  iter_warmup = 1000,
  iter_sampling = 1000,
  save_warmup = FALSE,
  thin = NULL,
  max_treedepth = NULL,
  adapt_engaged = TRUE,
  adapt_delta = NULL,
  step_size = NULL,
  metric = NULL,
  metric_file = NULL,
  inv_metric = NULL,
  init_buffer = NULL,
  term_buffer = NULL,
  window = NULL,
  fixed_param = FALSE,
  show_messages = TRUE,
  compile = TRUE,
  ...
) {
  # updated to use my custom stan model
  stan_file <- "updated_twostate.stan" # specify new path

  model <- cmdstan_model(stan_file, ...)

  m_fit <- model$sample(
    data = data$stan_d,
    seed = seed,
    refresh = refresh,
    init = init,
    save_latent_dynamics = save_latent_dynamics,
    output_dir = output_dir,
    output_basename = output_basename,
    sig_figs = sig_figs,
    chains = chains,
    parallel_chains = parallel_chains,
    chain_ids = chain_ids,
    threads_per_chain = threads_per_chain,
    opencl_ids = opencl_ids,
    iter_warmup = iter_warmup,
    iter_sampling = iter_sampling,
    save_warmup = save_warmup,
    thin = thin,
    max_treedepth = max_treedepth,
    adapt_engaged = adapt_engaged,
    adapt_delta = adapt_delta,
    step_size = step_size,
    metric = metric,
    metric_file = metric_file,
    inv_metric = inv_metric,
    init_buffer = init_buffer,
    term_buffer = term_buffer,
    window = window,
    fixed_param = fixed_param,
    show_messages = show_messages
  )

  # Load all the data and return the whole unserialized fit object
  m_fit$draws() # Do not specify variables or inc_warmup.
  try(m_fit$sampler_diagnostics(), silent = TRUE)
  try(m_fit$init(), silent = TRUE)
  try(m_fit$profiles(), silent = TRUE)

  list(m_fit = m_fit, data = data)
}

```

```{r}

# set paths for Trapdoor captures and surveys
captures <- read.csv("data/54188_capture.csv")  
surveys <- read.csv("data/54188_survey.csv")

data_TVD_model <- mrmr::clean_data(captures, surveys)
```

```{r}

# Fit the model using the custom function
fit_result <- fit_model_custom(data_TVD_model,
                                chains = 3, 
                                iter_sampling = 1000, 
                                iter_warmup = 400)

modTVD_summary <- fit_result$m_fit$summary()
View(modTVD_summary)

write.csv(modTVD_summary, "Trapdoor_custom_summary")

```

**Now, run the classic mrmr model on the Trapdoor data to compare with my version**

```{r}

# Fit the model using the custom function
Trapdoor1_mrmr <- mrmr::fit_model(data_TVD_model,
                                chains = 3, 
                                iter_sampling = 1000, 
                                iter_warmup = 400)

Trapdoor1_summary <- Trapdoor1_mrmr$m_fit$summary()
View(Trapdoor1_summary)

write.csv(Traodoor1_summary, "Trapdoor1_mrmr_summary.csv")

Trapdoor1_summary <- read.csv("Trapdoor1_mrmr_summary.csv")
View(trapdoor_summary)

```

### 2024-11-13 code/R/VaryingDetection_mrmr.Rmd
initial commit

--- 2024-11-13 code/R/TimeVaryingDetection_mrmr.Rmd
@@
 
-**Write custom version of mrmr::fit_model()**
-
-Change only line 53. Tell function to pull from my TVD_twostate.stan file
+Write custom version of mrmr::fit_model()
 
@@
 # set paths for Trapdoor captures and surveys
-captures <- read.csv("data/54188_capture.csv")  
-surveys <- read.csv("data/54188_survey.csv")
+captures <- "path/to/capture-example.csv"  
+surveys <- "path/to/survey-example.csv"
 
 data_TVD_model <- mrmr::clean_data(captures, surveys)
-```
-
-```{r}
 
 # Fit the model using the custom function
-fit_result <- fit_model_custom(data_TVD_model,
-                                chains = 3, 
-                                iter_sampling = 1000, 
-                                iter_warmup = 400)
-
-modTVD_summary <- fit_result$m_fit$summary()
-View(modTVD_summary)
-
-write.csv(modTVD_summary, "Trapdoor_custom_summary")
+fit_result <- fit_model_custom(data)
 
 ```
-
-**Now, run the classic mrmr model on the Trapdoor data to compare with my version**
-
-```{r}
-
-# Fit the model using the custom function
-Trapdoor1_mrmr <- mrmr::fit_model(data_TVD_model,
-                                chains = 3, 
-                                iter_sampling = 1000, 
-                                iter_warmup = 400)
-
-Trapdoor1_summary <- Trapdoor1_mrmr$m_fit$summary()
-View(Trapdoor1_summary)
-
-write.csv(Traodoor1_summary, "Trapdoor1_mrmr_summary.csv")
-
-Trapdoor1_summary <- read.csv("Trapdoor1_mrmr_summary.csv")
-View(trapdoor_summary)
-
-```

### 2024-11-13 code/R/mrmr_mod1.Rmd
initial commit

---
title: "Mossy_mrmr_model"
output: html_document
date: "2024-09-12"
---

Load packages

```{r}

library(dplyr)
library(readr)
library(tibble)
library(rstan)
library(mrmr)

```

Load capture data and format 'captures' data frame

```{r}

# Load .csv of all captures from the Mossy Pond CMR study period
captures <- read.csv("Capture_Data.csv")

# Remove "00:00:00 EDT" from Date column for compatibility with mrmr 
captures$Date <- gsub("00:00:00 EDT","", captures$Date)

# Convert 'Date' variable to date object class 
captures$Date <- as.Date(captures$Date, format = "%a %b %d  %Y")

# Change variable names for compatibility with mrmr 
names(captures)[names(captures) == "Date"] <- "survey_date"
names(captures)[names(captures) == "PIT_TagCode"] <- "pit_tag_id"

# Change data types for compatibility with mrmr
captures$survey_date <- as.character(captures$survey_date)
captures$pit_tag_id <- as.character(captures$pit_tag_id)

# Remove the single dead frog observation (instead, use it to create a 'removals' data frame)
captures <- captures[captures$frog_state != "dead", ]

```

Load survey date data and format 'surveys' data frame

```{r}

# load surveys .csv I made externally (manually in excel...)
surveys <- read.csv("survey.csv")

# change data types for compatibility with mrmr
surveys$survey_date <- as.character(surveys$survey_date)
surveys$primary_period <- as.integer(surveys$primary_period)
surveys$secondary_period <- as.integer(surveys$secondary_period)

```

Load capture data and format 'removals' data frame

```{r}

removals <- read.csv("Capture_Data.csv")

removals <- removals[removals$frog_state == "dead", ]

# Remove "00:00:00 EDT" from Date column for compatibility with 'mrmr' 
removals$Date <- gsub("00:00:00 EDT","", removals$Date)

# Convert 'Date' variable to object class date
removals$Date <- as.Date(removals$Date, format = "%a %b %d  %Y")

# Change variable names for compatibility with 'mrmr' 
names(removals)[names(removals) == "Date"] <- "removal_date"
names(removals)[names(removals) == "PIT_TagCode"] <- "pit_tag_id"

removals$survey_date <- as.character(removals$removal_date)
removals$pit_tag_id <- as.character(removals$pit_tag_id)

```

Run the clean_data function and fit the model

```{r}

# run the clean_data function
mrmr_mossy_mod1_data <- clean_data(captures, surveys, removals = removals)

# fit the model
mrmr_mossy_mod1 <- fit_model(mrmr_mossy_mod1_data,
                                chains = 3, 
                                iter_sampling = 1000, 
                                iter_warmup = 400)

# save the model fit summary to an object
mod1_summary <- mrmr_mossy_mod1$m_fit$summary()
view(mod1_summary)

```

### 2024-11-13 code/R/mrmr_mod1_mossy.Rmd
initial commit

--- 2024-11-13 code/R/mrmr_mod1.Rmd
@@
 ---
-title: "Mossy_mrmr_model"
-output: html_document
-date: "2024-09-12"
+title: "Mossy Pond mrmr model 1"
+output: GitHub_document
+date: "`r Sys.Date()`"
+author: "Your Name"
 ---
@@
 captures <- captures[captures$frog_state != "dead", ]
+
+# ATTEMPT TO FIGURE THE PROBLEM OUT
+# Include only a subset of columns from captures. This seems to have worked in the past. 
+### captures <- unique(captures[ , c(2,3,5,7)])
+  # when I create this subset, 5 observations disappear
+  # This subset didn't work. Lets try getting rid of CA lake id column
+captures <- unique(captures[ , c(3,5,7)])
+  # another 2 observations disappeared
+# I have not yet run this model
 
@@
 
-Run the clean_data function and fit the model
+Run the clean_data function
 
@@
 # run the clean_data function
-mrmr_mossy_mod1_data <- clean_data(captures, surveys, removals = removals)
+mrmr_mossy_mod1_data <- clean_data(captures, surveys)
+
+```
+
+fit the model
+
+```{r}
 
@@
 ```
+
+```{r}
+
+plot_model(mrmr_mossy_mod1, "abundance")
+
+```

### 2024-11-13 code/R/TimeVaryingDetection_mrmr.Rmd
document formatting updates

@@
 
-**Here, I make a custom version of mrmr functons, aiming to get time-varying detection probability estimates.
-I updated the twostate.stan model to this end, and am now customizing mrmr's fit_model() function to use my updated stan model**
+# Custom version of MLRG's mrmr package for mark-recapture data analysis
+
+-   fit_model_custom() is a copy of mrmr's fit_model() function, changed only to call on a modified version of the underlying stan model.
+
+    The modified stan model...
+
+    -   allows for independent session-varying estimates of frog detection probabilities
+    -   includes more parameter estimates in the generated quantities block
+        -   p (detection probability, varying by session)
+        -   phi_est (phi, or survival probability, estimated for the whole population between primary periods)
+
+**Load packages**
 
 ```{r}
-
 library(cmdstanr)
 library(mrmr)
-
 ```
 
-**Write custom version of mrmr::fit_model()**
-
-Change only line 53. Tell function to pull from my TVD_twostate.stan file
+**Custom version of mrmr::fit_model()**
 
@@
   # updated to use my custom stan model
-  stan_file <- "updated_twostate.stan" # specify new path
+  stan_file <- "JI_custom_twostate.stan" # specify local path
 

### 2024-11-13 code/R/TimeVaryingDetection_mrmr.Rmd
plot and visually compare model output

@@
 
-# Custom version of MLRG's mrmr package for mark-recapture data analysis
-
--   fit_model_custom() is a copy of mrmr's fit_model() function, changed only to call on a modified version of the underlying stan model.
+# Test my custom stan model with MLRG's public data from site 54188
 
@@
 **Load packages**
-
 ```{r}
@@
 library(mrmr)
+library(ggplot2)
+library(ggpubr)
 ```
 
-**Custom version of mrmr::fit_model()**
+## Modify mrmr::fit_model()
+
+-   fit_model_custom() is a copy of mrmr::fit_model(), changed only to call on a modified version of the underlying stan model.
 
@@
   # updated to use my custom stan model
-  stan_file <- "JI_custom_twostate.stan" # specify local path
+  stan_file <- "/Users/johnimperato/Desktop/Mossy-Pond/code/stan/JI_custom_twostate.stan" # specify local path
 
@@
 
+## Run 54188 data through my custom model
+
+**Load and clean data**
 ```{r}
+# set paths for 54188 captures and surveys
+captures <- read.csv("/Users/johnimperato/Desktop/Mossy-Pond/data/54188_capture.csv")  
+surveys <- read.csv("/Users/johnimperato/Desktop/Mossy-Pond/data/54188_survey.csv")
 
-# set paths for Trapdoor captures and surveys
-captures <- read.csv("data/54188_capture.csv")  
-surveys <- read.csv("data/54188_survey.csv")
-
-data_TVD_model <- mrmr::clean_data(captures, surveys)
+# run mrmr::clean_data()
+data_54188 <- mrmr::clean_data(captures, surveys)
 ```
@@
 ```{r}
-
-# Fit the model using the custom function
-fit_result <- fit_model_custom(data_TVD_model,
+# Fit the custom model using the modified function
+custom_mod_54188 <- fit_model_custom(data_custom_mod_54188,
                                 chains = 3, 
@@
 
-modTVD_summary <- fit_result$m_fit$summary()
-View(modTVD_summary)
+custom_mod_54188_summary <- custom_mod_54188$m_fit$summary()
+View(custom_mod_54188_summary)
 
-write.csv(modTVD_summary, "Trapdoor_custom_summary")
+write.csv(custom_mod_54188_summary, "54188_custom_mod_summary")
 
@@
 
-**Now, run the classic mrmr model on the Trapdoor data to compare with my version**
+**Plot abundance and recruitment**
+```{r}
+
+trapdoor_custom_abundance_plot <- mrmr::plot_model(custom_mod_54188, "abundance")
+trapdoor_custom_recruitment_plot <- mrmr::plot_model(custom_mod_54188, "recruitment")
+
+# arrange plots together
+mrmr_custom_plots_54188 <- ggarrange(trapdoor_custom_abundance_plot, trapdoor_custom_recruitment_plot, ncol =1)
+
+# add title
+custom_annotated <- annotate_figure(mrmr_custom_plots_54188, top = text_grob("Time-Varying Detection mrmr", size = 14, face = "bold"))
+
+```
+
+## Run classic mrmr model on 54188 data
 
@@
 
-# Fit the model using the custom function
-Trapdoor1_mrmr <- mrmr::fit_model(data_TVD_model,
+# Fit the model using mrmr's built in function
+classic_mod_54188 <- mrmr::fit_model(data_54188,
                                 chains = 3, 
@@
 
-Trapdoor1_summary <- Trapdoor1_mrmr$m_fit$summary()
-View(Trapdoor1_summary)
+classic_mod_54188_summary <- classic_mod_54188$m_fit$summary()
+View(classic_mod_54188_summary)
 
-write.csv(Traodoor1_summary, "Trapdoor1_mrmr_summary.csv")
+write.csv(classic_mod_54188_summary, "54188_classic_mod_summary.csv")
 
-Trapdoor1_summary <- read.csv("Trapdoor1_mrmr_summary.csv")
-View(trapdoor_summary)
+plot_model(classic_mod_54188, "abundance")
+plot_model(classic_mod_54188, "recruitment")
 
 ```
+
+**Plot abundance and recruitment**
+```{r}
+
+trapdoor_classic_abundance_plot <- mrmr::plot_model(classic_mod_54188, "abundance")
+trapdoor_classic_recruitment_plot <- mrmr::plot_model(classic_mod_54188, "recruitment")
+
+mrmr_classic_plots_54188 <- ggarrange(trapdoor_classic_abundance_plot, trapdoor_classic_recruitment_plot, ncol =1)
+
+# Add title
+classic_annotated <- annotate_figure(mrmr_classic_plots_54188, top = text_grob("Classic mrmr", size = 14, face = "bold"))
+
+```
+
+## Compare Model Output
+
+**Visual Comparison**
+```{r}
+
+abund_recruit_comparison_plot <- ggarrange(classic_annotated, custom_annotated, ncol=2)
+
+```

### 2024-11-14 code/R/MossyPond_Classic_mrmr.Rmd
rename file

--- 2024-11-13 code/R/mrmr_mod1_mossy.Rmd
@@
 # Load .csv of all captures from the Mossy Pond CMR study period
-captures <- read.csv("Capture_Data.csv")
+captures <- read.csv("/Users/johnimperato/Documents/Capture_Data.csv")
 
@@
   # This subset didn't work. Lets try getting rid of CA lake id column
-captures <- unique(captures[ , c(3,5,7)])
+captures <- unique(captures[ , c(3,5)])
   # another 2 observations disappeared

### 2024-11-14 code/R/MP_captures_formatting.R
initial commit of captures data formatting

@@ -1,0 +1,33 @@
+# Load capture data and format 'captures' data frame for mrmr model
+
+# Load .csv of all captures from the Mossy Pond CMR study period
+captures <- read.csv("/Users/johnimperato/Documents/Capture_Data.csv")
+
+# Remove "00:00:00 EDT" from Date column for compatibility with mrmr 
+captures$Date <- gsub("00:00:00 EDT","", captures$Date)
+
+# Convert 'Date' variable to date object class 
+captures$Date <- as.Date(captures$Date, format = "%a %b %d  %Y")
+
+# Change variable names for compatibility with mrmr 
+names(captures)[names(captures) == "Date"] <- "survey_date"
+names(captures)[names(captures) == "PIT_TagCode"] <- "pit_tag_id"
+
+# Change data types for compatibility with mrmr
+captures$survey_date <- as.character(captures$survey_date)
+captures$pit_tag_id <- as.character(captures$pit_tag_id)
+
+# Remove the single dead frog observation (instead, use it to create a 'removals' data frame)
+captures <- captures[captures$frog_state != "dead", ]
+
+# ATTEMPT TO FIGURE THE PROBLEM OUT
+# Include only a subset of columns from captures. This seems to have worked in the past. 
+### captures <- unique(captures[ , c(2,3,5,7)])
+# when I create this subset, 5 observations disappear
+# This subset didn't work. Lets try getting rid of CA lake id column
+captures <- unique(captures[ , c(3,5)])
+# another 2 observations disappeared
+# I have not yet run this model
+
+write.csv( captures , '/Users/johnimperato/Desktop/Mossy-Pond_companion/MP_captures.csv' )
+

### 2024-11-14 code/R/custom_fit_model_function.R
initial commit of custom fit_model() function derived from mrmr

@@ -1,0 +1,78 @@
+fit_model_custom <- function(
+    data,
+    seed = NULL,
+    refresh = NULL,
+    init = NULL,
+    save_latent_dynamics = FALSE,
+    output_dir = NULL,
+    output_basename = NULL,
+    sig_figs = NULL,
+    chains = 4,
+    parallel_chains = getOption("mc.cores", 1),
+    chain_ids = seq_len(chains),
+    threads_per_chain = NULL,
+    opencl_ids = NULL,
+    iter_warmup = 1000,
+    iter_sampling = 1000,
+    save_warmup = FALSE,
+    thin = NULL,
+    max_treedepth = NULL,
+    adapt_engaged = TRUE,
+    adapt_delta = NULL,
+    step_size = NULL,
+    metric = NULL,
+    metric_file = NULL,
+    inv_metric = NULL,
+    init_buffer = NULL,
+    term_buffer = NULL,
+    window = NULL,
+    fixed_param = FALSE,
+    show_messages = TRUE,
+    compile = TRUE,
+    ...
+) {
+  # updated to use my custom stan model
+  stan_file <- "/Users/johnimperato/Desktop/Mossy-Pond/code/stan/JI_custom_twostate.stan" # specify local path
+  
+  model <- cmdstan_model(stan_file, ...)
+  
+  m_fit <- model$sample(
+    data = data$stan_d,
+    seed = seed,
+    refresh = refresh,
+    init = init,
+    save_latent_dynamics = save_latent_dynamics,
+    output_dir = output_dir,
+    output_basename = output_basename,
+    sig_figs = sig_figs,
+    chains = chains,
+    parallel_chains = parallel_chains,
+    chain_ids = chain_ids,
+    threads_per_chain = threads_per_chain,
+    opencl_ids = opencl_ids,
+    iter_warmup = iter_warmup,
+    iter_sampling = iter_sampling,
+    save_warmup = save_warmup,
+    thin = thin,
+    max_treedepth = max_treedepth,
+    adapt_engaged = adapt_engaged,
+    adapt_delta = adapt_delta,
+    step_size = step_size,
+    metric = metric,
+    metric_file = metric_file,
+    inv_metric = inv_metric,
+    init_buffer = init_buffer,
+    term_buffer = term_buffer,
+    window = window,
+    fixed_param = fixed_param,
+    show_messages = show_messages
+  )
+  
[…]

### 2024-11-14 code/R/MossyPond_Classic_mrmr.Rmd
remove data formatting code and arrange output plots

--- 2024-11-13 code/R/mrmr_mod1.Rmd
@@
 ---
-title: "Mossy_mrmr_model"
-output: html_document
-date: "2024-09-12"
+title: "Mossy Pond mrmr model 1"
+output: GitHub_document
+date: "`r Sys.Date()`"
+author: "Your Name"
 ---
 
-Load packages
-
 ```{r}
-
 library(dplyr)
@@
 library(mrmr)
-
+library(ggpubr)
 ```
-
-Load capture data and format 'captures' data frame
 
@@
 
-# Load .csv of all captures from the Mossy Pond CMR study period
-captures <- read.csv("Capture_Data.csv")
+# load data
+captures <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond_companion/MP_captures.csv')
+surveys <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond/data/survey_MossyPond.csv')
 
-# Remove "00:00:00 EDT" from Date column for compatibility with mrmr 
-captures$Date <- gsub("00:00:00 EDT","", captures$Date)
+# clean data 
+mrmr_mossy_mod1_data <- clean_data(captures, surveys)
 
-# Convert 'Date' variable to date object class 
-captures$Date <- as.Date(captures$Date, format = "%a %b %d  %Y")
-
-# Change variable names for compatibility with mrmr 
-names(captures)[names(captures) == "Date"] <- "survey_date"
-names(captures)[names(captures) == "PIT_TagCode"] <- "pit_tag_id"
-
-# Change data types for compatibility with mrmr
-captures$survey_date <- as.character(captures$survey_date)
-captures$pit_tag_id <- as.character(captures$pit_tag_id)
-
-# Remove the single dead frog observation (instead, use it to create a 'removals' data frame)
-captures <- captures[captures$frog_state != "dead", ]
-
-```
-
-Load survey date data and format 'surveys' data frame
-
-```{r}
-
-# load surveys .csv I made externally (manually in excel...)
-surveys <- read.csv("survey.csv")
-
-# change data types for compatibility with mrmr
-surveys$survey_date <- as.character(surveys$survey_date)
-surveys$primary_period <- as.integer(surveys$primary_period)
-surveys$secondary_period <- as.integer(surveys$secondary_period)
-
-```
-
-Load capture data and format 'removals' data frame
-
-```{r}
-
-removals <- read.csv("Capture_Data.csv")
-
-removals <- removals[removals$frog_state == "dead", ]
-
-# Remove "00:00:00 EDT" from Date column for compatibility with 'mrmr' 
-removals$Date <- gsub("00:00:00 EDT","", removals$Date)
-
-# Convert 'Date' variable to object class date
-removals$Date <- as.Date(removals$Date, format = "%a %b %d  %Y")
-
-# Change variable names for compatibility with 'mrmr' 
-names(removals)[names(removals) == "Date"] <- "removal_date"
-names(removals)[names(removals) == "PIT_TagCode"] <- "pit_tag_id"
-
-removals$survey_date <- as.character(removals$removal_date)
-removals$pit_tag_id <- as.character(removals$pit_tag_id)
-
-```
-
-Run the clean_data function and fit the model
-
-```{r}
-
-# run the clean_data function
-mrmr_mossy_mod1_data <- clean_data(captures, surveys, removals = removals)
-
-# fit the model
+# fit classic mrmr model
 mrmr_mossy_mod1 <- fit_model(mrmr_mossy_mod1_data,
@@
 
-# save the model fit summary to an object
+# save model summary to an object
 mod1_summary <- mrmr_mossy_mod1$m_fit$summary()
@@
 
+write.csv(mod1_summary, "/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_classic_mod_sumary.csv")
+
 ```
+
+```{r}
+
+abundance_MP_classic <- plot_model(mrmr_mossy_mod1, "abundance")
+recruitment_MP_classic <- plot_model(mrmr_mossy_mod1, "recruitment")
+
+MP_classic_plots <- ggarrange(abundance_MP_classic, recruitment_MP_classic, ncol=1)
+MP_classic_plots
+
+```

### 2024-11-16 code/R/MossyPond_Custom_mrmr_TVD.Rmd
rename file

---
title: "Mossy_custom_mrmr"
output: html_document
date: "2024-11-13"
---

# Run my custom mrmr model with Mossy Pond data

```{r}
library(mrmr)
```

**Load and format data**

```{r}
getwd()
surveys <- read.csv("/Users/johnimperato/Desktop/Mossy-Pond/data/survey_MossyPond.csv")
captures <- read.csv("/Users/johnimperato/Documents/private_MP_data/captures_MP.csv")

mp_custom_data <- mrmr::clean_data(captures, surveys)

```

**Fit the custom model to MP data and summarize**

```{r}

# Fit the custom model using the modified function
custom_mod_MP <- fit_model_custom(mp_custom_data,
                                chains = 3, 
                                iter_sampling = 1000, 
                                iter_warmup = 400)

custom_mod_mp_summary <- custom_mod_MP$m_fit$summary()
View(custom_mod_mp_summary)

write.csv(custom_mod_mp_summary, "/Users/johnimperato/Documents/mp_custom_mod_summary")

```

```{r}

abundance_MP_custom <- mrmr::plot_model(custom_mod_MP, "abundance")
recruitment_MP_custom <- mrmr::plot_model(custom_mod_MP, "recruitment")

```

### 2024-11-16 code/R/54188_ClassicCustomComparison_mrmr.Rmd
remove custom fit model function code

--- 2024-11-13 code/R/TimeVaryingDetection_mrmr.Rmd
@@
 
-# Test my custom stan model with MLRG's public data from site 54188
+# Model Comparison
 
-    The modified stan model...
+**Classic mrmr model vs. modified mrmr model**
+
+    The modified model...
 
@@
 library(mrmr)
-library(ggplot2)
 library(ggpubr)
-```
-
-## Modify mrmr::fit_model()
-
--   fit_model_custom() is a copy of mrmr::fit_model(), changed only to call on a modified version of the underlying stan model.
-
-```{r}
-
-# Custom fit function using the updated Stan model file
-fit_model_custom <- function(
-  data,
-  seed = NULL,
-  refresh = NULL,
-  init = NULL,
-  save_latent_dynamics = FALSE,
-  output_dir = NULL,
-  output_basename = NULL,
-  sig_figs = NULL,
-  chains = 4,
-  parallel_chains = getOption("mc.cores", 1),
-  chain_ids = seq_len(chains),
-  threads_per_chain = NULL,
-  opencl_ids = NULL,
-  iter_warmup = 1000,
-  iter_sampling = 1000,
-  save_warmup = FALSE,
-  thin = NULL,
-  max_treedepth = NULL,
-  adapt_engaged = TRUE,
-  adapt_delta = NULL,
-  step_size = NULL,
-  metric = NULL,
-  metric_file = NULL,
-  inv_metric = NULL,
-  init_buffer = NULL,
-  term_buffer = NULL,
-  window = NULL,
-  fixed_param = FALSE,
-  show_messages = TRUE,
-  compile = TRUE,
-  ...
-) {
-  # updated to use my custom stan model
-  stan_file <- "/Users/johnimperato/Desktop/Mossy-Pond/code/stan/JI_custom_twostate.stan" # specify local path
-
-  model <- cmdstan_model(stan_file, ...)
-
-  m_fit <- model$sample(
-    data = data$stan_d,
-    seed = seed,
-    refresh = refresh,
-    init = init,
-    save_latent_dynamics = save_latent_dynamics,
-    output_dir = output_dir,
-    output_basename = output_basename,
-    sig_figs = sig_figs,
-    chains = chains,
-    parallel_chains = parallel_chains,
-    chain_ids = chain_ids,
-    threads_per_chain = threads_per_chain,
-    opencl_ids = opencl_ids,
-    iter_warmup = iter_warmup,
-    iter_sampling = iter_sampling,
-    save_warmup = save_warmup,
-    thin = thin,
-    max_treedepth = max_treedepth,
-    adapt_engaged = adapt_engaged,
-    adapt_delta = adapt_delta,
-    step_size = step_size,
-    metric = metric,
-    metric_file = metric_file,
-    inv_metric = inv_metric,
-    init_buffer = init_buffer,
-    term_buffer = term_buffer,
-    window = window,
-    fixed_param = fixed_param,
-    show_messages = show_messages
-  )
-
-  # Load all the data and return the whole unserialized fit object
-  m_fit$draws() # Do not specify variables or inc_warmup.
-  try(m_fit$sampler_diagnostics(), silent = TRUE)
-  try(m_fit$init(), silent = TRUE)
-  try(m_fit$profiles(), silent = TRUE)
-
-  list(m_fit = m_fit, data = data)
-}
-
 ```
@@
 
-abund_recruit_comparison_plot <- ggarrange(classic_annotated, custom_annotated, ncol=2)
+abund_recruit_comparison_plot_54188 <- ggarrange(classic_annotated, custom_annotated, ncol=2)
 

### 2024-11-16 code/R/54188_ClassicCustomComparison_mrmr.Rmd
update formatting and clean up code

@@
 
-**Classic mrmr model vs. modified mrmr model**
+**Classic mrmr model vs. custom mrmr model**
 
-    The modified model...
+The custom model...
 
@@
 **Load packages**
+
 ```{r}
@@
 
-## Run 54188 data through my custom model
+## Run 54188 data through custom model
 
 **Load and clean data**
+
 ```{r}
@@
 
+**Fit the custom model**
+
 ```{r}
 # Fit the custom model using the modified function
-custom_mod_54188 <- fit_model_custom(data_custom_mod_54188,
+custom_mod_54188 <- fit_model_custom(data_54188,
                                 chains = 3, 
@@
 
+# create and view model summary
 custom_mod_54188_summary <- custom_mod_54188$m_fit$summary()
@@
 
+# save results to .csv file
 write.csv(custom_mod_54188_summary, "54188_custom_mod_summary")
@@
 **Plot abundance and recruitment**
+
 ```{r}
@@
 **Visual Comparison**
+
 ```{r}

### 2024-11-20 code/R/classic_mrmr_loglik_fit_model().R
initial commit of custom fit_model function for mrmr with log_lik

@@ -1,0 +1,78 @@
+fit_model_classic_loglik <- function(
+    data,
+    seed = NULL,
+    refresh = NULL,
+    init = NULL,
+    save_latent_dynamics = FALSE,
+    output_dir = NULL,
+    output_basename = NULL,
+    sig_figs = NULL,
+    chains = 4,
+    parallel_chains = getOption("mc.cores", 1),
+    chain_ids = seq_len(chains),
+    threads_per_chain = NULL,
+    opencl_ids = NULL,
+    iter_warmup = 1000,
+    iter_sampling = 1000,
+    save_warmup = FALSE,
+    thin = NULL,
+    max_treedepth = NULL,
+    adapt_engaged = TRUE,
+    adapt_delta = NULL,
+    step_size = NULL,
+    metric = NULL,
+    metric_file = NULL,
+    inv_metric = NULL,
+    init_buffer = NULL,
+    term_buffer = NULL,
+    window = NULL,
+    fixed_param = FALSE,
+    show_messages = TRUE,
+    compile = TRUE,
+    ...
+) {
+  # updated to use my custom stan model
+  stan_file <- "/Users/johnimperato/Desktop/Mossy-Pond/code/stan/twostate_loglik_modified.stan" # specify local path
+  
+  model <- cmdstan_model(stan_file, ...)
+  
+  m_fit <- model$sample(
+    data = data$stan_d,
+    seed = seed,
+    refresh = refresh,
+    init = init,
+    save_latent_dynamics = save_latent_dynamics,
+    output_dir = output_dir,
+    output_basename = output_basename,
+    sig_figs = sig_figs,
+    chains = chains,
+    parallel_chains = parallel_chains,
+    chain_ids = chain_ids,
+    threads_per_chain = threads_per_chain,
+    opencl_ids = opencl_ids,
+    iter_warmup = iter_warmup,
+    iter_sampling = iter_sampling,
+    save_warmup = save_warmup,
+    thin = thin,
+    max_treedepth = max_treedepth,
+    adapt_engaged = adapt_engaged,
+    adapt_delta = adapt_delta,
+    step_size = step_size,
+    metric = metric,
+    metric_file = metric_file,
+    inv_metric = inv_metric,
+    init_buffer = init_buffer,
+    term_buffer = term_buffer,
+    window = window,
+    fixed_param = fixed_param,
+    show_messages = show_messages
+  )
+  
[…]

### 2024-11-21 code/R/54188_ClassicCustomComparison_mrmr.Rmd
Update formatting and add p and Phi parameter estimates to comparison table

@@
 
-# Model Comparison
+# Mrmr Model Comparison
 
-**Classic mrmr model vs. custom mrmr model**
+**Classic mrmr model** (fixed detection probability)
 
-The custom model...
+vs.
 
--   allows for independent session-varying estimates of frog detection probabilities
--   includes more parameter estimates in the generated quantities block
-    -   p (detection probability, varying by session)
-    -   phi_est (phi, or survival probability, estimated for the whole population between primary periods)
+**Custom mrmr model** (independent, time-varying detection probabilities)
 
-**Load packages**
+## Introduction
+
+### Generated Quantities Block
+
+Both the classic and custom models include changes to the generated quantities block of mrmr's underlying Stan model.
+The generated quantities block is composed of post-hoc computations based on the model's parameters after it has finished fitting the data.
+Changes to this block do not affect how the model runs or the estimated parameter values.
+The generated quantities blocks of both the custom and classic model are identical.
+Generated quantities include:
+
+1.  N - total abundance, estimated for each primary period. Included in the standard mrmr model.
+
+2.  B - number of adults recruited into the population, estimated for each primary period. Included in the standard mrmr model.
+
+3.
+p - detection probability, estimated for each session (secondary period).
+In the classic model, p is fixed so the estimate is identical for each session.
+In the custom model, p is estimated independently for each session (time-varying detection probabilities).
+
+4.
+phi_est - survival probability, estimated between primary periods for the whole population.
+phi_est[1] is the collective estimated survival probability between the first and second primary periods, phi_est[2] between the second and third, and so on.\*
+
+    -   \*I need to confirm my suspicion that the model generates "boundary estimates" for phi (an estimate before the first primary period, and after the last primary period).
+I am currently operating under this assumption.
+If I can't confirm this by myself, maybe I can ask Max.
+
+5.
+log_lik - log of the likelihood, calculated for each individual, real and pseudo.
+Likelihood is the probability of observing an individual's capture history given the model parameters.
+Pseudo-individuals don't actually have a capture histories, so their log_lik values are all identical.
+See the 'Compare Model Fit' section of this report for more information about this quantity.
+
+### Other Blocks...
+
+The custom mrmr model calls upon a Stan file that is modified from the classic mrmr Stan file. Compared to the classic, the modified Stan file...
+
+1.  allows for independent, session-varying estimates of frog detection probabilities
+
+### The data
+
+This analysis uses capture-mark-recapture (CMR) data for a population of endangered Sierra Nevada yellow-legged frogs (*Rana sierrae*) in the Sierra National Forest, California.
+This population is referred to by the numeric site id of its associated water body, 54188.
+The data set is publicly available from the Mountain Lakes Research Group.
+
+## Run Models
+
+Load packages
 
@@
 library(ggpubr)
+library(dplyr)
 ```
 
-## Run 54188 data through custom model
+### Prepare data
 
-**Load and clean data**
+Load data for population 54188 and format it with mrmr's clean_data( ) function.
 
 ```{r}
-# set paths for 54188 captures and surveys
+# set local paths for 54188 captures and surveys
 captures <- read.csv("/Users/johnimperato/Desktop/Mossy-Pond/data/54188_capture.csv")  
@@
 
-# run mrmr::clean_data()
+# clean data
 data_54188 <- mrmr::clean_data(captures, surveys)
@@
 
-**Fit the custom model**
+### Fit the Custom Model
 
@@
 
+with mrmr's built-in plotting functionality.
+
 ```{r}
-
 trapdoor_custom_abundance_plot <- mrmr::plot_model(custom_mod_54188, "abundance")
@@
 
-# arrange plots together
+# arrange plots 
 mrmr_custom_plots_54188 <- ggarrange(trapdoor_custom_abundance_plot, trapdoor_custom_recruitment_plot, ncol =1)
@@
 custom_annotated <- annotate_figure(mrmr_custom_plots_54188, top = text_grob("Time-Varying Detection mrmr", size = 14, face = "bold"))
-
 ```
 
-## Run classic mrmr model on 54188 data
+### Fit the classic model
 
 ```{r}
-
 # Fit the model using mrmr's built in function
-classic_mod_54188 <- mrmr::fit_model(data_54188,
+classic_mod_54188 <- fit_model_classic_loglik(data_54188,
                                 chains = 3, 
@@
 
+# create and view summary
 classic_mod_54188_summary <- classic_mod_54188$m_fit$summary()
@@
 
+# save summary to .csv file
 write.csv(classic_mod_54188_summary, "54188_classic_mod_summary.csv")
-
-plot_model(classic_mod_54188, "abundance")
-plot_model(classic_mod_54188, "recruitment")
 
@@
 
+with mrmr's built-in plotting functionality.
+
 ```{r}
-
 trapdoor_classic_abundance_plot <- mrmr::plot_model(classic_mod_54188, "abundance")
@@
 
+# arrange plots
 mrmr_classic_plots_54188 <- ggarrange(trapdoor_classic_abundance_plot, trapdoor_classic_recruitment_plot, ncol =1)
 
-# Add title
+# add title
 classic_annotated <- annotate_figure(mrmr_classic_plots_54188, top = text_grob("Classic mrmr", size = 14, face = "bold"))
@@
 
-## Compare Model Output
+## Model Comparison
 
-**Visual Comparison**
+### Generated Quantities
+
+**Create a table comparing the mean generated quantity values for each model**
+
+This table shows **only the mean** estimated value for each generated quantity.
+It is meant to serve as a quick reference for comparing model output.
+See the individual model summaries for complete generated quantity summaries.
 
 ```{r}
+# pull quantities from model fit summary
+N_classic <- classic_mod_54188_summary[4216:4220, 1:2]
+N_custom <- custom_mod_54188_summary[4229:4233, 1:2]
+B_classic <- classic_mod_54188_summary[4221:4225, 1:2]
+B_custom <- custom_mod_54188_summary[4234:4238, 1:2]
+Phi_classic <- classic_mod_54188_summary[4240:4243, 1:2]
+Phi_custom <- custom_mod_54188_summary[4253:4256, 1:2]
+p_classic <- classic_mod_54188_summary[4226:4239, 1:2]
+p_custom <- custom_mod_54188_summary[4239:4252, 1:2]
 
-abund_recruit_comparison_plot_54188 <- ggarrange(classic_annotated, custom_annotated, ncol=2)
+GenQuant_comparison <- bind_rows(
+  left_join(N_classic, N_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "abundance"),
+  left_join(B_classic, B_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "recruitment"),
+  left_join(Phi_classic, Phi_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "survival"),
+  left_join(p_classic, p_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "detection"),
+) %>% 
+  select(variable, parameter, everything())
+
+print(GenQuant_comparison)
 
 ```
+
+**Visually compare abundance and recruitment estimates between models**
+
+Axes are not standardized, which limits comparison value.
+
+```{r}
+# visual comparison of abundance and recruitment plots
+abund_recruit_comparison_plot_54188 <- ggarrange(classic_annotated, custom_annotated, ncol=2)
+print(abund_recruit_comparison_plot_54188)
+```
+
+### Quantifed Model Fit
+
+Use leave-one-out (LOO) cross-validation, a standard method to calculate the predictive accuracy of Bayesian models, to compare how well the classic and custom models fit the Trapdoor data.
+The cmdstanr package has built in LOO capabilities.
+Briefly,
+
+-   LOO cross-validation uses log likelihood values, which now appear in the model summary as generated quantities, to evaluate how well the model predicts each data point when that point is left out of the fitting process.
+Likelihood, calculated at the individual level, is the probability of observing the data (in this case, an individual frog's capture history) given the model parameters.
+
+The loo( ) function generates an expected log predictive density (elpd) value for each model.
+The loo_compare( ) function ranks models based on their elpd values and calculates the difference in those values.
+A difference of 2-3x the standard error is generally considered significant in a LOO comparison.
+In this case, the elpd_diff value is 2.95x the standard error, suggesting that the custom mrmr model fits the Trapdoor data significantly better than the classic model.
+
+```{r}
+loo_classic <- classic_mod_54188$m_fit$loo()
+
+loo_custom <- custom_mod_54188$m_fit$loo()
+
+loo_compare <- loo::loo_compare(loo_classic, loo_custom)
+
+print(loo_compare)
+
+```

### 2024-11-21 code/R/54188_ClassicCustomComparison_mrmr.Rmd
add code to interpret LOO comparison

@@
 
-with mrmr's built-in plotting functionality.
+...with mrmr's built-in plotting functionality.
 
@@
 
-Axes are not standardized, which limits comparison value.
+Note that axes are not standardized.
 
@@
 The loo( ) function generates an expected log predictive density (elpd) value for each model.
-The loo_compare( ) function ranks models based on their elpd values and calculates the difference in those values.
+The loo_compare( ) function ranks models based on their elpd values, with the best fit listed first, and calculates the difference in those values.
 A difference of 2-3x the standard error is generally considered significant in a LOO comparison.
-In this case, the elpd_diff value is 2.95x the standard error, suggesting that the custom mrmr model fits the Trapdoor data significantly better than the classic model.
 
@@
 
+# extract elpd_diff and se_diff for model1
+elpd_diff_model1 <- loo_compare["model1", "elpd_diff"]
+se_diff_model1 <- loo_compare["model1", "se_diff"]
+
+# calculate significance ratio
+result <- elpd_diff_model1 / se_diff_model1
+
+# interpret
+if (abs(result) > 2) {
+  cat("Result of elpd_diff / se_diff for model1:", result, "\n")
+  cat("The custom model fits the data significantly better than the classic model.\n")
+} else {
+  cat("Result of elpd_diff / se_diff for model1:", result, "\n")
+  cat("The custom model does not fit the data significantly better than the classic model.\n")
+}
+
 ```

### 2024-11-21 code/R/54188_ClassicCustomComparison_mrmr.Rmd
update introduction

@@
 
+The R package mrmr serves as an interface to prepare data and run population models.
+It relies on a site-specific multi-state Bayesian open-population Jolly-Seber CMR model (hereafter referred to as the multi-state model) for the mathematical and computational part of model fitting.The multi-state model is written in the Stan programming language, which is designed for probabilistic modeling and Bayesian inference.
+The 'changes to mrmr' outlined in this report are primarily changes to the multi-state model, via modifications to the Stan code.
+The Stan code is not included in this report.
+Instead, changes are described using the English language.
+Snippets of essential Stan code are provided, when necessary, in *italics*.
+
+The multi-state model is composed of seven blocks:
+
+1.  Functions - handle repeated calculations. Key functions are forward_prob( ) and partial_sum_lpmf( ).
+
+2.  Data - specifies input data.
+
+3.  Transformed Data
+
+4.  Parameters - declared the parameters the model will estimate.
+
+5.  Transformed parameters
+
+6.  Model - defines the statistical model and specifies the priors.
+
+7.  Generated quantities - calculates quantities after the model is fit, used for interpretation and diagnostics.
+
+Changes made to each block are described below.
+
 ### Generated Quantities Block
@@
 
-1.  N - total abundance, estimated for each primary period. Included in the standard mrmr model.
+1.  N - total abundance, estimated for each primary period.
 
-2.  B - number of adults recruited into the population, estimated for each primary period. Included in the standard mrmr model.
+2.  B - number of adults recruited into the population, estimated for each primary period.
 
@@
 
-### Other Blocks...
+### Parameters Block
 
-The custom mrmr model calls upon a Stan file that is modified from the classic mrmr Stan file. Compared to the classic, the modified Stan file...
+The changes made to this block are essential to achieving independent, time-varying estimates of detection probability.
 
-1.  allows for independent, session-varying estimates of frog detection probabilities
+**beta_detect** a vector of parameters that represent the logit-scale detection probability for survey sessions.
+The classic mrmr model defines beta_detect as a vector of size m_detect.
+m_detect refers to the number of detection covariates.
+The custom model defines beta_detect as a vector of size Jtot.
+Jtot refers to the total number of survey sessions.
+
+Simply, the original line *vector[m_detect] beta_detect*; is changed to *vector[Jtot] beta_detect;* where each entry corresponds to a specific survey session.
+
+Meaning is assigned to beta_detect in the functions block via transformed parameter logit_detect, within the forward_prob( ) function. No changes were made to this function.
+
+### Transformed Parameters Block
+
+*logit_detect = X_detect \* beta_detect;* —\> *logit_detect = beta_detect;*
+
+In the classic model, logit_detect is derived from beta_detect using the design matrix x_detect.
+This ties detection probabilities to shared covariates.
+In the custom model, logit_detect is assigned directly to beta_detect, making each session's detection probability independent.
+\*
+
+-   When I want to introduce covariates into my analysis, I will need to revisit this section.
 
@@
 The loo_compare( ) function ranks models based on their elpd values, with the best fit listed first, and calculates the difference in those values.
-A difference of 2-3x the standard error is generally considered significant in a LOO comparison.
+A difference of more than 2x the standard error is generally considered significant in a LOO comparison.
 

### 2024-11-21 code/R/54188_ClassicCustomComparison_mrmr.Rmd
continue writing model block sections

@@
 Instead, changes are described using the English language.
-Snippets of essential Stan code are provided, when necessary, in *italics*.
+Snippets of essential Stan code are provided, when necessary, in *italics*, following the format (*classic model code*) —\> (*custom model code*).
 
@@
 
-4.  Parameters - declared the parameters the model will estimate.
+4.  **Parameters** - declared the parameters the model will estimate.
 
-5.  Transformed parameters
+5.  **Transformed parameters**
 
@@
 
-7.  Generated quantities - calculates quantities after the model is fit, used for interpretation and diagnostics.
+7.  **Generated quantities** - calculates quantities after the model is fit, used for interpretation and diagnostics.
 
-Changes made to each block are described below.
+This report focuses on changes made to three of those blocks: generates quantities, parameters, and transformed parameters. Those changes, and their functionality, are outlined below.
 
@@
 
-The changes made to this block are essential to achieving independent, time-varying estimates of detection probability.
+(*vector[m_detect] beta_detect;*) —\> (*vector[Jtot] beta_detect;*)
 
-**beta_detect** a vector of parameters that represent the logit-scale detection probability for survey sessions.
+This change is essential to achieving independent, time-varying estimates of detection probability in the custom model.
+
+**beta_detect** is a vector of parameters that represent the logit-scale detection probability for survey sessions.
 The classic mrmr model defines beta_detect as a vector of size m_detect.
@@
 Jtot refers to the total number of survey sessions.
+Each entry in the vector now corresponds to a specific survey session (custom) rather than a covariate (classic).
 
-Simply, the original line *vector[m_detect] beta_detect*; is changed to *vector[Jtot] beta_detect;* where each entry corresponds to a specific survey session.
-
-Meaning is assigned to beta_detect in the functions block via transformed parameter logit_detect, within the forward_prob( ) function. No changes were made to this function.
+Meaning is assigned to beta_detect in the functions block via the transformed parameter logit_detect, within the forward_prob( ) function. No changes were made to this function.
 
@@
 
-In the classic model, logit_detect is derived from beta_detect using the design matrix x_detect.
+In the classic model, logit_detect is derived from beta_detect using the design matrix of covariates x_detect.
 This ties detection probabilities to shared covariates.
 In the custom model, logit_detect is assigned directly to beta_detect, making each session's detection probability independent.
-\*
 
--   When I want to introduce covariates into my analysis, I will need to revisit this section.
+Changes to the parameters and transformed parameters blocks reflect a purpose to obtain the most accurate estimates of N, B, p, and phi for the collective population.
+The classic model seems to reflect a purpose to model detection and survival by covariates, which makes sense in the context of its original purpose, described by Knapp and Joseph (2018).
+If and when I want to introduce covariates into my analysis, I will need to revisit this section.
+(I'm not sure if it is possible to include both covariates in the design matrix and indepentent, session-specific detection.)
 
-### The data
+## The Data
 
@@
 
-## Run Models
+## Analysis
+
+This analysis serves to...
 
@@
 
-## Model Comparison
+### Model Comparison
 
-### Generated Quantities
+#### Generated Quantities
 
@@
 
-### Quantifed Model Fit
+#### Model Fit
 

### 2024-11-21 code/R/54188_ClassicCustomComparison_mrmr.Rmd
write Analysis section

@@
 
-*logit_detect = X_detect \* beta_detect;* —\> *logit_detect = beta_detect;*
+(*logit_detect = X_detect \* beta_detect;*) —\> (*logit_detect = beta_detect;*)
 
@@
 Changes to the parameters and transformed parameters blocks reflect a purpose to obtain the most accurate estimates of N, B, p, and phi for the collective population.
-The classic model seems to reflect a purpose to model detection and survival by covariates, which makes sense in the context of its original purpose, described by Knapp and Joseph (2018).
+The classic model reflect a purpose to model detection and survival by covariates, which makes sense in the context of Knapp and Joseph (2018).
 If and when I want to introduce covariates into my analysis, I will need to revisit this section.
@@
 
-This analysis serves to...
+This section describes the analysis performed to compare the classic and custom models.
+Both models were fit to the same capture-mark-recapture (CMR) data for the Sierra Nevada yellow-legged frog population at site 54188.
+Key comparisons include estimates of population abundance (N), recruitment (B), detection probability (p), and survival (phi).
+To evaluate model performance, I used the **leave-one-out cross-validation (LOO)** method, which assesses model fit while penalizing complexity.
 
-Load packages
+The goal of this analysis is to determine whether the custom model, with independent, time-varying estimates of detection probability, provides a significantly better fit to the data compared to the classic model, which assumes detection probability is constant across sessions.
+Differences in model performance will be interpreted using the LOO expected log predictive density (ELPD) values, along with associated uncertainty estimates.
+
+Results and interpretations for each parameter and the overall model comparison are detailed in the following subsections.
+
+**Load packages**
 

### 2024-11-21 code/R/54188_ClassicCustomComparison_mrmr.Rmd
include fit_model function sources

@@
 
-# Mrmr Model Comparison
+# mrmr Model Comparison
 
@@
 The 'changes to mrmr' outlined in this report are primarily changes to the multi-state model, via modifications to the Stan code.
-The Stan code is not included in this report.
+The Stan file is not included in this report.
 Instead, changes are described using the English language.
-Snippets of essential Stan code are provided, when necessary, in *italics*, following the format (*classic model code*) —\> (*custom model code*).
+Snippets of essential Stan code are provided, when necessary, in *italics*, following the format (*classic model Stan code*) —\> (*custom model Stan code*).
 
@@
 phi_est - survival probability, estimated between primary periods for the whole population.
-phi_est[1] is the collective estimated survival probability between the first and second primary periods, phi_est[2] between the second and third, and so on.\*
+phi_est[1] is the collective estimated survival probability between the first and second primary periods, phi_est[2] between the second and third, etc.\*
 
-    -   \*I need to confirm my suspicion that the model generates "boundary estimates" for phi (an estimate before the first primary period, and after the last primary period).
-I am currently operating under this assumption.
+    -   \*I need to confirm my assumption that the model generates "boundary estimates" for phi (an estimate before the first primary period, and after the last primary period).
 If I can't confirm this by myself, maybe I can ask Max.
@@
 This section describes the analysis performed to compare the classic and custom models.
-Both models were fit to the same capture-mark-recapture (CMR) data for the Sierra Nevada yellow-legged frog population at site 54188.
+Both models were fit to the same CMR data for the Sierra Nevada yellow-legged frog population at site 54188.
 Key comparisons include estimates of population abundance (N), recruitment (B), detection probability (p), and survival (phi).
-To evaluate model performance, I used the **leave-one-out cross-validation (LOO)** method, which assesses model fit while penalizing complexity.
+To evaluate model performance, I used the leave-one-out cross-validation (LOO) method, which assesses model fit while penalizing complexity.
 
@@
 
+The fit_model_custom( ) function is identical to mrmr's built in fit_model( ) function, except it draws on the custom .stan file on my local machine, rather than the standard mrmr .stan file.
+The custom .stan file includes the previously described changes to the parameters, transformed parameters, and generated quantities block.
+
 ```{r}
+# load the modified fit_model function
+source("custom_fit_model_function.R")
+
 # Fit the custom model using the modified function
@@
 
+The fit_model_classic( ) function is identical to mrmr's built in fit_model( ) function, except it draws on a modified .stan file on my local machine, rather than the standard mrmr .stan file.
+The modified .stan file includes the previously described changes to the generated quantities block.
+
 ```{r}
+source("classic_fit_model_function.R")
+
 # Fit the model using mrmr's built in function
-classic_mod_54188 <- fit_model_classic_loglik(data_54188,
+classic_mod_54188 <- fit_model_classic(data_54188,
                                 chains = 3, 

### 2024-11-21 code/R/54188_ClassicCustomComparison_mrmr.Rmd
edit introduction

@@
 
-# mrmr Model Comparison
+# mrmr Modifications
 
-**Classic mrmr model** (fixed detection probability)
+# and
 
-vs.
-
-**Custom mrmr model** (independent, time-varying detection probabilities)
+# Model Comparison
 
@@
 The R package mrmr serves as an interface to prepare data and run population models.
-It relies on a site-specific multi-state Bayesian open-population Jolly-Seber CMR model (hereafter referred to as the multi-state model) for the mathematical and computational part of model fitting.The multi-state model is written in the Stan programming language, which is designed for probabilistic modeling and Bayesian inference.
-The 'changes to mrmr' outlined in this report are primarily changes to the multi-state model, via modifications to the Stan code.
-The Stan file is not included in this report.
-Instead, changes are described using the English language.
-Snippets of essential Stan code are provided, when necessary, in *italics*, following the format (*classic model Stan code*) —\> (*custom model Stan code*).
+It relies on a site-specific multi-state Bayesian open-population Jolly-Seber CMR model (hereafter referred to as the multi-state model) for the mathematical and computational part of model fitting.
+The multi-state model is written in the Stan programming language, which is designed for probabilistic modeling and Bayesian inference.
 
-The multi-state model is composed of seven blocks:
+This report describes modifications to the multi-state model, each of which can be categorized as either 'cosmetic' or 'substantive.' Cosmetic changes are those that do not not affect how the model runs or the values of its parameter estimates.
+Substantive changes do affect how the model runs and the values of parameter estimates.
 
-1.  Functions - handle repeated calculations. Key functions are forward_prob( ) and partial_sum_lpmf( ).
+These modifications are embodied by a 'classic model' and a 'custom model.' The models are named relative to the original multi-state model published in the R package mrmr.
+The classic model includes only cosmetic changes relative to the original, and the custom model includes both cosmetic and substantive changes.
 
-2.  Data - specifies input data.
-
-3.  Transformed Data
-
-4.  **Parameters** - declared the parameters the model will estimate.
-
-5.  **Transformed parameters**
-
-6.  Model - defines the statistical model and specifies the priors.
-
-7.  **Generated quantities** - calculates quantities after the model is fit, used for interpretation and diagnostics.
-
-This report focuses on changes made to three of those blocks: generates quantities, parameters, and transformed parameters. Those changes, and their functionality, are outlined below.
+To understand the cosmetic and substantive nature of those changes, it is helpful to know that the multi-state Stan model is composed of seven **code blocks**: functions, data, transformed data, parameters, transformed parameters, model, and generated quantities.
+Changes made to the generated quantities block are cosmetic.
+Changes to any other block are substantive.
+This report details changes within the **generated quantities, parameters,** and **transformed parameters** blocks, as described in the next three sections.
 
@@
 The generated quantities block is composed of post-hoc computations based on the model's parameters after it has finished fitting the data.
-Changes to this block do not affect how the model runs or the estimated parameter values.
+Changes to this block are cosmetic and do not affect how the model runs or the estimated parameter values.
 The generated quantities blocks of both the custom and classic model are identical.
@@
 
-    -   \*I need to confirm my assumption that the model generates "boundary estimates" for phi (an estimate before the first primary period, and after the last primary period).
+    -   \*I need to confirm my assumption that the model generates "boundary estimates" for phi (an estimate before the first primary period and after the last primary period).
 If I can't confirm this by myself, maybe I can ask Max.
@@
 Pseudo-individuals don't actually have a capture histories, so their log_lik values are all identical.
-See the 'Compare Model Fit' section of this report for more information about this quantity.
+See the Compare Model Fit section of this report for more information about this quantity.
 
@@
 
-(*vector[m_detect] beta_detect;*) —\> (*vector[Jtot] beta_detect;*)
+*vector[m_detect] beta_detect;* —\> *vector[Jtot] beta_detect;*
 
-This change is essential to achieving independent, time-varying estimates of detection probability in the custom model.
+This substantive change is essential to achieving independent, time-varying estimates of detection probability in the custom model.
 
@@
 Jtot refers to the total number of survey sessions.
-Each entry in the vector now corresponds to a specific survey session (custom) rather than a covariate (classic).
+Each entry in the vector in the custom model corresponds to a specific survey session rather than a covariate, as in the classic model.
 
@@
 
-In the classic model, logit_detect is derived from beta_detect using the design matrix of covariates x_detect.
+In the classic model, logit_detect is derived from beta_detect using the design matrix of covariates X_detect.
 This ties detection probabilities to shared covariates.
@@
 
-Changes to the parameters and transformed parameters blocks reflect a purpose to obtain the most accurate estimates of N, B, p, and phi for the collective population.
-The classic model reflect a purpose to model detection and survival by covariates, which makes sense in the context of Knapp and Joseph (2018).
+Changes to the parameters and transformed parameters blocks are substantive and aim for the most accurate estimates of N, B, p, and phi for the population as a whole.
+The classic version aims to model detection and survival by climate and disease-related covariates, which makes sense in the context of Knapp and Joseph (2018).
 If and when I want to introduce covariates into my analysis, I will need to revisit this section.
@@
 
-This analysis uses capture-mark-recapture (CMR) data for a population of endangered Sierra Nevada yellow-legged frogs (*Rana sierrae*) in the Sierra National Forest, California.
+The analysis in this report uses capture-mark-recapture (CMR) data for a population of endangered Sierra Nevada yellow-legged frogs (*Rana sierrae*) in the Sierra National Forest, California.
 This population is referred to by the numeric site id of its associated water body, 54188.
@@
 This section describes the analysis performed to compare the classic and custom models.
-Both models were fit to the same CMR data for the Sierra Nevada yellow-legged frog population at site 54188.
+Both models were fit to the same CMR data, described in the previous section.
 Key comparisons include estimates of population abundance (N), recruitment (B), detection probability (p), and survival (phi).
-To evaluate model performance, I used the leave-one-out cross-validation (LOO) method, which assesses model fit while penalizing complexity.
+To evaluate model performance, I used the leave-one-out (LOO) cross-validation method, which assesses model fit while penalizing complexity.
 
 The goal of this analysis is to determine whether the custom model, with independent, time-varying estimates of detection probability, provides a significantly better fit to the data compared to the classic model, which assumes detection probability is constant across sessions.
-Differences in model performance will be interpreted using the LOO expected log predictive density (ELPD) values, along with associated uncertainty estimates.
+Differences in model performance will be interpreted using the LOO expected log predictive density (elpd) values, along with associated uncertainty estimates.
 
@@
 The fit_model_custom( ) function is identical to mrmr's built in fit_model( ) function, except it draws on the custom .stan file on my local machine, rather than the standard mrmr .stan file.
-The custom .stan file includes the previously described changes to the parameters, transformed parameters, and generated quantities block.
+The custom .stan file includes the previously described changes to the parameters, transformed parameters, and generated quantities blocks.
 
@@
 custom_mod_54188_summary <- custom_mod_54188$m_fit$summary()
-View(custom_mod_54188_summary)
+print(custom_mod_54188_summary)
 
@@
 classic_mod_54188_summary <- classic_mod_54188$m_fit$summary()
-View(classic_mod_54188_summary)
+print(classic_mod_54188_summary)
 

### 2024-11-22 code/python/2017_April1SWE_MP.py
initial commit of 2017 %avg April1 SWE calculation for MP

@@ -1,0 +1,148 @@
+import os
+import pandas as pd
+
+def process_station_data(station_file, station_name, apr1_avg, lat, lon, date_col='DATE TIME', value_col='VALUE'):
+    """
+    Processes snow water equivalent (SWE) data for CA snow survey stations.
+
+    Args:
+        station_file (str): File name for the station data.
+        station_name (str): Three-letter station name.
+        apr1_avg (float): April 1st average value for the station.
+        lat (float): Latitude of the station.
+        lon (float): Longitude of the station.
+        date_col (str): Column name for the date/time data.
+        value_col (str): Column name for the SWE values.
+
+    Returns:
+        pd.DataFrame: Processed DataFrame with added columns for April 1st average, pctAvg, and location.
+    """
+    # Define the base directory where your data is stored
+    base_directory = r"C:\Modeling Geographical Objects\MossyPond\Data\SWE"
+    
+    # Join the base directory with the station file name to create the full file path
+    file_path = os.path.join(base_directory, station_file)
+    
+    # Load the data and keep necessary columns
+    data = pd.read_csv(file_path)
+    
+    # Filter only the relevant columns
+    data = data[[date_col, value_col]]
+    
+    # Convert the date column to a datetime object
+    data[date_col] = pd.to_datetime(data[date_col])
+    
+    # Define April 1st dates for filtering
+    april1_dates = pd.to_datetime(["2012-04-01", "2013-04-01", "2014-04-01", "2015-04-01",
+                                   "2016-04-01", "2017-04-01", "2018-04-01"])
+    
+    # Filter for April 1st observations
+    data_filtered = data.loc[data[date_col].isin(april1_dates)].copy()
+    
+    # Add April 1st average column
+    data_filtered.loc[:, 'Apr1Avg'] = apr1_avg
+    
+    # Calculate pctAvg (percentage of the April 1st average) and round to 2 decimals
+    data_filtered.loc[:, 'pctAvg'] = round((data_filtered[value_col] / apr1_avg) * 100, 2)
+    
[…]

### 2024-11-22 code/python/2017_April1SWE_MP.py
review and edit code comments

@@ -1,5 +1,6 @@
 import os
 import pandas as pd
+import arcpy
 
 def process_station_data(station_file, station_name, apr1_avg, lat, lon, date_col='DATE TIME', value_col='VALUE'):
     """
@@ -17,39 +18,39 @@ def process_station_data(station_file, station_name, apr1_avg, lat, lon, date_co
     Returns:
         pd.DataFrame: Processed DataFrame with added columns for April 1st average, pctAvg, and location.
     """
-    # Define the base directory where your data is stored
+    # Define base directory 
     base_directory = r"C:\Modeling Geographical Objects\MossyPond\Data\SWE"
     
-    # Join the base directory with the station file name to create the full file path
+    # join base directory station file name to create full file path
     file_path = os.path.join(base_directory, station_file)
     
-    # Load the data and keep necessary columns
+    # load data 
     data = pd.read_csv(file_path)
     
-    # Filter only the relevant columns
+    # filter columns
     data = data[[date_col, value_col]]
     
-    # Convert the date column to a datetime object
+    # convert date column to a datetime object
     data[date_col] = pd.to_datetime(data[date_col])
     
-    # Define April 1st dates for filtering
+    # define April 1st dates for filtering
     april1_dates = pd.to_datetime(["2012-04-01", "2013-04-01", "2014-04-01", "2015-04-01",
                                    "2016-04-01", "2017-04-01", "2018-04-01"])
     
-    # Filter for April 1st observations
+    # filter for April 1st observations
     data_filtered = data.loc[data[date_col].isin(april1_dates)].copy()
     
-    # Add April 1st average column
+    # add April 1st average column
     data_filtered.loc[:, 'Apr1Avg'] = apr1_avg
     
-    # Calculate pctAvg (percentage of the April 1st average) and round to 2 decimals
+    # Calculate pctAvg
     data_filtered.loc[:, 'pctAvg'] = round((data_filtered[value_col] / apr1_avg) * 100, 2)
     
-    # Add location information
+    # add location information
     data_filtered.loc[:, 'LAT'] = lat
     data_filtered.loc[:, 'LON'] = lon
     
-    # Add station name column
+    # add station name column
     data_filtered.loc[:, 'Station'] = station_name
     
     return data_filtered
[…]

### 2024-11-22 code/R/MossyPond_Classic_mrmr.Rmd
update fit_model_classic function and change object names

@@
 # clean data 
-mrmr_mossy_mod1_data <- clean_data(captures, surveys)
+mp_data <- clean_data(captures, surveys)
+
+# load fit_model_classic function
+source("classic_fit_model_function.R")
 
 # fit classic mrmr model
-mrmr_mossy_mod1 <- fit_model(mrmr_mossy_mod1_data,
+classic_mod_mp <- fit_model_classic(mp_data,
                                 chains = 3, 
@@
 # save model summary to an object
-mod1_summary <- mrmr_mossy_mod1$m_fit$summary()
-view(mod1_summary)
+classic_mod_mp_summary <- classic_mod_mp$m_fit$summary()
+View(classic_mod_mp_summary)
 
-write.csv(mod1_summary, "/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_classic_mod_sumary.csv")
+write.csv(classic_mod_mp_summary, "/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_classic_mod_sumary.csv")
 

### 2024-11-22 code/R/MossyPond_Custom_mrmr_TVD.Rmd
add model comaprison code

@@
 
-# Run my custom mrmr model with Mossy Pond data
+# custom TVD mrmr model with MP data
 
@@
 ```{r}
-getwd()
+
 surveys <- read.csv("/Users/johnimperato/Desktop/Mossy-Pond/data/survey_MossyPond.csv")
@@
 
-mp_custom_data <- mrmr::clean_data(captures, surveys)
+mp_data <- mrmr::clean_data(captures, surveys)
 
@@
 
-write.csv(custom_mod_mp_summary, "/Users/johnimperato/Documents/mp_custom_mod_summary")
+write.csv(custom_mod_mp_summary, "/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_custom_mod_sumary.csv")
 
@@
 ```
+
+### Model Comparison – custom/classic
+
+#### Generated Quantities
+
+**Create a table comparing the mean and sd generated quantity values for each model**
+
+This table shows **only the mean and sd** estimated value for each generated quantity.
+It is meant to serve as a quick reference for comparing model output.
+See the individual model summaries for complete generated quantity summaries.
+
+```{r}
+# pull quantities from model fit summary
+N_classic <- classic_mod_54188_summary[, c(1,2,4)]
+N_custom <- custom_mod_54188_summary[33659:33674, c(1,2,4)]
+B_classic <- classic_mod_54188_summary[4221:4225, c(1,2,4)]
+B_custom <- custom_mod_54188_summary[4234:4238, c(1,2,4)]
+Phi_classic <- classic_mod_54188_summary[4240:4243, c(1,2,4)]
+Phi_custom <- custom_mod_54188_summary[4253:4256, c(1,2,4)]
+p_classic <- classic_mod_54188_summary[4226:4239, c(1,2,4)]
+p_custom <- custom_mod_54188_summary[4239:4252, c(1,2,4)]
+
+GenQuant_comparison <- bind_rows(
+  left_join(N_classic, N_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "abundance"),
+  left_join(B_classic, B_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "recruitment"),
+  left_join(Phi_classic, Phi_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "survival"),
+  left_join(p_classic, p_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "detection"),
+) %>% 
+  select(variable, parameter, everything())
+
+print(GenQuant_comparison)
+
+# Convert table to grob (graphical object)
+table_grob <- tableGrob(GenQuant_comparison)
+
+# Save as an image
+ggsave("/Users/johnimperato/Desktop/Mossy-Pond/out/54188_comparison_table.png", plot = table_grob, width = 8, height = 8)
+
+```
+
+**Visually compare abundance and recruitment estimates between models**
+
+Note that axes are not standardized.
+
+```{r}
+# visual comparison of abundance and recruitment plots
+abund_recruit_comparison_plot_54188 <- ggarrange(classic_annotated, custom_annotated, ncol=2)
+print(abund_recruit_comparison_plot_54188)
+
+```
+
+#### Model Fit
+
+Use leave-one-out (LOO) cross-validation, a standard method to calculate the predictive accuracy of Bayesian models, to compare how well the classic and custom models fit the Trapdoor data.
+The cmdstanr package has built in LOO capabilities.
+Briefly,
+
+-   LOO cross-validation uses log likelihood values, which now appear in the model summary as generated quantities, to evaluate how well the model predicts each data point when that point is left out of the fitting process.
+Likelihood, calculated at the individual level, is the probability of observing the data (in this case, an individual frog's capture history) given the model parameters.
+Log likelihood calculated at the individual level in the generated quantities block is different than the total log likelihood, summed across all individuals and sessions, calculated during model fitting in the model block.
+
+The loo( ) function generates an expected log predictive density (elpd) value for each model.
+The loo_compare( ) function ranks models based on their elpd values, with the best fit listed first, and calculates the difference in those values.
+A difference of more than 2x the standard error is generally considered significant in a LOO comparison.
+
+```{r}
+loo_classic <- classic_mod_54188$m_fit$loo()
+
+loo_custom <- custom_mod_54188$m_fit$loo()
+
+loo_compare <- loo::loo_compare(loo_classic, loo_custom)
+
+print(loo_compare)
+
+# extract elpd_diff and se_diff for model1
+elpd_diff_model1 <- loo_compare["model1", "elpd_diff"]
+se_diff_model1 <- loo_compare["model1", "se_diff"]
+
+# calculate significance ratio
+result <- abs(elpd_diff_model1 / se_diff_model1)
+
+# interpret
+if (abs(result) > 2) {
+  cat("The absolute value of elpd_diff / se_diff for model1:", result, "\n")
+  cat("The custom model fits the data significantly better than the classic model.\n")
+} else {
+  cat("The absolute value of elpd_diff / se_diff for model1:", result, "\n")
+  cat("The custom model does not fit the data significantly better than the classic model.\n")
+}
+
+```

### 2024-11-22 code/R/classic_fit_model_function.R
change file path for classic stan model

@@ -1,4 +1,4 @@
-fit_model_classic_loglik <- function(
+fit_model_classic <- function(
     data,
     seed = NULL,
     refresh = NULL,
@@ -32,7 +32,7 @@ fit_model_classic_loglik <- function(
     ...
 ) {
   # updated to use my custom stan model
-  stan_file <- "/Users/johnimperato/Desktop/Mossy-Pond/code/stan/twostate_loglik_modified.stan" # specify local path
+  stan_file <- "/Users/johnimperato/Desktop/Mossy-Pond/code/stan/JI_classic_twostate.stan" # specify local path
   
   model <- cmdstan_model(stan_file, ...)
   

### 2024-11-22 code/R/MossyPond_Custom_mrmr_TVD.Rmd
format GQ comparison table

--- 2024-11-16 code/R/MossyPond_Custom_mrmr_TVD.Rmd
@@
 
-# Run my custom mrmr model with Mossy Pond data
+# custom TVD mrmr model with MP data
 
 ```{r}
+library(dplyr)
+library(readr)
+library(tibble)
+library(cmdstanr)
 library(mrmr)
+library(ggpubr)
 ```
@@
 
-write.csv(custom_mod_mp_summary, "/Users/johnimperato/Documents/mp_custom_mod_summary")
+write.csv(custom_mod_mp_summary, "/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_custom_mod_summary.csv")
 
 ```
+
+**Plot abundance and recruitment**
 
@@
 
+mp_custom_plots <- ggarrange(abundance_MP_custom, recruitment_MP_custom, ncol=1)
+
+# add title
+mp_custom_plots_annotated <- annotate_figure(mp_custom_plots, top = text_grob("Custom mrmr", size = 14, face = "bold"))
+
+mp_custom_plots_annotated
+
 ```
+
+### Model Comparison – custom/classic
+
+#### Generated Quantities
+
+*Parameter comparison table*
+
+```{r}
+# pull generated quantities from model fit summary
+N_classic <- classic_mod_54188_summary[, c(1,2,4)]
+N_custom <- custom_mod_54188_summary[33659:33674, c(1,2,4)]
+B_classic <- classic_mod_54188_summary[4221:4225, c(1,2,4)]
+B_custom <- custom_mod_54188_summary[33676:33691, c(1,2,4)]
+Phi_classic <- classic_mod_54188_summary[4240:4243, c(1,2,4)]
+Phi_custom <- custom_mod_54188_summary[33739:33753, c(1,2,4)]
+p_classic <- classic_mod_54188_summary[4226:4239, c(1,2,4)]
+p_custom <- custom_mod_54188_summary[33692:33738, c(1,2,4)]
+
+GenQuant_comparison_mp <- bind_rows(
+  left_join(N_classic, N_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "abundance"),
+  left_join(B_classic, B_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "recruitment"),
+  left_join(Phi_classic, Phi_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "survival"),
+  left_join(p_classic, p_custom, by = "variable", suffix = c("_classic", "_custom")) %>% mutate(parameter = "detection"),
+) %>% 
+  select(variable, parameter, everything())
+
+print(GenQuant_comparison_mp)
+
+# Convert table to grob (graphical object)
+table_grob <- tableGrob(GenQuant_comparison_mp)
+
+# Save as an image
+ggsave("/Users/johnimperato/Desktop/Mossy-Pond/out/MP_comparison_table.png", plot = table_grob, width = 8, height = 10)
+
+```
+
+**Abundance/Recruitment comparison plot**
+Note that axes are not standardized.
+
+```{r}
+# visual comparison of abundance and recruitment plots
+abund_recruit_comparison_plot_mp <- ggarrange(mp_classic_plots_annotated, mp_custom_plots_annotated, ncol=2)
+print(abund_recruit_comparison_plot_mp)
+
+```
+
+#### Model Fit
+
+**LOO comparison**
+
+```{r}
+loo_classic <- classic_mod_54188$m_fit$loo()
+
+loo_custom <- custom_mod_54188$m_fit$loo()
+
+loo_compare <- loo::loo_compare(loo_classic, loo_custom)
+
+print(loo_compare)
+
+# extract elpd_diff and se_diff for model1
+elpd_diff_model1 <- loo_compare["model1", "elpd_diff"]
+se_diff_model1 <- loo_compare["model1", "se_diff"]
+
+# calculate significance ratio
+result <- abs(elpd_diff_model1 / se_diff_model1)
+
+# interpret
+if (abs(result) > 2) {
+  cat("The absolute value of elpd_diff / se_diff for model1:", result, "\n")
+  cat("The custom model fits the data significantly better than the classic model.\n")
+} else {
+  cat("The absolute value of elpd_diff / se_diff for model1:", result, "\n")
+  cat("The custom model does not fit the data significantly better than the classic model.\n")
+}
+
+```

### 2024-11-22 code/R/MossyPond_Custom_mrmr_TVD.Rmd
create comparison table and plot with updated classic model

@@
 library(ggpubr)
+library(gridExtra)
+library(grid)
 ```
@@
 # pull generated quantities from model fit summary
-N_classic <- classic_mod_54188_summary[, c(1,2,4)]
-N_custom <- custom_mod_54188_summary[33659:33674, c(1,2,4)]
-B_classic <- classic_mod_54188_summary[4221:4225, c(1,2,4)]
-B_custom <- custom_mod_54188_summary[33676:33691, c(1,2,4)]
-Phi_classic <- classic_mod_54188_summary[4240:4243, c(1,2,4)]
-Phi_custom <- custom_mod_54188_summary[33739:33753, c(1,2,4)]
-p_classic <- classic_mod_54188_summary[4226:4239, c(1,2,4)]
-p_custom <- custom_mod_54188_summary[33692:33738, c(1,2,4)]
+N_classic <- classic_mod_mp_summary[33613:33628, c(1,2,4)]
+N_custom <- custom_mod_mp_summary[33658:33674, c(1,2,4)]
+B_classic <- classic_mod_mp_summary[33629:33644, c(1,2,4)]
+B_custom <- custom_mod_mp_summary[33675:33690, c(1,2,4)]
+Phi_classic <- classic_mod_mp_summary[33692:33706, c(1,2,4)]
+Phi_custom <- custom_mod_mp_summary[33738:33752, c(1,2,4)]
+p_classic <- classic_mod_mp_summary[33645:33691, c(1,2,4)]
+p_custom <- custom_mod_mp_summary[33691:33737, c(1,2,4)]
 
@@
 
-print(GenQuant_comparison_mp)
+View(GenQuant_comparison_mp)
 
@@
 ggsave("/Users/johnimperato/Desktop/Mossy-Pond/out/MP_comparison_table.png", plot = table_grob, width = 8, height = 10)
+
+  View(custom_mod_mp_summary)
 

### 2024-11-22 code/R/MossyPond_Classic_mrmr.Rmd
run updated model with new gen quants

@@
 library(tibble)
-library(rstan)
+library(cmdstanr)
 library(mrmr)
@@
 
-abundance_MP_classic <- plot_model(mrmr_mossy_mod1, "abundance")
-recruitment_MP_classic <- plot_model(mrmr_mossy_mod1, "recruitment")
+abundance_MP_classic <- plot_model(classic_mod_mp, "abundance")
+recruitment_MP_classic <- plot_model(classic_mod_mp, "recruitment")
 
-MP_classic_plots <- ggarrange(abundance_MP_classic, recruitment_MP_classic, ncol=1)
-MP_classic_plots
+mp_classic_plots <- ggarrange(abundance_MP_classic, recruitment_MP_classic, ncol=1)
+
+# add title
+mp_classic_plots_annotated <- annotate_figure(mp_classic_plots, top = text_grob("Classic mrmr", size = 14, face = "bold"))
+
+mp_classic_plots_annotated
 

### 2024-11-22 code/R/54188_ClassicCustomComparison_mrmr.Rmd
edit intro

@@
 library(dplyr)
+library(gridExtra)
+library(grid)
+library(ggplot2)
 ```
@@
 
-**Create a table comparing the mean generated quantity values for each model**
+**Create a table comparing the mean and sd generated quantity values for each model**
 
-This table shows **only the mean** estimated value for each generated quantity.
+This table shows **only the mean and sd** estimated value for each generated quantity.
 It is meant to serve as a quick reference for comparing model output.
@@
 # pull quantities from model fit summary
-N_classic <- classic_mod_54188_summary[4216:4220, 1:2]
-N_custom <- custom_mod_54188_summary[4229:4233, 1:2]
-B_classic <- classic_mod_54188_summary[4221:4225, 1:2]
-B_custom <- custom_mod_54188_summary[4234:4238, 1:2]
-Phi_classic <- classic_mod_54188_summary[4240:4243, 1:2]
-Phi_custom <- custom_mod_54188_summary[4253:4256, 1:2]
-p_classic <- classic_mod_54188_summary[4226:4239, 1:2]
-p_custom <- custom_mod_54188_summary[4239:4252, 1:2]
+N_classic <- classic_mod_54188_summary[4216:4220, c(1,2,4)]
+N_custom <- custom_mod_54188_summary[4229:4233, c(1,2,4)]
+B_classic <- classic_mod_54188_summary[4221:4225, c(1,2,4)]
+B_custom <- custom_mod_54188_summary[4234:4238, c(1,2,4)]
+Phi_classic <- classic_mod_54188_summary[4240:4243, c(1,2,4)]
+Phi_custom <- custom_mod_54188_summary[4253:4256, c(1,2,4)]
+p_classic <- classic_mod_54188_summary[4226:4239, c(1,2,4)]
+p_custom <- custom_mod_54188_summary[4239:4252, c(1,2,4)]
 
@@
 
+# Convert table to grob (graphical object)
+table_grob <- tableGrob(GenQuant_comparison)
+
+# Save as an image
+ggsave("/Users/johnimperato/Desktop/Mossy-Pond/out/54188_comparison_table.png", plot = table_grob, width = 8, height = 8)
+
 ```
@@
 print(abund_recruit_comparison_plot_54188)
+
 ```
@@
 Likelihood, calculated at the individual level, is the probability of observing the data (in this case, an individual frog's capture history) given the model parameters.
+Log likelihood calculated at the individual level in the generated quantities block is different than the total log likelihood, summed across all individuals and sessions, calculated during model fitting in the model block.
 
@@
 # calculate significance ratio
-result <- elpd_diff_model1 / se_diff_model1
+result <- abs(elpd_diff_model1 / se_diff_model1)
 
@@
 if (abs(result) > 2) {
-  cat("Result of elpd_diff / se_diff for model1:", result, "\n")
+  cat("The absolute value of elpd_diff / se_diff for model1:", result, "\n")
   cat("The custom model fits the data significantly better than the classic model.\n")
 } else {
-  cat("Result of elpd_diff / se_diff for model1:", result, "\n")
+  cat("The absolute value of elpd_diff / se_diff for model1:", result, "\n")
   cat("The custom model does not fit the data significantly better than the classic model.\n")

### 2024-11-22 code/R/custom_fit_model_function.R
remove function save attempt

@@ -76,3 +76,5 @@ fit_model_custom <- function(
   
   list(m_fit = m_fit, data = data)
 }
+
+

### 2024-12-05 code/R/WinterSeverity_analysis_MossyPond.Rmd
Initial commit of Winter Severity - Vital Rates analysis for Mossy Pond

---
title: "Yearly 1 April SWE and Vital Rates"
output: html_document
date: "2024-12-04"
---

# Analysis of Winter Severity and *Rana sierrae* Population Characteristics

**Winter severity measure**: percent of historical average 1 April SWE in the Mossy Pond study area.

-   snow data taken from 10 observation stations between 5.1 and 19.5 km from the MPSA

## Yearly % of historical average 1 April SWE in the Mossy Pond study area

```{r}
library(ggplot2)
```

```{r}

# create winter severity data frame
avg_swe <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond/data/Yearly_WinterSeverity_MPSA.csv')

avg_swe <- avg_swe %>%
  rename(Winter_Severity = Average.Pixel.Value) %>%
  mutate(Winter = c("2011-2012", "2012-2013", "2013-2014", "2014-2015", "2015-2016", "2016-2017", "2017-2018"))

# create data frame for winter severity one year prior
avg_swe_year_prior <- avg_swe %>%
                        mutate(Year = Year + 1)

```

## Population Vital Rates/Numbers and Growth Rate

**N, B, and phi**

```{r}

model_summary <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_custom_mod_sumary.csv')

# extract N values from model summary
N <- model_summary[33659:33674, ]

# calculate yearly average N
N_yearly_avg <- N %>%
  # Group rows and calculate the mean of numeric columns
  mutate(group = c(rep(1:4, each = 3), rep(5, 4))) %>%
  group_by(group) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
  ungroup() %>%
  select(-group) %>%
  # Add a new Year column
  mutate(Year = c(2014:2018)) %>%
  # Rename 'mean' to 'Abundance' and reorder columns
  rename(Mean_Abundance = mean) %>%
  select(Year, Mean_Abundance)

# extract B values from model summary
B <- model_summary[c(33675:33690), ]

# calculate yearly total B
B_yearly_avg <- B %>%
  # Group rows and calculate the mean of numeric columns
  mutate(group = c(rep(1:4, each = 3), rep(5, 4))) %>%
  group_by(group) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
  ungroup() %>%
  select(-group) %>%
  # Add a new Year column
  mutate(Year = c(2014:2018)) %>%
  # Rename 'mean' to 'Recruited_Individuals' and reorder columns
  rename(Recruited_Individuals = mean) %>%
  select(Year, Recruited_Individuals)

# extract overwinter survival values from model summary
phi <- B <- model_summary[c(33738:33752), ]

# extract overwinter survival
phi_overwinter <- phi %>%
  slice(c(3, 6, 9, 12)) %>%
  # Add a new Year column
  mutate(Winter = c("2014-2015", "2015-2016", "2016-2017", "2017-2018")) %>%
  # Rename 'mean' to 'Recruited_Individuals' and reorder columns
  rename(Overwinter_Survival_Probability = mean) %>%
  select(Winter, Overwinter_Survival_Probability)

```

**Growth rate**

```{r}

```

## Abundance-Winter severity Comaprison

### Winter 1 year ago

```{r}

# combine data fames
comparison_df_1 <- N_yearly_avg %>%
  left_join(avg_swe_year_prior, by = "Year")

# plot
abundance_WinterSeverity_1yr <- ggplot(comparison_df_1, aes(x = Year)) +
  # frog abundance line
  geom_line(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 1) +
  geom_point(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 2) +
  # winter severity line
  geom_line(aes(y = Winter_Severity, color = "Winter Severity One Year Prior\n(% of historical avg 1 April SWE)"), linetype = "dashed", size = 1) +
  geom_point(aes(y = Winter_Severity, color = "Winter Severity One Year Prior\n(% of historical avg 1 April SWE)"), size = 2) +
  # Labels and theme
  scale_color_manual(
    name = "Legend",
    values = c("Frog Abundance" = "blue", "Winter Severity One Year Prior\n(% of historical avg 1 April SWE)" = "red")
  ) +
  theme_minimal() +
  labs(
    title = "Frog Abundance and Winter Severity One Year Prior",
    x = "Year",
    y = "Value"
  )

# correlation
correlation1 <- cor(comparison_df_1$Mean_Abundance, comparison_df_1$Winter_Severity)
cor_test1 <- cor.test(comparison_df_1$Mean_Abundance, comparison_df_1$Winter_Severity)

correlation_plot1 <- ggplot(data = comparison_df_1, aes(x = Winter_Severity, y = Mean_Abundance)) +
  geom_point() +
  geom_smooth(method = "lm" ,se = FALSE) +
  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation1, 2)), size = 5, color = "red") +
  labs(
    title = "Correlation between Winter Severity and Abundance",
    x = "Winter Severity (% of Historical Average 1 April SWE)",
    y = "Yearly Mean Abundance"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

## Recruitment - Winter Severity Comparison

### Winter 1 year ago

```{r}

# combine data fames
comparison_df_2 <- B_yearly_avg %>%
  left_join(avg_swe_year_prior, by = "Year")

# plot
WinterSeverity_Recruitment <- ggplot(comparison_df_2, aes(x = Year)) +
  # frog abundance line
  geom_line(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 1) +
  geom_point(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 2) +
  # winter severity line
  geom_line(aes(y = Winter_Severity, color = "Winter Severity One Year Prior\n(% of historical avg 1 April SWE)"), linetype = "dashed", size = 1) +
  geom_point(aes(y = Winter_Severity, color = "Winter Severity One Year Prior\n(% of historical avg 1 April SWE)"), size = 2) +
  # Labels and theme
  scale_color_manual(
    name = "Legend",
    values = c("Individuals Recruited" = "blue", "Winter Severity One Year Prior\n(% of historical avg 1 April SWE)" = "red")
  ) +
  theme_minimal() +
  labs(
    title = "Recruitment and Winter Severity One Year Prior",
    x = "Year",
    y = "Value"
  )

# correlation
correlation2 <- cor(comparison_df_2$Winter_Severity, comparison_df_2$Recruited_Individuals)
cor_test2 <- cor.test(comparison_df_1$Winter_Severity, comparison_df_2$Recruited_Individuals)

correlation_plot2 <- ggplot(data = comparison_df_2, aes(x = Winter_Severity, y = Recruited_Individuals)) +
  geom_point() +
  geom_smooth(method = "lm" ,se = FALSE) +
  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation2, 2)), size = 5, color = "red") +
  labs(
    title = "Correlation between Winter Severity and Recruitment",
    x = "Winter Severity (% of Historical Average 1 April SWE)",
    y = "Total Yearly Recruitment (Number of individuals)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

## Survival Probability - Winter Severity Comparison

### Same Winter

```{r}

# combine data fames
comparison_df_3 <- phi_overwinter %>%
  left_join(avg_swe, by = "Winter")

# Ensure Winter is a factor in the correct order
comparison_df_3 <- comparison_df_3 %>%
  mutate(Winter = factor(Winter, levels = c("2014-2015", "2015-2016", "2016-2017", "2017-2018"))) %>%
  mutate(Survival_Percentage = Overwinter_Survival_Probability * 100)

# Plot
ggplot(comparison_df_3, aes(x = Winter)) +
  # survival line
  geom_line(aes(y = Survival_Percentage, color = "Survival Percentage"), size = 1, group = 1) +
  geom_point(aes(y = Survival_Percentage, color = "Survival Percentage"), size = 2) +
  # winter severity line
  geom_line(aes(y = Winter_Severity, color = "Winter Severity"), linetype = "dashed", size = 1, group = 2) +
  geom_point(aes(y = Winter_Severity, color = "Winter Severity"), size = 2) +
  # Labels and theme
  scale_color_manual(
    name = "Legend",
    values = c("Survival Percentage" = "blue", "Winter Severity" = "red")
  ) +
  theme_minimal() +
  labs(
    title = "Survival and Winter Severity",
    x = "Winter",
    y = "Value"
  )

```

### 2024-12-05 code/python/Snowpack_Interpolation_MossyPond.py
Initial commit of yearly snowpack interpolation

@@ -1,0 +1,166 @@
+import os
+import pandas as pd
+import arcpy
+
+def process_station_data_by_year(station_file, station_name, apr1_avg, lat, lon, years, base_directory, date_col='DATE TIME', value_col='VALUE'):
+    """
+    Processes SWE data for multiple years and a given station.
+    
+    Args:
+        station_file (str): File name for the station data.
+        station_name (str): Three-letter station name.
+        apr1_avg (float): April 1st average value for the station.
+        lat (float): Latitude of the station.
+        lon (float): Longitude of the station.
+        years (list): List of years to process.
+        base_directory (str): Base directory containing the station files.
+        date_col (str): Column name for the date/time data.
+        value_col (str): Column name for the SWE values.
+    
+    Returns:
+        pd.DataFrame: Combined DataFrame for all years with additional information.
+    """
+    file_path = os.path.join(base_directory, station_file)
+    data = pd.read_csv(file_path)
+    data[date_col] = pd.to_datetime(data[date_col])
+    
+    # Generate April 1st dates for filtering
+    april1_dates = [pd.Timestamp(f"{year}-04-01") for year in years]
+    
+    data_filtered = data[data[date_col].isin(april1_dates)].copy()
+    data_filtered['Apr1Avg'] = apr1_avg
+    data_filtered['pctAvg'] = round((data_filtered[value_col] / apr1_avg) * 100, 2)
+    data_filtered['LAT'] = lat
+    data_filtered['LON'] = lon
+    data_filtered['Station'] = station_name
+    
+    return data_filtered
+
+# List of years to process
+years = list(range(2012, 2019))
+
+# Base directory for SWE files
+base_directory = r"C:\Mossy Pond\SWE_MossyPond\SWE_Data"
+
+# Process each station
+stations = [
+    ("RDM.csv", "RDM", 48.7, 39.343, -120.508),
+    ("DNS.csv", "DNS", 33.1, 39.31, -120.338),
+    ("FOD.csv", "FOD", 36.5, 39.36, -120.5),
+    ("CSL.csv", "CSL", 34.2, 39.325, -120.367),
[…]

### 2025-01-10 doc/Manuscript.qmd
add shell manuscript document

---
title: "manuscript"
format: html
editor: visual
---

## Title Title Title

John Imperato^1^

^1^School of the Environment, Yale University, New Haven, CT, 06511

### Abstract

### Introduction

-   Setup: which scale?

    -   Broadest: global biodiversity loss (as in Knapp et al. 2024)

    -   Middle: global amphibian declines (as in Keung et al. 2019)

    -   Narrowest: SNYLF are an endangered species... (as in Pope and Matthews 2004)

-   Contextualize:

    -   reintroduction as a conservation method (as in Keung et al. 2019)

    -   Reintroductions of Bd resistant Rana sierrae have proved to be a successful method for recovery in the Yosemite National park (Knapp et al. 2024).

-   Bridge:

    -   Reintroductions rely on donations from naturally persistent populations with evolved resistance to Bd

    -   It is vital to understand what drives the population dynamics of these persistent donor populations to inform conservation efforts

    -   Large, persistent populations in the northern part of the species' range are relatively rare compared to the central and southern parts of he range.

        -   especially populations occupying atypical habitat (streams, ephemeral ponds, shallow lake)

    -   Mossy Pond is one such population

        -   In the TNF, managed largely by CDFW (?)

        -   other relevant descriptions

    -   Mossy Pond population is a candidate donor population for areas identified as Native Species Reserves for SNYLF

-   Study Overview

    -   in 2014, CDFW began a 5-year CMR study to gain insight into this population because of its conservation significance

        -   get some of the "original directive" from Sara M & Isaac

    -   This paper describes analysis of the resulting data and establishes a relationship between (local climate variables and SNYLF population viability).

### Methods

### Results

### References

### Figures

### 2025-01-10 doc/Manuscript.qmd
delete introduction notes

@@
 
--   Setup: which scale?
-
-    -   Broadest: global biodiversity loss (as in Knapp et al. 2024)
-
-    -   Middle: global amphibian declines (as in Keung et al. 2019)
-
-    -   Narrowest: SNYLF are an endangered species... (as in Pope and Matthews 2004)
-
--   Contextualize:
-
-    -   reintroduction as a conservation method (as in Keung et al. 2019)
-
-    -   Reintroductions of Bd resistant Rana sierrae have proved to be a successful method for recovery in the Yosemite National park (Knapp et al. 2024).
-
--   Bridge:
-
-    -   Reintroductions rely on donations from naturally persistent populations with evolved resistance to Bd
-
-    -   It is vital to understand what drives the population dynamics of these persistent donor populations to inform conservation efforts
-
-    -   Large, persistent populations in the northern part of the species' range are relatively rare compared to the central and southern parts of he range.
-
-        -   especially populations occupying atypical habitat (streams, ephemeral ponds, shallow lake)
-
-    -   Mossy Pond is one such population
-
-        -   In the TNF, managed largely by CDFW (?)
-
-        -   other relevant descriptions
-
-    -   Mossy Pond population is a candidate donor population for areas identified as Native Species Reserves for SNYLF
-
--   Study Overview
-
-    -   in 2014, CDFW began a 5-year CMR study to gain insight into this population because of its conservation significance
-
-        -   get some of the "original directive" from Sara M & Isaac
-
-    -   This paper describes analysis of the resulting data and establishes a relationship between (local climate variables and SNYLF population viability).
-
 ### Methods

### 2025-01-15 code/python/WinterSeverity_MossyPond_2011_2018.py
debug winter severity calculation script

@@ -1,0 +1,191 @@
+import os
+import pandas as pd
+import arcpy
+
+# Enable overwriting of outputs in ArcGIS
+arcpy.env.overwriteOutput = True
+
+def process_station_data_by_year(station_file, station_name, apr1_avg, lat, lon, years, base_directory, date_col='DATE TIME', value_col='VALUE'):
+    """
+    Processes SWE data for multiple years and a given station.
+    
+    Args:
+        station_file (str): File name for the station data.
+        station_name (str): Three-letter station name.
+        apr1_avg (float): April 1st average value for the station.
+        lat (float): Latitude of the station.
+        lon (float): Longitude of the station.
+        years (list): List of years to process.
+        base_directory (str): Base directory containing the station files.
+        date_col (str): Column name for the date/time data.
+        value_col (str): Column name for the SWE values.
+    
+    Returns:
+        pd.DataFrame: Combined DataFrame for all years with additional information.
+    """
+    file_path = os.path.join(base_directory, station_file)
+    data = pd.read_csv(file_path)
+    data[date_col] = pd.to_datetime(data[date_col])
+    
+    # Generate April 1st dates for filtering
+    april1_dates = [pd.Timestamp(f"{year}-04-01") for year in years]
+    
+    data_filtered = data[data[date_col].isin(april1_dates)].copy()
+    data_filtered['Apr1Avg'] = apr1_avg
+    data_filtered['pctAvg'] = round((data_filtered[value_col] / apr1_avg) * 100, 2)
+    data_filtered['LAT'] = lat
+    data_filtered['LON'] = lon
+    data_filtered['Station'] = station_name
+    
+    return data_filtered
+
+# Define the years to process
+years = list(range(2010, 2019))
+
+# Base directory for SWE files
+base_directory = r"C:\Mossy Pond\SWE_MossyPond\SWE_Data"
+
+# List of stations with metadata
+stations = [
+    ("RDM.csv", "RDM", 48.7, 39.343, -120.508),
+    ("DNS.csv", "DNS", 33.1, 39.31, -120.338),
+    ("FOD.csv", "FOD", 36.5, 39.36, -120.5),
[…]

### 2025-01-16 code/python/Snowpack_Interpolation_MossyPond.py
remove outdated version of winter severity calculation script

@@ -1,166 +1,0 @@
-import os
-import pandas as pd
-import arcpy
-
-def process_station_data_by_year(station_file, station_name, apr1_avg, lat, lon, years, base_directory, date_col='DATE TIME', value_col='VALUE'):
-    """
-    Processes SWE data for multiple years and a given station.
-    
-    Args:
-        station_file (str): File name for the station data.
-        station_name (str): Three-letter station name.
-        apr1_avg (float): April 1st average value for the station.
-        lat (float): Latitude of the station.
-        lon (float): Longitude of the station.
-        years (list): List of years to process.
-        base_directory (str): Base directory containing the station files.
-        date_col (str): Column name for the date/time data.
-        value_col (str): Column name for the SWE values.
-    
-    Returns:
-        pd.DataFrame: Combined DataFrame for all years with additional information.
-    """
-    file_path = os.path.join(base_directory, station_file)
-    data = pd.read_csv(file_path)
-    data[date_col] = pd.to_datetime(data[date_col])
-    
-    # Generate April 1st dates for filtering
-    april1_dates = [pd.Timestamp(f"{year}-04-01") for year in years]
-    
-    data_filtered = data[data[date_col].isin(april1_dates)].copy()
-    data_filtered['Apr1Avg'] = apr1_avg
-    data_filtered['pctAvg'] = round((data_filtered[value_col] / apr1_avg) * 100, 2)
-    data_filtered['LAT'] = lat
-    data_filtered['LON'] = lon
-    data_filtered['Station'] = station_name
-    
-    return data_filtered
-
-# List of years to process
-years = list(range(2012, 2019))
-
-# Base directory for SWE files
-base_directory = r"C:\Mossy Pond\SWE_MossyPond\SWE_Data"
-
-# Process each station
-stations = [
-    ("RDM.csv", "RDM", 48.7, 39.343, -120.508),
-    ("DNS.csv", "DNS", 33.1, 39.31, -120.338),
-    ("FOD.csv", "FOD", 36.5, 39.36, -120.5),
-    ("CSL.csv", "CSL", 34.2, 39.325, -120.367),
[…]

### 2025-01-16 code/python/2017_April1SWE_MP.py
remove outdated script

@@ -1,149 +1,0 @@
-import os
-import pandas as pd
-import arcpy
-
-def process_station_data(station_file, station_name, apr1_avg, lat, lon, date_col='DATE TIME', value_col='VALUE'):
-    """
-    Processes snow water equivalent (SWE) data for CA snow survey stations.
-
-    Args:
-        station_file (str): File name for the station data.
-        station_name (str): Three-letter station name.
-        apr1_avg (float): April 1st average value for the station.
-        lat (float): Latitude of the station.
-        lon (float): Longitude of the station.
-        date_col (str): Column name for the date/time data.
-        value_col (str): Column name for the SWE values.
-
-    Returns:
-        pd.DataFrame: Processed DataFrame with added columns for April 1st average, pctAvg, and location.
-    """
-    # Define base directory 
-    base_directory = r"C:\Modeling Geographical Objects\MossyPond\Data\SWE"
-    
-    # join base directory station file name to create full file path
-    file_path = os.path.join(base_directory, station_file)
-    
-    # load data 
-    data = pd.read_csv(file_path)
-    
-    # filter columns
-    data = data[[date_col, value_col]]
-    
-    # convert date column to a datetime object
-    data[date_col] = pd.to_datetime(data[date_col])
-    
-    # define April 1st dates for filtering
-    april1_dates = pd.to_datetime(["2012-04-01", "2013-04-01", "2014-04-01", "2015-04-01",
-                                   "2016-04-01", "2017-04-01", "2018-04-01"])
-    
-    # filter for April 1st observations
-    data_filtered = data.loc[data[date_col].isin(april1_dates)].copy()
-    
-    # add April 1st average column
-    data_filtered.loc[:, 'Apr1Avg'] = apr1_avg
-    
-    # Calculate pctAvg
-    data_filtered.loc[:, 'pctAvg'] = round((data_filtered[value_col] / apr1_avg) * 100, 2)
-    
-    # add location information
-    data_filtered.loc[:, 'LAT'] = lat
-    data_filtered.loc[:, 'LON'] = lon
-    
[…]

### 2025-01-17 code/R/WinterSeverity_analysis_MossyPond.Rmd
add code for analaysis with 3-yr avg winters

@@
 
--   snow data taken from 10 observation stations between 5.1 and 19.5 km from the MPSA
+-   1 April SWE data interpolated from 10 observation stations between 5.1 and 19.5 km from the MPSA
+-   Data downloaded from: <https://cdec.water.ca.gov/dynamicapp/wsSensorData>
 
@@
 library(ggplot2)
+library(tidyverse)
+library(ggpubr)
 ```
@@
 # create winter severity data frame
-avg_swe <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond/data/Yearly_WinterSeverity_MPSA.csv')
+avg_swe <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond/data/Yearly_Winter_Severity_2011_2018_MPSA.csv')
 
@@
   rename(Winter_Severity = Average.Pixel.Value) %>%
-  mutate(Winter = c("2011-2012", "2012-2013", "2013-2014", "2014-2015", "2015-2016", "2016-2017", "2017-2018"))
+  mutate(Winter = c("2010-2011", "2011-2012", "2012-2013", "2013-2014", "2014-2015", "2015-2016", "2016-2017", "2017-2018"))
 
@@
                         mutate(Year = Year + 1)
+
+# create data frame for winter severity two years prior
+avg_swe_2year_prior <- avg_swe %>%
+                        mutate(Year = Year + 2)
+
+# three-year rolling average 
+avg_swe_3yrAvg <- avg_swe %>% 
+  mutate(
+    swe_3yrAvg = (
+      Winter_Severity + 
+      lag(Winter_Severity, 1) +
+      lag(Winter_Severity, 2)
+      ) / 3
+  )
+
+# three-year rolling average with one-year lag
+avg_swe_3yrAvg_1yrLag <- avg_swe %>% 
+  mutate(
+    swe_3yrAvg_1yrLag = (
+      lag(Winter_Severity, 1) +
+      lag(Winter_Severity, 2) +
+      lag(Winter_Severity, 3)
+      ) / 3
+  )
 
@@
   rename(Mean_Abundance = mean) %>%
+  select(Year, Mean_Abundance)
+
+# calculate yearly median N
+N_yearly_median <- N %>%
+  # Group rows and calculate the mean of numeric columns
+  mutate(group = c(rep(1:4, each = 3), rep(5, 4))) %>%
+  group_by(group) %>%
+  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
+  ungroup() %>%
+  select(-group) %>%
+  # Add a new Year column
+  mutate(Year = c(2014:2018)) %>%
+  # Rename 'mean' to 'Abundance' and reorder columns
+  rename(Mean_Abundance = median) %>%
   select(Year, Mean_Abundance)
@@
 
-```
+print(N)
+print(B)
 
-**Growth rate**
-
-```{r}
+print(B_yearly_avg)
+print(N_yearly_avg)
+print(phi_overwinter)
 
@@
 
-### Winter 1 year ago
+### Same Year
 
@@
 # combine data fames
-comparison_df_1 <- N_yearly_avg %>%
+comparison_df_N0 <- N_yearly_avg %>%
+  left_join(avg_swe, by = "Year")
+
+# plot
+WinterSeverity_N0 <- ggplot(comparison_df_N0, aes(x = Year)) +
+  # frog abundance line
+  geom_line(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 1) +
+  geom_point(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 2) +
+  # winter severity line
+  geom_line(aes(y = Winter_Severity, color = "Winter Severity Same Year\n(% of historical avg 1 April SWE)"), linetype = "dashed", size = 1) +
+  geom_point(aes(y = Winter_Severity, color = "Winter Severity Same Year\n(% of historical avg 1 April SWE)"), size = 2) +
+  # Labels and theme
+  scale_color_manual(
+    name = "Legend",
+    values = c("Frog Abundance" = "blue", "Winter Severity Same Year\n(% of historical avg 1 April SWE)" = "red")
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Frog Abundance and Winter Severity Same Year",
+    x = "Year",
+    y = "Value"
+  )
+
+# correlation
+correlation_N0 <- cor(comparison_df_N0$Mean_Abundance, comparison_df_N0$Winter_Severity)
+cor_test_N0 <- cor.test(comparison_df_N0$Mean_Abundance, comparison_df_N0$Winter_Severity)
+
+correlation_plot_N0 <- ggplot(data = comparison_df_S_N, aes(x = Winter_Severity, y = Mean_Abundance)) +
+  geom_point() +
+  geom_smooth(method = "lm" ,se = FALSE) +
+  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_N0, 2)), size = 5, color = "red") +
+  labs(
+    title = "Correlation between Same-Year Winter Severity and Abundance",
+    x = "Winter Severity (% of Historical Average 1 April SWE)",
+    y = "Yearly Mean Abundance"
+  ) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+######################## Print
+print(WinterSeverity_N0)
+print(correlation_plot_N0)
+
+```
+
+### One year ago
+
+```{r}
+
+# combine data fames
+comparison_df_N1 <- N_yearly_avg %>%
   left_join(avg_swe_year_prior, by = "Year")
 
-# plot
-abundance_WinterSeverity_1yr <- ggplot(comparison_df_1, aes(x = Year)) +
+###################### plot
+WinterSeverity_N1 <- ggplot(comparison_df_N1, aes(x = Year)) +
   # frog abundance line
@@
     y = "Value"
-  )
+  ) +
+  theme(legend.position = c(0.8, 0.2))
 
-# correlation
-correlation1 <- cor(comparison_df_1$Mean_Abundance, comparison_df_1$Winter_Severity)
-cor_test1 <- cor.test(comparison_df_1$Mean_Abundance, comparison_df_1$Winter_Severity)
+###################### correlation
+correlation_N1 <- cor(comparison_df_N1$Mean_Abundance, comparison_df_N1$Winter_Severity)
+cor_test_1_N <- cor.test(comparison_df_N1$Mean_Abundance, comparison_df_N1$Winter_Severity)
 
-correlation_plot1 <- ggplot(data = comparison_df_1, aes(x = Winter_Severity, y = Mean_Abundance)) +
+correlation_plot_N1 <- ggplot(data = comparison_df_N1, aes(x = Winter_Severity, y = Mean_Abundance)) +
   geom_point() +
   geom_smooth(method = "lm" ,se = FALSE) +
-  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation1, 2)), size = 5, color = "red") +
+  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_N1, 2)), size = 5, color = "red") +
   labs(
@@
 
+###################### print
+print(WinterSeverity_N1)
+print(correlation_plot_N1)
+
+ggsave("WinterSeverityN1.png", plot = WinterSeverity_N1, width = 7, height = 4, bg = "white")
+ggsave("CorrelationN1.png", plot = correlation_plot_N1, width = 7, height = 4, bg = "white")
+
+N1_arrange <- ggarrange(WinterSeverity_N1, correlation_plot_N1)
+ggsave("test12345.png", plot = N1_arrange, width = 10, bg = "white")
+
+```
+
+### Two Years Ago
+
+```{r}
+
+# combine data fames
+comparison_df_2_N <- N_yearly_avg %>%
+  left_join(avg_swe_2year_prior, by = "Year")
+
+# plot
+abundance_WinterSeverity_2yr <- ggplot(comparison_df_2_N, aes(x = Year)) +
+  # frog abundance line
+  geom_line(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 1) +
+  geom_point(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 2) +
+  # winter severity line
+  geom_line(aes(y = Winter_Severity, color = "Winter Severity Two Years Prior\n(% of historical avg 1 April SWE)"), linetype = "dashed", size = 1) +
+  geom_point(aes(y = Winter_Severity, color = "Winter Severity Two Years Prior\n(% of historical avg 1 April SWE)"), size = 2) +
+  # Labels and theme
+  scale_color_manual(
+    name = "Legend",
+    values = c("Frog Abundance" = "blue", "Winter Severity Two Years Prior\n(% of historical avg 1 April SWE)" = "red")
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Frog Abundance and Winter Severity Two Years Prior",
+    x = "Year",
+    y = "Value"
+  )
+
+# correlation
+correlation_2_N <- cor(comparison_df_2_N$Mean_Abundance, comparison_df_2_N$Winter_Severity)
+cor_test_2_N <- cor.test(comparison_df_2_N$Mean_Abundance, comparison_df_2_N$Winter_Severity)
+
+correlation_plot_2_N <- ggplot(data = comparison_df_2_N, aes(x = Winter_Severity, y = Mean_Abundance)) +
+  geom_point() +
+  geom_smooth(method = "lm" ,se = FALSE) +
+  annotate("text", x = 50, y = 185, label = paste0("Correlation: ", round(correlation_2_N, 2)), size = 5, color = "red") +
+  labs(
+    title = "Correlation between Winter Severity Two Years Prior and Abundance",
+    x = "Winter Severity (% of Historical Average 1 April SWE)",
+    y = "Yearly Mean Abundance"
+  ) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+```
+
+### Three-year rolling average
+
+```{r}
+
+# combine data fames
+comparison_df_N3 <- N_yearly_avg %>%
+  left_join(avg_swe_3yrAvg, by = "Year")
+
+# plot
+WinterSeverity_N3 <- ggplot(comparison_df_N3, aes(x = Year)) +
+  # frog abundance line
+  geom_line(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 1) +
+  geom_point(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 2) +
+  # winter severity line
+  geom_line(aes(y = swe_3yrAvg, color = "Winter Severity 3 yr avg\n(% of historical avg 1 April SWE)"), linetype = "dashed", size = 1) +
+  geom_point(aes(y = swe_3yrAvg, color = "Winter Severity 3 yr avg\n(% of historical avg 1 April SWE)"), size = 2) +
+  # Labels and theme
+  scale_color_manual(
+    name = "Legend",
+    values = c("Frog Abundance" = "blue", "Winter Severity 3 yr avg\n(% of historical avg 1 April SWE)" = "red")
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Frog Abundance and Winter Severity Same Year",
+    x = "Year",
+    y = "Value"
+  )
+
+# correlation
+correlation_N3 <- cor(comparison_df_N3$Mean_Abundance, comparison_df_N3$swe_3yrAvg)
+cor_test_N3 <- cor.test(comparison_df_N0$Mean_Abundance, comparison_df_N3$swe_3yrAvg)
+
+correlation_plot_N3 <- ggplot(data = comparison_df_N3, aes(x = swe_3yrAvg, y = Mean_Abundance)) +
+  geom_point() +
+  geom_smooth(method = "lm" ,se = FALSE) +
+  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_N3, 2)), size = 5, color = "red") +
+  labs(
+    title = "Correlation between 3-yr avg Winter Severity and Abundance",
+    x = "Winter Severity (% of Historical Average 1 April SWE)",
+    y = "Yearly Mean Abundance"
+  ) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+######################## Print
+print(WinterSeverity_N3)
+print(correlation_plot_N3)
+
+```
+### Three-year rolling average w/ 1-yr lag
+
+```{r}
+
+# combine data fames
+comparison_df_N3.1 <- N_yearly_avg %>%
+  left_join(avg_swe_3yrAvg_1yrLag, by = "Year")
+
+# plot
+WinterSeverity_N3.1 <- ggplot(comparison_df_N3.1, aes(x = Year)) +
+  # frog abundance line
+  geom_line(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 1) +
+  geom_point(aes(y = Mean_Abundance, color = "Frog Abundance"), size = 2) +
+  # winter severity line
+  geom_line(aes(y = swe_3yrAvg_1yrLag, color = "Winter Severity 3 yr avg 1 yr lag\n(% of historical avg 1 April SWE)"), linetype = "dashed", size = 1) +
+  geom_point(aes(y = swe_3yrAvg_1yrLag, color = "Winter Severity 3 yr avg 1 yr lag\n(% of historical avg 1 April SWE)"), size = 2) +
+  # Labels and theme
+  scale_color_manual(
+    name = "Legend",
+    values = c("Frog Abundance" = "blue", "Winter Severity 3 yr avg 1 yr lag\n(% of historical avg 1 April SWE)" = "red")
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Frog Abundance and Winter Severity Same Year",
+    x = "Year",
+    y = "Value"
+  )
+
+# correlation
+correlation_N3.1 <- cor(comparison_df_N3.1$Mean_Abundance, comparison_df_N3.1$swe_3yrAvg_1yrLag)
+cor_test_N3.1 <- cor.test(comparison_df_N0$Mean_Abundance, comparison_df_N3.1$swe_3yrAvg_1yrLag)
+
+correlation_plot_N3.1 <- ggplot(data = comparison_df_N3.1, aes(x = swe_3yrAvg_1yrLag, y = Mean_Abundance)) +
+  geom_point() +
+  geom_smooth(method = "lm" ,se = FALSE) +
+  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_N3.1, 2)), size = 5, color = "red") +
+  labs(
+    title = "Correlation between 3-yr avg Winter Severity with 1 yr lag and Abundance",
+    x = "Winter Severity (% of Historical Average 1 April SWE)",
+    y = "Yearly Mean Abundance"
+  ) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+######################## Print
+print(WinterSeverity_N3.1)
+print(correlation_plot_N3.1)
+
 ```
@@
 ## Recruitment - Winter Severity Comparison
+
+### Winter Same Year
+
+```{r}
+
+# combine data fames
+comparison_df_S_B <- B_yearly_avg %>%
+  left_join(avg_swe, by = "Year")
+
+# plot
+WinterSeverity_S_B <- ggplot(comparison_df_S_B, aes(x = Year)) +
+  # frog abundance line
+  geom_line(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 1) +
+  geom_point(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 2) +
+  # winter severity line
+  geom_line(aes(y = Winter_Severity, color = "Winter Severity Same Year"), linetype = "dashed", size = 1) +
+  geom_point(aes(y = Winter_Severity, color = "Winter Severity Same Year"), size = 2) +
+  # Labels and theme
+  scale_color_manual(
+    name = "Legend",
+    values = c("Individuals Recruited" = "blue", "Winter Severity Same Year" = "red")
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Recruitment and Same-Year Winter Severity",
+    x = "Year",
+    y = "Value"
+  )
+
+# correlation
+correlation_S_B <- cor(comparison_df_S_B$Winter_Severity, comparison_df_S_B$Recruited_Individuals)
+cor_test_S_B <- cor.test(comparison_df_S_B$Winter_Severity, comparison_df_S_B$Recruited_Individuals)
+
+correlation_plot_S_B <- ggplot(data = comparison_df_S_B, aes(x = Winter_Severity, y = Recruited_Individuals)) +
+  geom_point() +
+  geom_smooth(method = "lm" ,se = FALSE) +
+  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_S_B, 2)), size = 5, color = "red") +
+  labs(
+    title = "Correlation between Same-Year Winter Severity and Recruitment",
+    x = "Winter Severity (% of Historical Average 1 April SWE)",
+    y = "Total Yearly Recruitment (Number of individuals)"
+  ) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+print(WinterSeverity_S_B)
+print(correlation_plot_S_B)
+
+```
 
@@
 # combine data fames
-comparison_df_2 <- B_yearly_avg %>%
+comparison_df_B1 <- B_yearly_avg %>%
   left_join(avg_swe_year_prior, by = "Year")
@@
 # plot
-WinterSeverity_Recruitment <- ggplot(comparison_df_2, aes(x = Year)) +
+WinterSeverity_B1 <- ggplot(comparison_df_B1, aes(x = Year)) +
   # frog abundance line
@@
     y = "Value"
-  )
+  ) +
+  theme(legend.position = c(0.8, 0.2))
 
 # correlation
-correlation2 <- cor(comparison_df_2$Winter_Severity, comparison_df_2$Recruited_Individuals)
-cor_test2 <- cor.test(comparison_df_1$Winter_Severity, comparison_df_2$Recruited_Individuals)
+correlation_B1 <- cor(comparison_df_B1$Winter_Severity, comparison_df_B1$Recruited_Individuals)
+cor_test_B1 <- cor.test(comparison_df_B1$Winter_Severity, comparison_df_B1$Recruited_Individuals)
 
-correlation_plot2 <- ggplot(data = comparison_df_2, aes(x = Winter_Severity, y = Recruited_Individuals)) +
+correlation_plot_B1 <- ggplot(data = comparison_df_B1, aes(x = Winter_Severity, y = Recruited_Individuals)) +
   geom_point() +
   geom_smooth(method = "lm" ,se = FALSE) +
-  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation2, 2)), size = 5, color = "red") +
+  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_B1, 2)), size = 5, color = "red") +
   labs(
@@
   theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5),
+        )
+
+print(WinterSeverity_B1)
+print(correlation_plot_B1)
+
+ggsave("WinterSeverityB1.png", plot = WinterSeverity_B1, bg = "white")
+ggsave("CorrelationB1.png", plot = correlation_plot_B1, bg = "white")
+
+```
+
+### Winter 2 years ago
+
+```{r}
+
+# combine data fames
+comparison_df_2_B <- B_yearly_avg %>%
+  left_join(avg_swe_2year_prior, by = "Year")
+
+# plot
+WinterSeverity_2_B <- ggplot(comparison_df_2_B, aes(x = Year)) +
+  # frog abundance line
+  geom_line(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 1) +
+  geom_point(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 2) +
+  # winter severity line
+  geom_line(aes(y = Winter_Severity, color = "Winter Severity Two Years Prior\n(% of historical avg 1 April SWE)"), linetype = "dashed", size = 1) +
+  geom_point(aes(y = Winter_Severity, color = "Winter Severity Two Years Prior\n(% of historical avg 1 April SWE)"), size = 2) +
+  # Labels and theme
+  scale_color_manual(
+    name = "Legend",
+    values = c("Individuals Recruited" = "blue", "Winter Severity Two Years Prior\n(% of historical avg 1 April SWE)" = "red")
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Recruitment and Winter Severity Two Years Prior",
+    x = "Year",
+    y = "Value"
+  )
+
+# correlation
+correlation_2_B <- cor(comparison_df_2_B$Winter_Severity, comparison_df_2_B$Recruited_Individuals)
+cor_test_2_B <- cor.test(comparison_df_2_B$Winter_Severity, comparison_df_2_B$Recruited_Individuals)
+
+correlation_plot_2_B <- ggplot(data = comparison_df_2_B, aes(x = Winter_Severity, y = Recruited_Individuals)) +
+  geom_point() +
+  geom_smooth(method = "lm" ,se = FALSE) +
+  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_2_B, 2)), size = 5, color = "red") +
+  labs(
+    title = "Correlation between Winter Severity (2-Year lag) and Recruitment",
+    x = "Winter Severity (% of Historical Average 1 April SWE)",
+    y = "Total Yearly Recruitment (Number of individuals)"
+  ) +
+  theme_minimal() +
   theme(plot.title = element_text(hjust = 0.5))
+
+print(WinterSeverity_2_B)
+print(correlation_plot_2_B)
+
+```
+### Three-year rolling average no lag
+
+```{r}
+
+# combine data fames
+comparison_df_B3 <- B_yearly_avg %>%
+  left_join(avg_swe_3yrAvg, by = "Year")
+
+# plot
+WinterSeverity_B3 <- ggplot(comparison_df_B3, aes(x = Year)) +
+  # frog abundance line
+  geom_line(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 1) +
+  geom_point(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 2) +
+  # winter severity line
+  geom_line(aes(y = swe_3yrAvg, color = "Winter Severity 3 yr avg\n(% of historical avg 1 April SWE)"), linetype = "dashed", size = 1) +
+  geom_point(aes(y = swe_3yrAvg, color = "Winter Severity 3 yr avg\n(% of historical avg 1 April SWE)"), size = 2) +
+  # Labels and theme
+  scale_color_manual(
+    name = "Legend",
+    values = c("Individuals Recruited" = "blue", "Winter Severity 3 yr avg\n(% of historical avg 1 April SWE)" = "red")
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Recruitment and 3-yr avg Winter Severity",
+    x = "Year",
+    y = "Value"
+  )
+
+# correlation
+correlation_B3 <- cor(comparison_df_B3$Recruited_Individuals, comparison_df_B3$swe_3yrAvg)
+cor_test_B3 <- cor.test(comparison_df_B3$Recruited_Individuals, comparison_df_B3$swe_3yrAvg)
+
+correlation_plot_B3 <- ggplot(data = comparison_df_B3, aes(x = swe_3yrAvg, y = Recruited_Individuals)) +
+  geom_point() +
+  geom_smooth(method = "lm" ,se = FALSE) +
+  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_B3, 2)), size = 5, color = "red") +
+  labs(
+    title = "Correlation between 3-yr avg Winter Severity and Recruitment",
+    x = "Winter Severity (% of Historical Average 1 April SWE)",
+    y = "Yearly Mean Abundance"
+  ) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+######################## Print
+print(WinterSeverity_B3)
+print(correlation_plot_B3)
+
+```
+
+### Three-year rolling average w/ 1-yr lag
+
+```{r}
+
+# combine data fames
+comparison_df_B3.1 <- B_yearly_avg %>%
+  left_join(avg_swe_3yrAvg_1yrLag, by = "Year")
+
+# plot
+WinterSeverity_B3.1 <- ggplot(comparison_df_B3.1, aes(x = Year)) +
+  # frog abundance line
+  geom_line(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 1) +
+  geom_point(aes(y = Recruited_Individuals, color = "Individuals Recruited"), size = 2) +
+  # winter severity line
+  geom_line(aes(y = swe_3yrAvg_1yrLag, color = "Winter Severity 3 yr avg 1 yr lag\n(% of historical avg 1 April SWE)"), linetype = "dashed", size = 1) +
+  geom_point(aes(y = swe_3yrAvg_1yrLag, color = "Winter Severity 3 yr avg 1 yr lag\n(% of historical avg 1 April SWE)"), size = 2) +
+  # Labels and theme
+  scale_color_manual(
+    name = "Legend",
+    values = c("Individuals Recruited" = "blue", "Winter Severity 3 yr avg 1 yr lag\n(% of historical avg 1 April SWE)" = "red")
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Recruitment and 3-yr avg Winter Severity with 1-yr lag",
+    x = "Year",
+    y = "Value"
+  )
 
@@
 # Plot
-ggplot(comparison_df_3, aes(x = Winter)) +
+WinterSeverity_Survival <- ggplot(comparison_df_3, aes(x = Winter)) +
   # survival line
@@
 
+# correlation
+correlation3 <- cor(comparison_df_3$Winter_Severity, comparison_df_3$Overwinter_Survival_Probability)
+cor_test3 <- cor.test(comparison_df_3$Winter_Severity, comparison_df_3$Overwinter_Survival_Probability)
+
+correlation_plot3 <- ggplot(data = comparison_df_3, aes(x = Winter_Severity, y = Survival_Percentage)) +
+  geom_point() +
+  geom_smooth(method = "lm" ,se = FALSE) +
+  annotate("text", x = 75, y = 100, label = paste0("Correlation: ", round(correlation3, 2)), size = 5, color = "red") +
+  labs(
+    title = "Correlation between Winter Severity and Survival",
+    x = "Winter Severity (% of Historical Average 1 April SWE)",
+    y = "Overwinter Survuval Probability"
+  ) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))
+
 ```

### 2025-01-20 doc/Manuscript.qmd
begin introduction

@@
 ---
-title: "manuscript"
-format: html
-editor: visual
+format: pdf
+bibliography: references.bib
 ---
@@
 
+Amphibians are the most threatened class of vertebrates.
+Disease, habitat loss, and, increasingly, climate change drive their global decline [@luedtke2023].
+The endangered mountain yellow-legged frog species complex (*Rana muscosa* and *R.
+sierrae*) is emblematic of such declines.
+**Citation**.
+These frogs are endemic to the Sierra Nevada and Transverse mountain ranges of California and Nevada, and are characterized by \_\_\_\_\_\_.
+Non-native fish introductions across their range have severely reduced suitable habitat for the species[@knapp2000], and the fungal pathogen *Batrachochytrium dendrobatadis* (Bd) causes mass mortality and rapid population declines [@vredenburg2010].
+As a result, more than 90% of *R.
+muscosa* and *R.
+sierrae* populations have been extirpated from their historical range [@vredenburg2007].
+
+Despite these substantial losses, mountain yellow-legged frogs provide rare examples of amphibian resilience in the face of multiple stressors.
+Following the removal of non-native fish from alpine lakes, *R.
+muscosa* population density can increase significantly [@vredenburg2004][@knapp2007].
+Additionally, some mountain yellow-legged frog populations exhibit resistance to Bd infection and naturally recover despite the pathogen's continued presence in the environment [@knapp2016].
+These Bd-persistent populations are increasingly important to recovery efforts for the species.
+A recent study shows that reintroducing frogs from persistent populations into vacant habitat can establish self-sustaining populations with low long-term extinction probability [@knapp2024].
+It is critical, therefore, to study the dynamics of these unique populations to gain insights into how they are persisting in the presence of Bd and to assess their value to future conservation efforts.
+
+Among Bd-persistent mountain yellow-legged frog populations, little is known about the ecology and demographics of those in the northern part of their range [@brown2019].
+Compared to their relatively well-studied counterparts in the central and southern Sierra Nevada, northern populations are fewer in number, have lower abundances, utilize a more diverse range of habitat, and occupy lower elevations [@interagency2018].
+To implement effective conservation measures for mountain yellow-legged frogs in the northern portion of their range, it is vital to better understand the demographics and dynamics of these rare and understudied disease-persistent populations [@brown2020].
+
+In this study, we describe one such population, located in the \_\_\_\_\_\_\_\_\_\_ area of the Tahoe National Forest, California...
+
 ### Methods

### 2025-01-26 doc/Manuscript.qmd
Begin Methods, taken from HML memo

@@
 
-Among Bd-persistent mountain yellow-legged frog populations, little is known about the ecology and demographics of those in the northern part of their range [@brown2019].
+Among Bd-persistent mountain yellow-legged frog populations, little is known about the ecology and demographics of those in the northern part of their range [@interagency2018][@brown2019].
 Compared to their relatively well-studied counterparts in the central and southern Sierra Nevada, northern populations are fewer in number, have lower abundances, utilize a more diverse range of habitat, and occupy lower elevations [@interagency2018].
@@
 
-In this study, we describe one such population, located in the \_\_\_\_\_\_\_\_\_\_ area of the Tahoe National Forest, California...
+In this study, we describe one such population, located in the Mossy Pond area of the Tahoe National Forest, California...
 
 ### Methods
+
+**Study area and species**
+
+The Mossy Pond Complex is located in the Tahoe National Forest, north of highway 80 in Nevada County, California.
+It consists of approximately 60 lakes, ponds, and streams set on granite benches southeast of Fordyce Lake.
+Elevation ranges from 6,400 feet to 7,100 feet.
+The study area is a one-square-mile section of the complex, containing the namesake Mossy Pond, its seasonally-flowing outlet stream, and 12 ephemeral ponds.
+**\*\*Rationale for study area selection\*\*.** The United States Geological Survey (USGS) first detected SNYLF in the complex in 1998, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.
+
+**Capture-mark-recapture sampling**
+
+*Sampling schedule*
+
+\
+
+*Capturing & marking animals*
+
+Field crews captured frogs by hand or dip net, and processed frogs at the point of capture.
+Crews first scanned each captured frog with a BioMark 601 passive integrated transponder (PIT) tag reader to detect if the frog was marked (i.e., a recapture).
+Crews then used calipers to measure the snout to urostyle (SUL) length of each captured frog, and released frogs \<40 mm SUL (which are considered subadults) without further processing.
+Crews continued collecting data on larger (\>40 mm SUL) frogs, which are considered adults.
+For new adult captures, crews inserted an 8 X 1.4 mm PIT tag under the dorsal skin using methods recommended by McAllister et al.
+(2004).
+Crews collected a GPS point (estimated precision error \~3 m) for each capture using a handheld Garmin GPS unit.
+Crews weighed each adult frog inside a tared plastic bag to the nearest 0.1 g using a Pesola spring scale.
+For each captured adult individual, crews only collected length and weight data during the first capture event within a three-day primary period.
+Afterward, if crews captured the same individual on a subsequent day duringthe same primary period, crews only recorded PIT tag, sex, and location data.
+
+**Population modeling**
+
+\
+\
+**Winter severity calculations**
 

### 2025-01-29 doc/Manuscript.qmd
continue methods: robust design

@@
 
-## Title Title Title
+# Title Title Title
 
@@
 
-### Abstract
+## Abstract
 
-### Introduction
+## Introduction
 
@@
 
-### Methods
+## Methods
 
-**Study area and species**
+### Study area and species
 
@@
 
-**Capture-mark-recapture sampling**
+### Capture-mark-recapture surveys
 
-*Sampling schedule*
-
-\
-
-*Capturing & marking animals*
+Capture-recapture sampling was conducted according to Pollock's [@pollock1982] robust design.
+Under the robust design, three primary survey periods occurred each summer from 2014 to 2018, each consisting of three consecutive survey days (secondary periods).
 
@@
 For new adult captures, crews inserted an 8 X 1.4 mm PIT tag under the dorsal skin using methods recommended by McAllister et al.
-(2004).
+(2004) [@mcallister2004].
 Crews collected a GPS point (estimated precision error \~3 m) for each capture using a handheld Garmin GPS unit.
@@
 For each captured adult individual, crews only collected length and weight data during the first capture event within a three-day primary period.
-Afterward, if crews captured the same individual on a subsequent day duringthe same primary period, crews only recorded PIT tag, sex, and location data.
+Afterward, if crews captured the same individual on a subsequent day during the same primary period, crews only recorded PIT tag, sex, and location data.
 
-**Population modeling**
+### Population modeling
 
 \
+
+### Winter severity calculations
+
 \
-**Winter severity calculations**
 
-### Results
+## Results
 
-### References
+## References
 
-### Figures
+## Figures

### 2025-01-29 doc/Manuscript.qmd
Continue methods. Modeling section.

@@
 
-Capture-recapture sampling was conducted according to Pollock's [@pollock1982] robust design.
+Capture-mark-recapture (CMR) sampling was conducted according to Pollock's [@pollock1982] robust design.
 Under the robust design, three primary survey periods occurred each summer from 2014 to 2018, each consisting of three consecutive survey days (secondary periods).
+**More needed?**
 
@@
 
-\
+We estimated survival probability, detection probability, recruitment of new frogs into the adult population, and adult population size for the Mossy Pond *R.
+sierrae* population using a site-specific Jolly-Seber CMR model implemented in a Bayesian framework with R statistical software [@R](v4.4.2).
+Our model was adapted from the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), which was developed for the analysis of robust design CMR data collected from *R.
+sierrae* populations.
+The model represents a superpopulation of M individuals whose states (not recruited, alive, dead) and observations (detected, not detected) evolve through time, with any new frogs assumed to arise from local reproduction [@knapp2024].
+We assessed model convergence via trace plots and Gelman–Rubin statistics (rhat).
+
+Models copmpared with LOO cross-validation.
 

### 2025-01-31 code/R/Survival_Probability_Difference_MF.Rmd
calculates and plots M-F difference in survival

---
title: "Survival Probability Difference M-F"
output: html_document
date: "2025-01-30"
---

```{r}

library(data.table)
library(tidyverse)
library(ggplot2)

```

## Extract draws from the model with sex as a survival covariate

```{r}

# laod the model with sex as a covariate of survival
mod <- readRDS("/Users/johnimperato/Desktop/Mossy-Pond_companion/sex_survival_covariate_model.rds")

# extract the draws of the relevant variables
phi_draws <- as.data.table(mod_sex$m_fit$draws(variables = c('beta_phi', 'eps_phi'), format = 'draws_df'))

# add indexing row
phi_draws <- phi_draws %>%
  mutate(draw_id = row_number()) %>%
  select(draw_id, everything())

```

```{r}

# Identify all 'eps_phi[t]' columns using regex
eps_phi_cols <- grep("^eps_phi\\[", names(phi_draws), value = TRUE)

# Extract 'eps_phi[t]' into a separate matrix (M x T)
eps_phi_matrix <- as.matrix(phi_draws %>% select(all_of(eps_phi_cols)))

# Verify dimensions (M = number of draws, T = number of time periods)
dim(eps_phi_matrix)  # Should return c(M, T)

# Extract 'beta_phi[1]' and 'beta_phi[2]' as numeric vectors
beta_phi1_draws <- phi_draws$`beta_phi[1]`
beta_phi2_draws <- phi_draws$`beta_phi[2]`

```

```{r}

# Determine number of posterior draws (M) and time periods (T)
M <- nrow(phi_draws)
T <- ncol(eps_phi_matrix)

# Initialize matrices to store survival probabilities and their differences
female_survival <- matrix(NA, nrow = M, ncol = T)
male_survival   <- matrix(NA, nrow = M, ncol = T)
diff_survival   <- matrix(NA, nrow = M, ncol = T)

# Assign column names for clarity (optional)
colnames(female_survival) <- paste0("time_", 1:T)
colnames(male_survival)   <- paste0("time_", 1:T)
colnames(diff_survival)   <- paste0("time_", 1:T)

# Compute survival probabilities and their differences
for(m in 1:M){
  for(t in 1:T){
    # Logit-scale survival for females: beta_phi[1] + eps_phi[t]
    logit_female <- beta_phi1_draws[m] + eps_phi_matrix[m, t]

    # Logit-scale survival for males: beta_phi[1] + beta_phi[2] + eps_phi[t]
    logit_male <- beta_phi1_draws[m] + beta_phi2_draws[m] + eps_phi_matrix[m, t]

    # Convert logits to probabilities using the logistic function
    p_female <- 1 / (1 + exp(-logit_female))
    p_male   <- 1 / (1 + exp(-logit_male))

    # Store the probabilities
    female_survival[m, t] <- p_female
    male_survival[m, t]   <- p_male

    # Compute and store the difference (male - female)
    diff_survival[m, t]   <- p_male - p_female
  }
}

```

```{r}

# Define a helper function to compute summary statistics
summarize_draws <- function(x) {
  data.frame(
    mean          = mean(x),
    median        = median(x),
    sd            = sd(x),
    lower95       = quantile(x, 0.025),
    upper95       = quantile(x, 0.975),
    prob_positive = mean(x > 0)   # Proportion of draws where male > female
  )
}

# Apply the summary function to each time period's difference
time_summaries_list <- lapply(1:T, function(t) {
  summarize_draws(diff_survival[, t])
})

# Combine the summaries into a single data frame
time_summaries_df <- bind_rows(time_summaries_list, .id = "time_index") %>%
  mutate(time_period = paste0("time_", time_index)) %>%
  select(time_period, everything())

# View the summaries
print(time_summaries_df)

```

```{r}

# Plot the difference with credible intervals
ggplot(time_summaries_df, aes(x = time_period, y = mean)) +
  geom_point(color = "blue") +
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.2, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Difference in Survival Probabilities (Male - Female) Over Time",
    x = "Time Period",
    y = "Difference in Survival Probability"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}

###############################################################################
# 5. Plot the Differences with Correct Time Order
###############################################################################

# Ensure 'time_period' is an ordered factor (if not already done)
time_summaries_df <- time_summaries_df %>%
  mutate(time_period = factor(time_period, 
                              levels = paste0("time_", 1:T),  # Define the correct chronological order
                              ordered = TRUE))

# Verify the ordering
print(levels(time_summaries_df$time_period))

# Create the plot
difference_plot <- ggplot(time_summaries_df, aes(x = time_period, y = mean)) +
  geom_point(color = "blue", size = 2) +  # Increased point size for better visibility
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.2, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Difference in Survival Probabilities (Male - Female) Over Time",
    x = "Time Period",
    y = "Difference in Survival Probability"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Tilt x-axis labels for readability
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")  # Center and style the title
  ) +
  scale_x_discrete(limits = paste0("time_", 1:T))  # Ensure correct order on x-axis

# Display the plot
print(difference_plot)

# Save the plot to a file
ggsave("difference_in_survival_probabilities.png", plot = difference_plot, width = 10, height = 6, dpi = 300)

```

### 2025-01-31 code/R/Survival_Probability_Difference_MF.Rmd
formatting update

@@
 
-# Plot the difference with credible intervals
-ggplot(time_summaries_df, aes(x = time_period, y = mean)) +
-  geom_point(color = "blue") +
-  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.2, color = "blue") +
-  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
-  labs(
-    title = "Difference in Survival Probabilities (Male - Female) Over Time",
-    x = "Time Period",
-    y = "Difference in Survival Probability"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-```
-
-```{r}
-
-###############################################################################
-# 5. Plot the Differences with Correct Time Order
-###############################################################################
+# plot
 
@@
   ) +
-  theme_minimal() +
   theme(
@@
 # Save the plot to a file
-ggsave("difference_in_survival_probabilities.png", plot = difference_plot, width = 10, height = 6, dpi = 300)
+ggsave("/Users/johnimperato/Desktop/Mossy-Pond_companion/difference_in_survival_probabilities.png", plot = difference_plot, width = 10, height = 6, dpi = 300)
 

### 2025-01-31 doc/Manuscript.qmd
draft new intro

@@
 ## Introduction
+
+The well-documented decline of the endangered Sierra Nevada yellow-legegd frog in its namesake mountain range of California (bunch of citations) is emblematic of the crisis facing amphibians globally (citations).
+As is the case with numerous other species..the fungal pathogen Bd is the primary driver of SNYLF decline in the Sierra Nevada (citations), and the effects of climate change loom large, especially in an alpine environment expriencing accelerated climate change (mountain ranges are warming faster than other places?) (citation).
+
+Though SNYLF decline fits the gloomy narrative, the examples of resilience and recovery that this species provides do not.
+Bd will typically wipe out (extirpate?) SNYLF populations (Vredenburg?), but a rare subset of individuals exhibit resistance to the pathogen ('resistant individuals') and contribute to the natural recovery of populations that persist in the presence of Bd (Knapp).
+These disease-persistent populations have become instrumental to conservation efforts for this species in their provision of resistant individuals as founders to establish new, resilient populations in previously-occupied habitat (Knapp 2024).
+We have the chance to pull off a rare feat: landscape scale recovery of an endangered species.
+
+To successfully leverage these disease-persistent populations as a conservation resource, we must gain a detailed understanding of their demographics, dynamics, and the factors that influence their viability.
+Without good information, we can't make smart decisions and we will waste limited resources (Brown xxxx).
+
+The population we focus on in this study-Mossy Pond-provides an excellent opportunity to fill the void of information surrounding this species in the northern portion of its range (cite interagency 2018 and Brown).
+Weird habitat, funky frogs, and little knowledge, etc..
+
+In this investigation we set out to estimate demographic parameters (getting absolute abundance is really important, as Os for citations on that) and to investigate the rolethat climate plays in population viability.
+Ultimately, we hope our insights are useful for the management of this paprticualr population, for a deeper understanding of the species as a whole, and as a hint of how climate change may affect alpine amphibian populations in the future.
+
+--------------------------------
+
+By acting as donor populations for the human-facilitated reintroduction of disease-resistant animals into extirpated habitat, these unique Bd persistent populations have become instrumental to conservation efforts.
+
+Some populations of MYLF exhibit resistence to Bd and will recover naturally (citations).
+These naturally-recovering populations (persistent populations) have become instrumental to human-facilitated conservation efforts, and present an opportunity for landscape scale recovery of an endangered species (Knapp 2024).
+
+What effects may climate have on these recovery efforts?
+
+-------------------------------------------
 
@@
 We estimated survival probability, detection probability, recruitment of new frogs into the adult population, and adult population size for the Mossy Pond *R.
-sierrae* population using a site-specific Jolly-Seber CMR model implemented in a Bayesian framework with R statistical software [@R](v4.4.2).
+sierrae* population using a site-specific Jolly-Seber CMR model implemented in a Bayesian framework with R statistical software [\@R](v4.4.2).
 Our model was adapted from the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), which was developed for the analysis of robust design CMR data collected from *R.

### 2025-01-31 code/R/Survival_Probability_Difference_MF.Rmd
update figure

@@
 M <- nrow(phi_draws)
-T <- ncol(eps_phi_matrix)
+T <- (ncol(eps_phi_matrix))
 
@@
 
-# Define a helper function to compute summary statistics
+# function to compute summary statistics
 summarize_draws <- function(x) {
@@
 # Apply the summary function to each time period's difference
-time_summaries_list <- lapply(1:T, function(t) {
+time_summaries_list <- lapply(2:(T-1), function(t) {
   summarize_draws(diff_survival[, t])
@@
 
+# Convert from factor to numeric via character
+time_summaries_df <- time_summaries_df %>%
+  mutate(time_index = as.numeric(as.character(time_index)))
+
 # Combine the summaries into a single data frame
 time_summaries_df <- bind_rows(time_summaries_list, .id = "time_index") %>%
-  mutate(time_period = paste0("time_", time_index)) %>%
-  select(time_period, everything())
+  mutate(time_index = as.numeric(time_index)) %>%
+  mutate(primary_period = paste("PPs_", time_index, "-", time_index + 1)) %>%
+  select(primary_period, everything())
 
-# View the summaries
+# View the summaries# Viewpaste0() the summaries
 print(time_summaries_df)
@@
 
-# plot
-
-# Ensure 'time_period' is an ordered factor (if not already done)
+# Ensure 'primary_period' is an ordered factor to maintain chronological order in the plot
 time_summaries_df <- time_summaries_df %>%
-  mutate(time_period = factor(time_period, 
-                              levels = paste0("time_", 1:T),  # Define the correct chronological order
+  mutate(primary_period = factor(primary_period, 
+                                 levels = unique(primary_period), 
                                  ordered = TRUE))
@@
 # Verify the ordering
-print(levels(time_summaries_df$time_period))
+print(levels(time_summaries_df$primary_period))
 
 # Create the plot
-difference_plot <- ggplot(time_summaries_df, aes(x = time_period, y = mean)) +
+difference_plot <- ggplot(time_summaries_df, aes(x = primary_period, y = mean)) +
   geom_point(color = "blue", size = 2) +  # Increased point size for better visibility
@@
   labs(
-    title = "Difference in Survival Probabilities (Male - Female) Over Time",
-    x = "Time Period",
+    title = "Difference in Survival Probabilities (M - F) Between Primary Periods",
+    x = "Primary Period",
     y = "Difference in Survival Probability"
@@
   ) +
-  scale_x_discrete(limits = paste0("time_", 1:T))  # Ensure correct order on x-axis
+  # Correct the x-axis limits to match 'primary_period' labels
+  scale_x_discrete(limits = levels(time_summaries_df$primary_period))
 
@@
 # Save the plot to a file
-ggsave("/Users/johnimperato/Desktop/Mossy-Pond_companion/difference_in_survival_probabilities.png", plot = difference_plot, width = 10, height = 6, dpi = 300)
-
+ggsave("/Users/johnimperato/Desktop/Mossy-Pond_companion/difference_in_survival_probabilities.png", 
+       plot = difference_plot, width = 10, height = 6, dpi = 300)
 ```

### 2025-01-31 doc/Manuscript.qmd
continue introduction

@@
 
-The well-documented decline of the endangered Sierra Nevada yellow-legegd frog in its namesake mountain range of California (bunch of citations) is emblematic of the crisis facing amphibians globally (citations).
-As is the case with numerous other species..the fungal pathogen Bd is the primary driver of SNYLF decline in the Sierra Nevada (citations), and the effects of climate change loom large, especially in an alpine environment expriencing accelerated climate change (mountain ranges are warming faster than other places?) (citation).
+The well-documented decline of the endangered Sierra Nevada yellow-legegd frog in its namesake mountain range (bunch of citations) is emblematic of the crisis facing amphibians globally (citations).
+As is the case with numerous other species...the fungal pathogen Bd is the primary driver of SNYLF decline in the Sierra Nevada (citations), and the effects of climate change loom large-yet remain understudied- especially in an alpine environment expriencing accelerated climate change (mountain ranges are warming faster than other places?) (citation).
 
@@
 
-To successfully leverage these disease-persistent populations as a conservation resource, we must gain a detailed understanding of their demographics, dynamics, and the factors that influence their viability.
+To successfully leverage these disease-persistent populations as conservation resources, we must gain a detailed understanding of their demographics, dynamics, and the factors that influence their viability.
 Without good information, we can't make smart decisions and we will waste limited resources (Brown xxxx).
 
-The population we focus on in this study-Mossy Pond-provides an excellent opportunity to fill the void of information surrounding this species in the northern portion of its range (cite interagency 2018 and Brown).
-Weird habitat, funky frogs, and little knowledge, etc..
+The population we focus on in this study-Mossy Pond-provides an excellent opportunity to fill the void of information around disease-persistent populations of this species in the northern portion of its range (cite interagency 2018 and Brown).
+Pop description: Weird habitat, funky frogs, and little knowledge, etc..
+Complexity of habitat here is representative of northern pops generally: streams, complexity, etc.
 
-In this investigation we set out to estimate demographic parameters (getting absolute abundance is really important, as Os for citations on that) and to investigate the rolethat climate plays in population viability.
+In this investigation we set out to estimate demographic parameters (getting absolute abundance is really important, as Os for citations on that) and to investigate the (factors that influence population viability in the Bd-persistent state) (role that climate plays in population viability).
 Ultimately, we hope our insights are useful for the management of this paprticualr population, for a deeper understanding of the species as a whole, and as a hint of how climate change may affect alpine amphibian populations in the future.

### 2025-02-01 doc/Manuscript.qmd
update introduction

@@
 The well-documented decline of the endangered Sierra Nevada yellow-legegd frog in its namesake mountain range (bunch of citations) is emblematic of the crisis facing amphibians globally (citations).
-As is the case with numerous other species...the fungal pathogen Bd is the primary driver of SNYLF decline in the Sierra Nevada (citations), and the effects of climate change loom large-yet remain understudied- especially in an alpine environment expriencing accelerated climate change (mountain ranges are warming faster than other places?) (citation).
+As is the case with numerous other species...the fungal pathogen Bd is a primary driver of SNYLF decline in the Sierra Nevada (citations), and the effects of climate change loom large-yet remain understudied- especially in an alpine environment expriencing accelerated climate change (mountain ranges are warming faster than other places?) (citation).
+
+-   get rid of climate *change* here. Introduce just climate somewhere in the intro, then get into climate *change* later, like int he discussion.
 

### 2025-04-12 doc/Manuscript.qmd
clean up intro

--- 2025-01-31 doc/Manuscript.qmd
@@
 bibliography: references.bib
+execute: 
+  echo: true
 ---
@@
 
-The well-documented decline of the endangered Sierra Nevada yellow-legegd frog in its namesake mountain range of California (bunch of citations) is emblematic of the crisis facing amphibians globally (citations).
-As is the case with numerous other species..the fungal pathogen Bd is the primary driver of SNYLF decline in the Sierra Nevada (citations), and the effects of climate change loom large, especially in an alpine environment expriencing accelerated climate change (mountain ranges are warming faster than other places?) (citation).
+Amphibians continue to experience unprecedented global declines, with infectious diseases emerging as primary drivers of biodiversity loss (Luedtke et al.
+2023).
+Chytridiomycosis, caused by the fungal pathogen *Batrachochytrium dendrobatidis* (Bd), has been directly implicated in the decline or extinction of approximately 200 amphibian species across six continents (Skerratt et al.
+2007).
+Amphibian populations exhibit highly variable responses to Bd infection; some appear relatively unaffected, others rapidly succumb to extinction, some persist at chronically low numbers, and a few rare cases demonstrate population recovery following severe declines (Lips 2016).
+The mechanisms underlying persistence or recovery in Bd-affected populations remain incompletely understood and likely involve complex interactions among environmental, evolutionary, immunological, and demographic factors (Scheele et al.
+2017).
 
-Though SNYLF decline fits the gloomy narrative, the examples of resilience and recovery that this species provides do not.
-Bd will typically wipe out (extirpate?) SNYLF populations (Vredenburg?), but a rare subset of individuals exhibit resistance to the pathogen ('resistant individuals') and contribute to the natural recovery of populations that persist in the presence of Bd (Knapp).
-These disease-persistent populations have become instrumental to conservation efforts for this species in their provision of resistant individuals as founders to establish new, resilient populations in previously-occupied habitat (Knapp 2024).
-We have the chance to pull off a rare feat: landscape scale recovery of an endangered species.
+The mountain yellow-legged frog species complex (*Rana sierrae* and *R.
+muscosa*), endemic to California's Sierra Nevada and Transverse mountain ranges, provides one of the most thoroughly documented examples of amphibian decline, persistence, and recovery in the face of Bd infection.
+Beginning in the late 1970s, Bd spread across the Sierra Nevada, causing widespread extirpations of mountain yellow-legged frog populations (Vredenburg et al.
+2010).
+By the early 2000s, these frogs had disappeared from over 90% of their historical range due to Bd impacts, compounded by habitat degradation and non-native fish introductions (Knapp 2000; Vredenburg et al.
+2007).
+However, unlike many other amphibian species affected by chytridiomycosis, the mountain yellow-legged frog story did not end with extinction.
+A small subset of populations survived their initial contact with Bd and, after decades of decline, began exhibiting increases in abundance and apparent resistance to infection (Knapp et al.
+2016).
+These dynamics align with patterns of evolutionary rescue (Carlson et al.
+2014), wherein a species rapidly acquires adaptive genetic variation that restores demographic viability in the presence of a persistent threat (Knapp et al.
+2024).
 
-To successfully leverage these disease-persistent populations as a conservation resource, we must gain a detailed understanding of their demographics, dynamics, and the factors that influence their viability.
-Without good information, we can't make smart decisions and we will waste limited resources (Brown xxxx).
+While research has thoroughly described this shift from decline to recovery at broader geographic scales, fine-grained demographic studies at the individual population level remain scarce.
+Understanding the mechanisms and ecological processes underlying population recovery requires detailed demographic data at fine spatial and temporal resolutions.
+Unfortunately, such studies are time and resource-intensive, resulting in relatively few well-documented examples.
+Furthermore, most detailed demographic studies have concentrated in the southern portion of the species' range, primarily within or south of Yosemite National Park (Fellers et al.
+2013; Joseph and Knapp 2018).
+Relatively little is known about population dynamics in the northern Sierra Nevada (MYLF ITT 2018), an area characterized by smaller, lower-elevation frog populations occupying more heterogeneous habitats (Brown et al.
+2020).
 
-The population we focus on in this study-Mossy Pond-provides an excellent opportunity to fill the void of information surrounding this species in the northern portion of its range (cite interagency 2018 and Brown).
-Weird habitat, funky frogs, and little knowledge, etc..
+To address these knowledge gaps, we conducted a five-year capture-mark-recapture (CMR) study of a Bd-persistent population of *R.
+sierrae* in California's Tahoe National Forest.
+Our objectives were to estimate demographic parameters critical for understanding population dynamics, including abundance, survival, recruitment, and detection probabilities.
+By quantifying these demographic parameters at a fine scale, we sought to elucidate the ecological processes supporting persistence and recovery in amphibian populations living with endemic Bd infection.
+Additionally, we examined how climatic variables influence these demographic parameters, as climate is expected to be a primary driver of population dynamics in these high-elevation systems where Bd has become an established component of the ecosystem.
 
-In this investigation we set out to estimate demographic parameters (getting absolute abundance is really important, as Os for citations on that) and to investigate the rolethat climate plays in population viability.
-Ultimately, we hope our insights are useful for the management of this paprticualr population, for a deeper understanding of the species as a whole, and as a hint of how climate change may affect alpine amphibian populations in the future.
-
---------------------------------
-
-By acting as donor populations for the human-facilitated reintroduction of disease-resistant animals into extirpated habitat, these unique Bd persistent populations have become instrumental to conservation efforts.
-
-Some populations of MYLF exhibit resistence to Bd and will recover naturally (citations).
-These naturally-recovering populations (persistent populations) have become instrumental to human-facilitated conservation efforts, and present an opportunity for landscape scale recovery of an endangered species (Knapp 2024).
-
-What effects may climate have on these recovery efforts?
-
--------------------------------------------
-
-Amphibians are the most threatened class of vertebrates.
-Disease, habitat loss, and, increasingly, climate change drive their global decline [@luedtke2023].
-The endangered mountain yellow-legged frog species complex (*Rana muscosa* and *R.
-sierrae*) is emblematic of such declines.
-**Citation**.
-These frogs are endemic to the Sierra Nevada and Transverse mountain ranges of California and Nevada, and are characterized by \_\_\_\_\_\_.
-Non-native fish introductions across their range have severely reduced suitable habitat for the species[@knapp2000], and the fungal pathogen *Batrachochytrium dendrobatadis* (Bd) causes mass mortality and rapid population declines [@vredenburg2010].
-As a result, more than 90% of *R.
-muscosa* and *R.
-sierrae* populations have been extirpated from their historical range [@vredenburg2007].
-
-Despite these substantial losses, mountain yellow-legged frogs provide rare examples of amphibian resilience in the face of multiple stressors.
-Following the removal of non-native fish from alpine lakes, *R.
-muscosa* population density can increase significantly [@vredenburg2004][@knapp2007].
-Additionally, some mountain yellow-legged frog populations exhibit resistance to Bd infection and naturally recover despite the pathogen's continued presence in the environment [@knapp2016].
-These Bd-persistent populations are increasingly important to recovery efforts for the species.
-A recent study shows that reintroducing frogs from persistent populations into vacant habitat can establish self-sustaining populations with low long-term extinction probability [@knapp2024].
-It is critical, therefore, to study the dynamics of these unique populations to gain insights into how they are persisting in the presence of Bd and to assess their value to future conservation efforts.
-
-Among Bd-persistent mountain yellow-legged frog populations, little is known about the ecology and demographics of those in the northern part of their range [@interagency2018][@brown2019].
-Compared to their relatively well-studied counterparts in the central and southern Sierra Nevada, northern populations are fewer in number, have lower abundances, utilize a more diverse range of habitat, and occupy lower elevations [@interagency2018].
-To implement effective conservation measures for mountain yellow-legged frogs in the northern portion of their range, it is vital to better understand the demographics and dynamics of these rare and understudied disease-persistent populations [@brown2020].
-
-In this study, we describe one such population, located in the Mossy Pond area of the Tahoe National Forest, California...
+These Bd-persistent populations have significant conservation implications for recovery efforts of this endangered species.
+Recent research has demonstrated that individuals from these naturally recovering populations can serve as effective founders for reintroduction programs, establishing self-sustaining populations with reduced extinction probability when translocated to previously occupied habitats (Knapp et al.
+2024).
+This approach leverages the apparent disease resistance that has evolved in these populations, potentially enabling landscape-scale recovery.
+Our demographic study provides critical information to support such conservation interventions, particularly for the understudied northern Sierra Nevada populations.
+By quantifying vital rates and identifying factors that influence population viability in Bd-endemic environments, our results can inform selection of source populations, optimal timing of translocations, and post-translocation monitoring protocols.
+This knowledge is essential for developing effective, evidence-based conservation strategies for mountain yellow-legged frogs throughout their range.
 
@@
 The study area is a one-square-mile section of the complex, containing the namesake Mossy Pond, its seasonally-flowing outlet stream, and 12 ephemeral ponds.
-**\*\*Rationale for study area selection\*\*.** The United States Geological Survey (USGS) first detected SNYLF in the complex in 1998, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.
+**\*\*Rationale for study area selection\*\*.**
+
+The United States Geological Survey (USGS) first detected SNYLF in the complex in 1998, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.
+
+### Bd testing
 
@@
 
-Capture-mark-recapture (CMR) sampling was conducted according to Pollock's [@pollock1982] robust design.
+Capture-mark-recapture (CMR) sampling was conducted according to Pollock's robust design [@pollock1982].
 Under the robust design, three primary survey periods occurred each summer from 2014 to 2018, each consisting of three consecutive survey days (secondary periods).
@@
 
-### Population modeling
+### Statistical Analysis
 
 We estimated survival probability, detection probability, recruitment of new frogs into the adult population, and adult population size for the Mossy Pond *R.
-sierrae* population using a site-specific Jolly-Seber CMR model implemented in a Bayesian framework with R statistical software [\@R](v4.4.2).
+sierrae* population using a site-specific Jolly-Seber CMR model implemented in a Bayesian framework with R statistical software \[[\@R](v4.4.2)\].
 Our model was adapted from the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), which was developed for the analysis of robust design CMR data collected from *R.
@@
 
-Models copmpared with LOO cross-validation.
+**Model selection and model comparison with LOO cross-validation.**
+
+To extend our analysis and investigate the environmental drivers of variation in our estimated population parameters, we employed a two-stage Bayesian framework.
+First, we obtained full posterior distributions of the values of key population parameters—recruitment, abundance, and survival—from our mark-recapture model (described above).
+In the second stage, we propagated this uncertainty into subsequent analyses by performing regression and correlation analyses on the full set of posterior draws.
+In one approach, we calculated the Pearson correlation coefficient between winter severity (lagged by one year) and recruitment for each draw, resulting in a posterior distribution of correlation coefficients that summarizes the strength and direction of the association.
+In another approach, we performed separate linear regressions on each draw to derive a distribution of regression coefficients.
+Both methods allowed us to integrate the uncertainty from the capture–recapture model into our inference on the effects of environmental variables, ensuring that our conclusions about the relationship between winter severity and recruitment reflect the full uncertainty of the underlying estimates.
+
+Our model tracks survival at the individual level, and we computed a population-wide probability of surviving the interval between primary periods by summing those individual probabilities and dividing by the total number of frogs in the superpopulation (*M*).
+
+To complete analyses in an annual framework (i.e., investigate the relationship between winter severity and recruitment), we needed to aggregate each aprameter to an annual value.
+....
+For recruitment, posterior draws were extracted from the model output and then aggregated by summing estimates across primary periods to yield annual values.
+For abundance,
 
@@
 
-\
+We defined winter severity as the percent of historical average snowpack—measured as snow-water equivalent (SWE) in inches—remaining on 1 April of a given year (e.g., 100% of historical average snow pack indicates an average-severity winter).
+We calculated the severity of each winter between 2011 and 2018 at 10 snow observation stations surrounding our study area, then used ArcGIS Pro (citation) to interpolate those values across the landscape and calculate the average pixel value in our study area (figure).
+Snow observation stations were between 5.1 and 19.5 km from the Mossy Pond study area, and the diversity of local conditions captured by these 10 stations insulated our measurement from site-specific factors that may have produced a non-representative measurement in any given year.
+All snow pack data was provided by the California Department of Water Resources, accessed via the California Data Exchange Center (CDEC--[http://cdec.water.ca.gov](http://cdec.water.ca.gov/)).
+
+Each of 10 snow monitoring stations surrounding the MPSA is used as a point estimate of winter severity, and those points are used to interpolate winter severity values across the landscape, providing a robust estimate of winter severity in the MPSA.\
 
 ## Results
+
+**Bd status**
+
+xx / xx samples tested positive for Bd, with low to moderate loads well below the lethal level described by Vredenburg et al.
+(2010).
+Infection levels in this zone are typical of other enzootic and recovered mountain yellow-legged frog populations (figure xx).
+
+**Abundance**
+
+We calculated the abundance of the Mossy Pond *R.
+sierrae* population for each of the 16 primary survey periods of the study.
+The lowest abundance in the population was observed in the final period, September 2018 (117 individuals, 95% Bayesian credible interval: 103-132) and a maximum of 245 individuals (95% credible interval 237-256) was observed during the fifth primary period, in July 2015.
+Across all primary periods, the mean abundance of the population was 168 individuals.
+The population grew from 2014 to 2016 and declined in 2017 and 2018.
+Within years, abundance typically peaked during the second of three yearly primary periods, which was most often in July.
+
+**Recruitment**
+
+A total of 629 frogs recruited into the adult population over the course of the study.
+Annual recruitment ranged from 60 individuals (95% credible interval: 45-78) in 2018 to 166 individuals (95% credible interval: 143-188) in 2015.
+
+**Survival**
+
+The probability of surviving between primary periods, averaged over all frogs, was relatively high (average \_\_\_) and constant (\_\_\_), excluding one outlier period over the 2015-2016 winter (prob = 0.57).
+
+**Detection**
+
+For each session (secondary period; n=47) of the study, we calculated the probability that an individual frog that is alive and available for detection is actually detected.
+Detection probability was lowest (median 0.26, CrI 0.2-0.31) during the sixteenth session (09 September, 2015) and highest (median 60.9, CrI: 0.54-0.67) during the forty-third session (07 August, 2018).
+The average detection probability over 47 individual surveys was 0.454.
+
+**Winter severity, Recruitment, and Abundance**
+
+Our study period coincided with high variation in winter severity at the Mossy Pond study area, with snowpack SWE ranging from 3.5% of historical average in 2015 to 150% in 2017.
+This natural variability provided an opportunity to explore how winter conditions influence population dynamics.
+Our analyses reveal a strong negative correlation between winter severity (lagged by one year) and the number of frogs recruited into the adult population (median correlation coefficient: -0.92; credible interval: –0.70 to -0.98).
+For example, the most severe winter in our study (2016–2017) corresponded to the lowest annual recruitment in 2018.
+
+## Discussion
+
+**Context: The *Rana sierrae / muscosa* story in 3 papers:**
+
+1\. [Vredenburg et al. 2010](https://www.pnas.org/doi/10.1073/pnas.0914111107) – *Bd*-induced decline of *R. sierrae*
+
+Careful monitoring of *R.
+sierrae* populations through *Bd* epizootic events from 2004-2008 reveals large-scale declines and extirpations.
+69 of 88 observed populations were wiped out by 2008 and the rest are expected to follow.
+Disease is the overwhelming driver of population dynamics in this system.
+
+2.  [Knapp at al. 2016](https://www.pnas.org/doi/full/10.1073/pnas.1600983113) – Persistence and natural recovery
+
+After decades of decline, *R.
+sierrae* in Yosemite National Park are increasing in abundance.
+The trend is observed at the landscape scale, across hundreds of populations, providing a rare example of amphibian resilience.
+A laboratory experiment reveals that resistance to *Bd* (reduced infection susceptibility) may be the mechanism enabling persistence and recovery.
+
+3.  [Knapp et al 2024](https://www.nature.com/articles/s41467-024-53608-4) – Translocations & reintroductions: harnessing natural *Bd-*resistance for human-assisted recovery
+
+Individuals from *Bd*-persistent populations exhibit resistance to infection, consistent with expectations under an “evolutionary rescue” scenario.
+A 15-year study shows that reintroducing resistant individuals from naturally recovering populations to unoccupied habitat is an effective method for *R.
+sierrae* recovery.
+ A majority of new populations exhibit successful reproduction and recruitment (becoming self-sustaining) and models show low 50-year extinction probabilities.
+
+**How does my work fit into this story?**
+
+1.  ***Bd*****-persistent populations: what’s going on?**
+
+-   To better understand the *R.
+sierrae* story (and the greater global story of amphibian-Bd dynamics) we need detailed information about how populations are persisting and recovering in the presence of *Bd*.
+
+-   Detailed demographic data from large persistent populations are rare in the literature.
+We need fine-scale information about population structure and dynamics not included in the landscape-level studies outlined above.
+
+-   This 5-year capture-mark-recapture study of the Mossy Pond *R.
+sierrae* population (Tahoe National Forest, CA) provides an excellent case study of a large persistent population in the northern extent of the species’ range.
+
+-   Relatively little is known about northern populations compared to their southern counterparts, adding another layer of usefulness to this study (expanded geographic representation).
+
+-   How do my results help to fill this information gap?
+
+2.  **Demonstrating a return to ‘normal’ ecological regulation**
+
+-   The most compelling finding of this study is the strong negative correlation between the number of individuals recruited into the adult population annually and winter severity (with a 1-year lag).
+The relationship holds for abundance and winter severity as well.
+
+-   This shows a shift from the disease-driven dynamics described by Vredenburg et al. 2010 to climate-driven dynamics, which is typical of ‘normal’ amphibian populations where *Bd* is not a factor.
+
+-   A paragraph on observed historical trends & future projections of winter severity in the Sierra Nevada. What might the future hold for these frogs under changing climatic conditions?
+
+-   Conservation applications
+
+-   Knapp et al.
+2024 show that large persistent populations are key to the cutting edge of frog conservation.
+They provide *bd-*resistant individuals for reintroductions, which can establish self-sustaining populations that exhibit the founders’ *Bd* resistance.
+
+-   Utilizing persistent populations as conservation resources requires decision making informed by detailed population studies.
+
+    -   When is the best time to remove individuals?
+
+    -   How many individuals can we remove?
+
+-   Specifically, the Mossy Pond population is a prime candidate for donation to the nearby 5-Lakes Basin, where CDFW recently finished a multi-year habitat restoration project.
+These results will help inform that project.
 

### 2025-04-17 doc/Manuscript.qmd
updates to all sections

@@
 
+*In amphibians, the what is traditionally referred to as a 'biodiversity crisis' is particualrly pronounced.
+Of xxx described amphibian species, xxx are declining and xxx have gone extinct.
+Stressors are numerous.*
+
+*Frame the paper around biotic resilience.
+Amphibians are not typically considered very resilient.
+We have so many stories of doom and extinction adn collapse.
+More generally, when thinking about the future coexistense of people and other species on this planet, we tend to discount species' ability to change and evolve rapidly.*
+
+*One example from the Sierra Nevada of California provides a cautiously hopeful example of evolved resistance to a highly-lethal pathogen that has caused numerous extinctions.* Probable mechanisms of recovery have been identified and the situation looks a lot like an evolutionary rescue scenario.
+
+*But this example shows us that capacity might be limited for evolution to rescue vertebrates that have imited dispersal ability.
+Enter translocations, reintroductions, and human-assisted recovery of species we deem worthy.
+This is a new fronteir in amphibian conservation.
+A real strong example of evolutionary rescue and successful methods for people to facilitate wider scale recovery beyond what is natural.* Outcomes have previously been mixed (citations), but we have a new shining example here in Yosemite National Park that provides hope for MYLF across the rest of their range, and also a blueprint for other declining amphibian species affected by Bd.
+
+*Recovery and resilience relies on these persistent populations, both in understanding the mechanisms and processes of persistence and recovery, and as conservation resources for human-assisted recovery strategies.*
+
+*The MYLF example provides an exciting opportunity to study the persistent populaitons of this species.
+Most studies have so far been t large scale, relying on VES and mroe coarse grained observations.
+CMR is expensive and hard to do.
+Thus, CMR studies of these persistent populations are limited to sub-sections fo the species' geographic range.
+We took the opportunitiy to do a detailed study of one of these rare and special persistent population to contribute to scientific knowledge of amphibian resilience and for practical conservation management.*
+
 Amphibians continue to experience unprecedented global declines, with infectious diseases emerging as primary drivers of biodiversity loss (Luedtke et al.
@@
 2007).
-Amphibian populations exhibit highly variable responses to Bd infection; some appear relatively unaffected, others rapidly succumb to extinction, some persist at chronically low numbers, and a few rare cases demonstrate population recovery following severe declines (Lips 2016).
+Amphibian populations exhibit highly variable responses to Bd infection; some appear relatively unaffected, others rapidly succumb and go extinct, some persist at chronically low numbers, and a few rare cases demonstrate population recovery following severe declines (Lips 2016).
 The mechanisms underlying persistence or recovery in Bd-affected populations remain incompletely understood and likely involve complex interactions among environmental, evolutionary, immunological, and demographic factors (Scheele et al.
@@
 muscosa*), endemic to California's Sierra Nevada and Transverse mountain ranges, provides one of the most thoroughly documented examples of amphibian decline, persistence, and recovery in the face of Bd infection.
-Beginning in the late 1970s, Bd spread across the Sierra Nevada, causing widespread extirpations of mountain yellow-legged frog populations (Vredenburg et al.
+Beginning in the 1970s, Bd began spreading across the Sierra Nevada, causing widespread extirpations of mountain yellow-legged frog populations (Vredenburg et al.
 2010).
@@
 2007).
-However, unlike many other amphibian species affected by chytridiomycosis, the mountain yellow-legged frog story did not end with extinction.
+However, unlike many other amphibian species that are highly-susceptible to Bd infection, the mountain yellow-legged frog's story did not end with extinction.
 A small subset of populations survived their initial contact with Bd and, after decades of decline, began exhibiting increases in abundance and apparent resistance to infection (Knapp et al.
@@
 Furthermore, most detailed demographic studies have concentrated in the southern portion of the species' range, primarily within or south of Yosemite National Park (Fellers et al.
-2013; Joseph and Knapp 2018).
+2013; Joseph and Knapp 2018; Knapp et al.
+2024).
 Relatively little is known about population dynamics in the northern Sierra Nevada (MYLF ITT 2018), an area characterized by smaller, lower-elevation frog populations occupying more heterogeneous habitats (Brown et al.
@@
 
-To address these knowledge gaps, we conducted a five-year capture-mark-recapture (CMR) study of a Bd-persistent population of *R.
+To address these knowledge gaps, we conducted a five-year capture-mark-recapture (CMR) study of a naturally recovering, Bd-persistent population of *R.
 sierrae* in California's Tahoe National Forest.
@@
 By quantifying these demographic parameters at a fine scale, we sought to elucidate the ecological processes supporting persistence and recovery in amphibian populations living with endemic Bd infection.
-Additionally, we examined how climatic variables influence these demographic parameters, as climate is expected to be a primary driver of population dynamics in these high-elevation systems where Bd has become an established component of the ecosystem.
+Additionally, we examined how climatic variables influence these demographic parameters, as climate is expected to be a primary driver of population dynamics in these high-elevation systems where Bd has become an established component of the ecosystem (citation for climate effects).
 
-These Bd-persistent populations have significant conservation implications for recovery efforts of this endangered species.
-Recent research has demonstrated that individuals from these naturally recovering populations can serve as effective founders for reintroduction programs, establishing self-sustaining populations with reduced extinction probability when translocated to previously occupied habitats (Knapp et al.
+Understanding the dynamics of these naturally recovering populations is particularly important given their significant potential to contribute to conservation efforts.
+Recent research has demonstrated that individuals from these populations can serve as effective founders for reintroduction programs, establishing self-sustaining populations with low long-term extinction probability when translocated to previously occupied habitats (Knapp et al.
 2024).
-This approach leverages the apparent disease resistance that has evolved in these populations, potentially enabling landscape-scale recovery.
+This approach leverages the apparent disease resistance that has evolved in these populations, facilitating recovery at ecologically-relevant scales.
 Our demographic study provides critical information to support such conservation interventions, particularly for the understudied northern Sierra Nevada populations.
 By quantifying vital rates and identifying factors that influence population viability in Bd-endemic environments, our results can inform selection of source populations, optimal timing of translocations, and post-translocation monitoring protocols.
-This knowledge is essential for developing effective, evidence-based conservation strategies for mountain yellow-legged frogs throughout their range.
 
@@
 
-The Mossy Pond Complex is located in the Tahoe National Forest, north of highway 80 in Nevada County, California.
+The Mossy Pond study area is located in the Tahoe National Forest, north of highway 80 in Nevada County, California.
 It consists of approximately 60 lakes, ponds, and streams set on granite benches southeast of Fordyce Lake.
@@
 
-### Bd testing
+### Quantifying Bd loads
 
-### Capture-mark-recapture surveys
+To investigate the Mossy Pond *R.
+sierrae* population's Bd status, we collected epithelial swabs from 24 adult individuals over three years (2008, 2010, and 2021).
+Partner scientists quantified the amount of Bd DNA on each swab using real-time quantitative polymerase chain reaction (qPCR) analysis (Boyle et al.
+2004).
+We report infection loads as the number of ITS1 (Internal Transcribed Spacer 1) gene copies detected (Longo et al.
+2013).
 
-Capture-mark-recapture (CMR) sampling was conducted according to Pollock's robust design [@pollock1982].
-Under the robust design, three primary survey periods occurred each summer from 2014 to 2018, each consisting of three consecutive survey days (secondary periods).
-**More needed?**
+Quantifying Bd loads rather than simply testing for Bd presence allowed us to compare our study population with other known recovering populations.
+Partner scientists provided *R.
+sierrae* Bd load data from three naturally recovering populations in Yosemite National Park for this comparison (Knapp et al.
+2024).
 
-Field crews captured frogs by hand or dip net, and processed frogs at the point of capture.
-Crews first scanned each captured frog with a BioMark 601 passive integrated transponder (PIT) tag reader to detect if the frog was marked (i.e., a recapture).
-Crews then used calipers to measure the snout to urostyle (SUL) length of each captured frog, and released frogs \<40 mm SUL (which are considered subadults) without further processing.
-Crews continued collecting data on larger (\>40 mm SUL) frogs, which are considered adults.
-For new adult captures, crews inserted an 8 X 1.4 mm PIT tag under the dorsal skin using methods recommended by McAllister et al.
+### Frog population surveys
+
+We studied *R.
+sierrae* with capture-mark-recapture (CMR) methods, which are widely used for estimating wildlife population parameters such as abundance, survival, and recruitment (Williams et al.
+2001).
+We conducted CMR sampling according to Pollock's robust design [@pollock1982], which involves sampling at two temporal scales: primary periods (between which the population is open to demographic change) and secondary periods, during which the population is assumed closed.
+Each summer from 2014 to 2018, we visited the study site 3-4 times (primary periods) for three consecutive days of survey (secondary periods), for a total of 47 individual surveys.
+
+Surveys occurred generally from 0900 to 1700, when *R.
+sierrae* are most active.
+During these hours, we continuously walked through and around all water bodies in the study area looking for frogs.
+We captured frogs by hand or dip net and processed them at the point of capture.
+We first scanned each captured frog with a BioMark 601 passive integrated transponder (PIT) tag reader (BioMark, Boise, Idaho, USA) to detect if the frog was marked (i.e., a recapture).
+We then used calipers to measure the snout to urostyle (SUL) length of each captured frog, and released frogs \<40 mm SUL (which are considered subadults) without further processing.
+**Frogs larger than 40mm (adults) required further processing.** *We continued collecting data on larger (\>40 mm SUL) frogs, which are considered adults*.
+For new adult captures, we inserted an 8 X 1.4 mm PIT tag (BioMark, Boise, Idaho, USA) under the dorsal skin using methods recommended by McAllister et al.
 (2004) [@mcallister2004].
-Crews collected a GPS point (estimated precision error \~3 m) for each capture using a handheld Garmin GPS unit.
-Crews weighed each adult frog inside a tared plastic bag to the nearest 0.1 g using a Pesola spring scale.
-For each captured adult individual, crews only collected length and weight data during the first capture event within a three-day primary period.
+Next, we recorded the frog's sex and weighed it inside a tared plastic bag to the nearest 0.1 g using a Pesola spring scale.
+Finally, we collected a GPS point (estimated precision error \~3 m) for each capture using a handheld Garmin GPS unit (Garmin citation).
+We only collected length and weight data during the first capture event within a three-day primary period.
 Afterward, if crews captured the same individual on a subsequent day during the same primary period, crews only recorded PIT tag, sex, and location data.
@@
 
-We estimated survival probability, detection probability, recruitment of new frogs into the adult population, and adult population size for the Mossy Pond *R.
-sierrae* population using a site-specific Jolly-Seber CMR model implemented in a Bayesian framework with R statistical software \[[\@R](v4.4.2)\].
-Our model was adapted from the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), which was developed for the analysis of robust design CMR data collected from *R.
+We conducted our statistical analyses using a two-stage approach that leverages Bayesian uncertainty quantification.
+In the first stage, we estimated survival probability, detection probability, recruitment, and adult population size for the Mossy Pond *R.
+sierrae* population using a site-specific open population Jolly-Seber CMR model implemented in a Bayesian framework with R statistical software \[[\@R](v4.4.2)\].
+Our model was adapted from the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), developed expressley to analyze robust design CMR data from *R.
 sierrae* populations.
 The model represents a superpopulation of M individuals whose states (not recruited, alive, dead) and observations (detected, not detected) evolve through time, with any new frogs assumed to arise from local reproduction [@knapp2024].
-We assessed model convergence via trace plots and Gelman–Rubin statistics (rhat).
+We assessed model convergence via trace plots and Gelman–Rubin statistics (rhat) (find a citation) and compared alternative parameterizations using leave-one-out (LOO) cross-validation (citation).
 
-**Model selection and model comparison with LOO cross-validation.**
+In the second stage of our statistical analysis, we leveraged the posterior distributions from our Bayesian CMR model to investigate environmental drivers of temporal variation in population parameters.
+For each of the resulting 4000 posterior samples, we calculated correlation coefficients between winter severity and demographic parameters (recruitment, abundance, and survival), yielding posterior distributions of these correlations.
+This approach allowed us to quantify both the strength of relationships between environmental factors and population dynamics and the uncertainty in these relationships, while maintaining the uncertainty structure estimated in the first stage.
+We ran multiple correlation analyses on these variables, lagging winter severity by 0, 1, 2, and 3 years, as well as a 3-year rolling average, to account for possible effects of the species' life history strategy, where individuals will overwinter as tadpoles 1-4 times before morphing to frogs.
 
-To extend our analysis and investigate the environmental drivers of variation in our estimated population parameters, we employed a two-stage Bayesian framework.
-First, we obtained full posterior distributions of the values of key population parameters—recruitment, abundance, and survival—from our mark-recapture model (described above).
-In the second stage, we propagated this uncertainty into subsequent analyses by performing regression and correlation analyses on the full set of posterior draws.
-In one approach, we calculated the Pearson correlation coefficient between winter severity (lagged by one year) and recruitment for each draw, resulting in a posterior distribution of correlation coefficients that summarizes the strength and direction of the association.
-In another approach, we performed separate linear regressions on each draw to derive a distribution of regression coefficients.
-Both methods allowed us to integrate the uncertainty from the capture–recapture model into our inference on the effects of environmental variables, ensuring that our conclusions about the relationship between winter severity and recruitment reflect the full uncertainty of the underlying estimates.
+*In one approach, we calculated the Pearson correlation coefficient between winter severity and recruitment for each draw, resulting in a posterior distribution of correlation coefficients that summarizes the strength and direction of the association.*
 
-Our model tracks survival at the individual level, and we computed a population-wide probability of surviving the interval between primary periods by summing those individual probabilities and dividing by the total number of frogs in the superpopulation (*M*).
+*In another approach, we performed separate linear regressions on each draw to derive a distribution of regression coefficients.
+Both methods allowed us to integrate the uncertainty from the capture–recapture model into our inference on the effects of environmental variables, ensuring that our conclusions about the relationship between winter severity and recruitment reflect the full uncertainty of the underlying estimates.*
 
-To complete analyses in an annual framework (i.e., investigate the relationship between winter severity and recruitment), we needed to aggregate each aprameter to an annual value.
-....
-For recruitment, posterior draws were extracted from the model output and then aggregated by summing estimates across primary periods to yield annual values.
-For abundance,
+### Winter severity calcualtion
 
-### Winter severity calculations
+We defined winter severity as the percent of historical average snow-water equivalent (SWE) remaining on 1 April of each year (100% indicating average severity).
+This relative metric, rather than absolute SWE depth, allowed us to incorporate data from ten snow observation stations surrounding our study area, thereby addressing the high spatial variability of snowpack distributions documented in mountainous regions (Revuelto et al., 2014).
+Snowpack distribution is strongly influenced by topographic position indices and variables such as elevation, aspect, and slope, which can create significant heterogeneity even within small catchments.
+By using multiple observation points (located between 5.1 and 19.5 km from the study area), we minimized site-specific biases that could result from topographic factors affecting individual stations.
 
-We defined winter severity as the percent of historical average snowpack—measured as snow-water equivalent (SWE) in inches—remaining on 1 April of a given year (e.g., 100% of historical average snow pack indicates an average-severity winter).
-We calculated the severity of each winter between 2011 and 2018 at 10 snow observation stations surrounding our study area, then used ArcGIS Pro (citation) to interpolate those values across the landscape and calculate the average pixel value in our study area (figure).
-Snow observation stations were between 5.1 and 19.5 km from the Mossy Pond study area, and the diversity of local conditions captured by these 10 stations insulated our measurement from site-specific factors that may have produced a non-representative measurement in any given year.
-All snow pack data was provided by the California Department of Water Resources, accessed via the California Data Exchange Center (CDEC--[http://cdec.water.ca.gov](http://cdec.water.ca.gov/)).
-
-Each of 10 snow monitoring stations surrounding the MPSA is used as a point estimate of winter severity, and those points are used to interpolate winter severity values across the landscape, providing a robust estimate of winter severity in the MPSA.\
+We calculated winter severity between 2011-2018 for all ten stations, then used ArcGIS Pro (citation) to interpolate these values across the landscape using inverse-distance weighting methods, ensuring stations closer to our study area had proportionally greater influence on the final measurements.
+This spatial interpolation approach produced a continuous surface of winter severity values from which we extracted the average for our study area (Figure X).
+This methodology is appropriate for ecological studies in topographically complex regions where a single measurement station would be unlikely to represent conditions across the entire habitat of a mobile species.
+All snowpack data was obtained from the California Department of Water Resources via the California Data Exchange Center (CDEC—http://cdec.water.ca.gov).
 
@@
 
-**Bd status**
+### Bd status
 
-xx / xx samples tested positive for Bd, with low to moderate loads well below the lethal level described by Vredenburg et al.
-(2010).
-Infection levels in this zone are typical of other enzootic and recovered mountain yellow-legged frog populations (figure xx).
+xx / xx samples tested positive for Bd, with low to moderate infection intensities ranging from xx to xx ITS1 gene copies.
+These infection intensities are well below the levels associated with increased MYL frog mortality; a rough threshold of 580,000 ITS1 gene copies is considered lethal for MYL frogs, and loads reach as high as xxxxxxxxx ITS1 gene copies during epizootic events leading to extirpation (Vredenburg et al.
+2010).
+The infection intensities observed in our study population are typical of other naturally recovering *R.
+sierrae* populations (figure xx), such as those now serving as translocation donors in Yosemite National Park (Knapp et al.
+2024).
+\_\_\_\_\_\_\_ describe enzootic dynamics?
+(Briggs et al.
+2016).
 
-**Abundance**
+### Abundance, Survival, Recruitment, and Detection
+
+*Abundance*
 
@@
 
-**Recruitment**
+*Recruitment*
 
@@
 
-**Survival**
+*Survival*
 
@@
 
-**Detection**
+*Detection*
 
@@
 
-**Winter severity, Recruitment, and Abundance**
+## Winter severity and its effects on population dynamics
 
-Our study period coincided with high variation in winter severity at the Mossy Pond study area, with snowpack SWE ranging from 3.5% of historical average in 2015 to 150% in 2017.
-This natural variability provided an opportunity to explore how winter conditions influence population dynamics.
-Our analyses reveal a strong negative correlation between winter severity (lagged by one year) and the number of frogs recruited into the adult population (median correlation coefficient: -0.92; credible interval: –0.70 to -0.98).
+Our study period coincided with high variation in winter severity at the Mossy Pond study area, with snowpack SWE ranging from 3.5% of historical average in 2015 to 150% in 2017 (Figure x).
+This natural variability provided an opportunity to explore if and how winter conditions influence *R.
+sierrae* population dynamics.
+Our analyses reveal a strong negative correlation between winter severity (lagged by one year) and both the number of frogs recruited into the adult population (median correlation coefficient: -0.92; credible interval: –0.70 to -0.98) and the abundance of the population (median correlation coefficient: -xx, credible interval: -xx to -xx).
 For example, the most severe winter in our study (2016–2017) corresponded to the lowest annual recruitment in 2018.
@@
 
-**Context: The *Rana sierrae / muscosa* story in 3 papers:**
+### A recovering population, back fromt he brink of extinction
 
-1\. [Vredenburg et al. 2010](https://www.pnas.org/doi/10.1073/pnas.0914111107) – *Bd*-induced decline of *R. sierrae*
+Our results describe an *R.
+sierrae* population that is large, relatively stable, and coexisting with Bd in the environment.
+Its population dynamics appear to be driven primarily by yearly variation in weather rather than a devastating disease, just like populations of frog species that are not as susceptible to Bd (bunch of citations for weather driving population dynamics).
+This is a far cry from the raging infection intensities, mass mortality, sharp declines, and extirpations observed across the species' range over the past two decades (Vredenburg et al.
+2010).
+**Need transition here** There is no single documented instances of an *R.
+sierrae* population not crashing soon after its first Bd epizootic event.
+Some of the last holdout Bd-naive R.
+sierrae populations are currently experiencing the familiar disease dynamics.
+This circumstantial evidence suggests overwhelmingly that our study population experienced similar disease dynamics in the not so distant past.
+Further, it's current size, relative stability, and low Bd infection intensities suggest that this population recovered from such a Bd-induced crash just like its counterparts in YNP.
+Low infection intensities in those populations are attributed to resistance to infection, likely evolved through natural selection for more resistant genotypes (Knapp et al.
+2024).
 
-Careful monitoring of *R.
-sierrae* populations through *Bd* epizootic events from 2004-2008 reveals large-scale declines and extirpations.
-69 of 88 observed populations were wiped out by 2008 and the rest are expected to follow.
-Disease is the overwhelming driver of population dynamics in this system.
+The geographic novelty of our study shows that the Bd-persistence phenomenon described in this species extends beyond the southern portion of the species range into places where the species looks different.
+**Does the geographic distribution of resistance suggest anything about the mechanisms?
+Resistance is unlikely to to be predominantly environmental, as the habitat here is, by some measures, more hospitable to Bd growth (lower elevation, warmer temps...).
+Address compensatory recruitment?
+Address immune responses?
+Use the 2024 paper to talk about evolutionary rescue.**
 
-2.  [Knapp at al. 2016](https://www.pnas.org/doi/full/10.1073/pnas.1600983113) – Persistence and natural recovery
+### Implications for conservation management
 
-After decades of decline, *R.
-sierrae* in Yosemite National Park are increasing in abundance.
-The trend is observed at the landscape scale, across hundreds of populations, providing a rare example of amphibian resilience.
-A laboratory experiment reveals that resistance to *Bd* (reduced infection susceptibility) may be the mechanism enabling persistence and recovery.
+### Climate and population viability
 
-3.  [Knapp et al 2024](https://www.nature.com/articles/s41467-024-53608-4) – Translocations & reintroductions: harnessing natural *Bd-*resistance for human-assisted recovery
+-   Intro to climate becoming a more important driver of population dynamics (Luedetke et al. 2023)
 
-Individuals from *Bd*-persistent populations exhibit resistance to infection, consistent with expectations under an “evolutionary rescue” scenario.
-A 15-year study shows that reintroducing resistant individuals from naturally recovering populations to unoccupied habitat is an effective method for *R.
-sierrae* recovery.
- A majority of new populations exhibit successful reproduction and recruitment (becoming self-sustaining) and models show low 50-year extinction probabilities.
+-   In the Sierra Nevada, the mountain range to which *R. sierrae* is endemic, past climate trends have been \_\_\_\_ and future trends are forecasted to be \_\_\_\_\_.
 
-**How does my work fit into this story?**
-
-1.  ***Bd*****-persistent populations: what’s going on?**
-
--   To better understand the *R.
-sierrae* story (and the greater global story of amphibian-Bd dynamics) we need detailed information about how populations are persisting and recovering in the presence of *Bd*.
-
--   Detailed demographic data from large persistent populations are rare in the literature.
-We need fine-scale information about population structure and dynamics not included in the landscape-level studies outlined above.
-
--   This 5-year capture-mark-recapture study of the Mossy Pond *R.
-sierrae* population (Tahoe National Forest, CA) provides an excellent case study of a large persistent population in the northern extent of the species’ range.
-
--   Relatively little is known about northern populations compared to their southern counterparts, adding another layer of usefulness to this study (expanded geographic representation).
-
--   How do my results help to fill this information gap?
-
-2.  **Demonstrating a return to ‘normal’ ecological regulation**
-
--   The most compelling finding of this study is the strong negative correlation between the number of individuals recruited into the adult population annually and winter severity (with a 1-year lag).
-The relationship holds for abundance and winter severity as well.
-
--   This shows a shift from the disease-driven dynamics described by Vredenburg et al. 2010 to climate-driven dynamics, which is typical of ‘normal’ amphibian populations where *Bd* is not a factor.
-
--   A paragraph on observed historical trends & future projections of winter severity in the Sierra Nevada. What might the future hold for these frogs under changing climatic conditions?
-
--   Conservation applications
-
--   Knapp et al.
-2024 show that large persistent populations are key to the cutting edge of frog conservation.
-They provide *bd-*resistant individuals for reintroductions, which can establish self-sustaining populations that exhibit the founders’ *Bd* resistance.
-
--   Utilizing persistent populations as conservation resources requires decision making informed by detailed population studies.
-
-    -   When is the best time to remove individuals?
-
-    -   How many individuals can we remove?
-
--   Specifically, the Mossy Pond population is a prime candidate for donation to the nearby 5-Lakes Basin, where CDFW recently finished a multi-year habitat restoration project.
-These results will help inform that project.
+-   In light of our results, what does this mean?
 

### 2025-04-24 code/R/Survival_by_sex.Rmd
Initial commit of M-F survival probability

--- 2025-01-31 code/R/Survival_Probability_Difference_MF.Rmd
@@
 # laod the model with sex as a covariate of survival
-mod <- readRDS("/Users/johnimperato/Desktop/Mossy-Pond_companion/sex_survival_covariate_model.rds")
+mod_sex <- readRDS("/Users/johnimperato/Desktop/Mossy-Pond_companion/sex_survival_covariate_model_4.rds")
 
@@
 
-# Convert from factor to numeric via character
-time_summaries_df <- time_summaries_df %>%
-  mutate(time_index = as.numeric(as.character(time_index)))
-
 # Combine the summaries into a single data frame
@@
 
+# Convert from factor to numeric via character
+time_summaries_df <- time_summaries_df %>%
+  mutate(time_index = as.numeric(as.character(time_index)))
+
 # View the summaries# Viewpaste0() the summaries
 print(time_summaries_df)
+
+View(time_summaries_df)
 
@@
   labs(
-    title = "Difference in Survival Probabilities (M - F) Between Primary Periods",
+    title = "Difference in Survival Probabilities (M - F) Between Primary Periods\nsurvival_fill_value = 'M'",
     x = "Primary Period",
@@
 # Save the plot to a file
-ggsave("/Users/johnimperato/Desktop/Mossy-Pond_companion/difference_in_survival_probabilities.png", 
-       plot = difference_plot, width = 10, height = 6, dpi = 300)
+ggsave("/Users/johnimperato/Desktop/Mossy-Pond_companion/difference_in_survival_probabilities_M_fill.png", 
+       plot = difference_plot, width = 10, height = 6, dpi = 300, bg = "white")
 ```

### 2025-04-24 code/R/sex_covariate_DataFormatting.Rmd
Initial commit of sex covariate daata formatting

---
title: "Data Formatting for Sex Covariate"
output: html_document
date: "2025-01-21"
---

```{r}

library(tidyverse)
library(rstan)
library(cmdstanr)
library(data.table)
library(mrmrmr)

captures <- read.csv("/Users/johnimperato/Documents/private_MP_data/captures_MP.csv")

full_captures <- read.csv("/Users/johnimperato/Desktop/Mossy-Pond_companion/Capture_Data.csv")

surveys <- read.csv("/Users/johnimperato/Desktop/Mossy-Pond/data/survey_MossyPond.csv")

```

## Data Formatting

### Clean up frog sex data

Resolve inconsistencies in frog_sex column. For frogs with multiple sexes listed over multiple captures, use the most recent observation. If the most recent observation is blank, use the most common sex identification.

```{r}

# Load the full capture data set
capture_data_full <- read.csv("/Users/johnimperato/Desktop/Mossy-Pond_companion/Capture_Data.csv")

# Select relevant columns, remove duplicates, and arrange by PIT tag
capture_data_full <- capture_data_full %>% 
  select(Date, PIT_TagCode, frog_sex) %>% 
  unique() %>% 
  arrange(PIT_TagCode)

# Clean up the Date column (remove time and convert to Date format)
capture_data_full$Date <- gsub("00:00:00 EDT", "", capture_data_full$Date)
capture_data_full$Date <- as.Date(capture_data_full$Date, format = "%a %b %d %Y")

# Write function to resolve the frog)_sex, taking the most recent, or the most common non-empty sex identification if the most recent is empty
resolve_sex <- function(data) {
  # Check if there's any non-empty sex data
  non_empty_data <- data %>% filter(frog_sex != "")

  if (nrow(non_empty_data) == 0) {
    return(NA)  # Return NA if no valid sex is found
  }

  # Find the most recent sex
  most_recent <- data %>% filter(Date == max(Date)) %>% pull(frog_sex)

  if (most_recent != "") {
    return(most_recent)
  } else {
    # If most recent sex is empty, return the most common non-empty sex
    return(non_empty_data %>% 
             count(frog_sex) %>% 
             slice_max(n, n = 1) %>% 
             pull(frog_sex))
  }
}

# Apply the function to resolve the sex for each frog
resolved_sex_data <- capture_data_full %>%
  group_by(PIT_TagCode) %>%
  group_modify(~ mutate(.x, Sex_resolved = resolve_sex(.x))) %>%
  ungroup() %>% 
  select(PIT_TagCode, Sex_resolved) %>% 
  unique() %>% 
  rename(sex = Sex_resolved)%>%
  mutate(sex_binary = ifelse(sex == "M", 1, ifelse(sex == "F", 0, NA))) %>% 
  filter(!is.na(sex) & sex != "U") %>% 
  rename(pit_tag_id = PIT_TagCode)

# compoute observed sex ratio
sex_ratio <- resolved_sex_data %>% 
  count(sex) %>% 
  mutate(prop = n / sum(n))

```

### Join sex data with capture data

```{r}

captures <- captures %>% 
  left_join(resolved_sex_data,
            by = "pit_tag_id")

write.csv(captures, "/Users/johnimperato/Desktop/Mossy-Pond_companion/MP_caps_sex.csv")

```

### Run updated clean_data function

1 = female, 2 = male

Fill in survival sex values according to observed sex ratio

```{r}

data <- mrmr::clean_data(captures = captures,
                              surveys = surveys,
                              survival_formula = ~ sex,
                              survival_fill_value = c(sex = "MISSING"))

# index individuals with NA sex value
missing_idx <- which(is.na(data$survival_covariate_df$sex) |
                     data$survival_covariate_df$sex == "MISSING" |
                     data$survival_covariate_df$sex == "")
# replace NA values with a sample of sex assignments with probability derived from the observed data 
if(length(missing_idx) > 0) {
  data$survival_covariate_df$sex[missing_idx] <- sample(
    sex_ratio$sex,
    size = length(missing_idx),
    replace = TRUE,
    prob = sex_ratio$prop
  )
}

```

```{r}

mod_sex <- mrmrmr::fit_model_TVD(
  data = data,
  chains = 3,
  iter_warmup = 1000,
  iter_sampling = 1000,
  parallel_chains = 4,
  threads_per_chain = 2
)

modsex_summary <- mod_sex$m_fit$summary()
View(modsex_summary)

saveRDS(mod_sex, "/Users/johnimperato/Desktop/Mossy-Pond_companion/sex_survival_covariate_model_4.rds")

```

### 2025-04-29 doc/Manuscript.qmd
first draft for circulation

--- 2025-02-01 doc/Manuscript.qmd
@@
 ---
-format: pdf
+format:
+  pdf:
+    documentclass: article
+    linestretch: 2
+    number-sections: true
+    include-in-header: draft-options.tex
 bibliography: references.bib
+execute: 
+  echo: true
 ---
 
-# Title Title Title
+# Dynamics of an Endangered Frog Population Recovering from Bd-Driven Decline
 
@@
 
-^1^School of the Environment, Yale University, New Haven, CT, 06511
+^1^Yale School of the Environment
 
@@
 
-The well-documented decline of the endangered Sierra Nevada yellow-legegd frog in its namesake mountain range (bunch of citations) is emblematic of the crisis facing amphibians globally (citations).
-As is the case with numerous other species...the fungal pathogen Bd is a primary driver of SNYLF decline in the Sierra Nevada (citations), and the effects of climate change loom large-yet remain understudied- especially in an alpine environment expriencing accelerated climate change (mountain ranges are warming faster than other places?) (citation).
+Amphibians continue to experience global declines, with infectious diseases recognized as primary drivers of biodiversity loss [@luedtke2023a] (Luedtke et al.
+2023).
+Chytridiomycosis, caused by the fungal pathogen *Batrachochytrium dendrobatidis* (Bd), has driven declines or extinctions of approximately 200 amphibian species globally [@skerratt2007] (Skerratt et al.
+2007).
+Amphibian responses to Bd infection vary dramatically; some populations show minimal impact, others rapidly go extinct, and a rare few exhibit persistent recovery following severe declines [@lips2016] (Lips 2016).
+The mechanisms supporting persistence or recovery remain poorly understood but likely involve interactions among ecological, evolutionary, immunological, and demographic factors [@scheele2017] (Scheele et al.
+2017).
 
--   get rid of climate *change* here. Introduce just climate somewhere in the intro, then get into climate *change* later, like int he discussion.
+The mountain yellow-legged (MYL) frog species complex (*Rana sierrae* and *R.
+muscosa*), endemic to California's Sierra Nevada and Transverse mountain ranges, provides a particularly well-documented case of amphibian decline and recovery from Bd infection.
+Starting in the 1970s, Bd spread through the Sierra Nevada, causing widespread and rapid extirpations of MYL frog populations [@vredenburg2010] [@vredenburg2019] (Vredenburg et al.
+2010, 2019).
+By the early 2000s, over 90% of their historical range had been lost, due to combined impacts from Bd and habitat degradation caused by introduced trout [@knapp2000] (Knapp and Matthews 2000).
+Despite these losses, a few populations survived initial Bd outbreaks, eventually exhibiting increased abundance and apparent disease resistance [\@knapp2016](Knapp%20et%20al.%202016).
+These dynamics exemplify patterns of evolutionary rescue [@carlson2014] (Carlson et al.
+2014), in which populations rapidly acquire adaptive genetic variation that restores demographic viability under persistent threats [@knapp2024] (Knapp et al.
+2024).
 
-Though SNYLF decline fits the gloomy narrative, the examples of resilience and recovery that this species provides do not.
-Bd will typically wipe out (extirpate?) SNYLF populations (Vredenburg?), but a rare subset of individuals exhibit resistance to the pathogen ('resistant individuals') and contribute to the natural recovery of populations that persist in the presence of Bd (Knapp).
-These disease-persistent populations have become instrumental to conservation efforts for this species in their provision of resistant individuals as founders to establish new, resilient populations in previously-occupied habitat (Knapp 2024).
-We have the chance to pull off a rare feat: landscape scale recovery of an endangered species.
+While broad-scale studies have described these declines and recoveries, detailed demographic analyses at the individual population level remain scarce, primarily due to logistical challenges.
+Most fine-scale demographic research has focused on southern populations, particularly within Yosemite National Park [@fellers2013] [@joseph2018] [\@knapp2024](Fellers%20et%20al.%202013;%20Joseph%20and%20Knapp%202018;%20Knapp%20et%20al.%202024).
+In contrast, the population dynamics of frogs in the northern Sierra Nevada—where habitats are more heterogeneous and populations are smaller and lower in elevation—remain understudied [@brown2019] (MYLF ITT 2018; Brown et al.
+2020).
 
-To successfully leverage these disease-persistent populations as conservation resources, we must gain a detailed understanding of their demographics, dynamics, and the factors that influence their viability.
-Without good information, we can't make smart decisions and we will waste limited resources (Brown xxxx).
+To address this gap, we conducted a five-year capture-mark-recapture (CMR) study of a naturally recovering, Bd-persistent *R.
+sierrae* population in Tahoe National Forest.
+Our objectives were to estimate key demographic parameters—abundance, survival, recruitment, and detection probability—to provide insights into the ecological processes and factors that contribute to population viability under endemic Bd infection.
+We also examined the role of climate variability in shaping these demographic parameters, as climate change is increasingly recognized as a primary driver of amphibian declines worldwide, comparable in impact to disease and habitat loss [\@luedtke2023a](Luedtke%20et%20al.%202023).
+Understanding how climate influences abundance, survival, and recruitment is particularly important for high-elevation amphibian populations living with endemic Bd infection.
 
-The population we focus on in this study-Mossy Pond-provides an excellent opportunity to fill the void of information around disease-persistent populations of this species in the northern portion of its range (cite interagency 2018 and Brown).
-Pop description: Weird habitat, funky frogs, and little knowledge, etc..
-Complexity of habitat here is representative of northern pops generally: streams, complexity, etc.
-
-In this investigation we set out to estimate demographic parameters (getting absolute abundance is really important, as Os for citations on that) and to investigate the (factors that influence population viability in the Bd-persistent state) (role that climate plays in population viability).
-Ultimately, we hope our insights are useful for the management of this paprticualr population, for a deeper understanding of the species as a whole, and as a hint of how climate change may affect alpine amphibian populations in the future.
-
---------------------------------
-
-By acting as donor populations for the human-facilitated reintroduction of disease-resistant animals into extirpated habitat, these unique Bd persistent populations have become instrumental to conservation efforts.
-
-Some populations of MYLF exhibit resistence to Bd and will recover naturally (citations).
-These naturally-recovering populations (persistent populations) have become instrumental to human-facilitated conservation efforts, and present an opportunity for landscape scale recovery of an endangered species (Knapp 2024).
-
-What effects may climate have on these recovery efforts?
-
--------------------------------------------
-
-Amphibians are the most threatened class of vertebrates.
-Disease, habitat loss, and, increasingly, climate change drive their global decline [@luedtke2023].
-The endangered mountain yellow-legged frog species complex (*Rana muscosa* and *R.
-sierrae*) is emblematic of such declines.
-**Citation**.
-These frogs are endemic to the Sierra Nevada and Transverse mountain ranges of California and Nevada, and are characterized by \_\_\_\_\_\_.
-Non-native fish introductions across their range have severely reduced suitable habitat for the species[@knapp2000], and the fungal pathogen *Batrachochytrium dendrobatadis* (Bd) causes mass mortality and rapid population declines [@vredenburg2010].
-As a result, more than 90% of *R.
-muscosa* and *R.
-sierrae* populations have been extirpated from their historical range [@vredenburg2007].
-
-Despite these substantial losses, mountain yellow-legged frogs provide rare examples of amphibian resilience in the face of multiple stressors.
-Following the removal of non-native fish from alpine lakes, *R.
-muscosa* population density can increase significantly [@vredenburg2004][@knapp2007].
-Additionally, some mountain yellow-legged frog populations exhibit resistance to Bd infection and naturally recover despite the pathogen's continued presence in the environment [@knapp2016].
-These Bd-persistent populations are increasingly important to recovery efforts for the species.
-A recent study shows that reintroducing frogs from persistent populations into vacant habitat can establish self-sustaining populations with low long-term extinction probability [@knapp2024].
-It is critical, therefore, to study the dynamics of these unique populations to gain insights into how they are persisting in the presence of Bd and to assess their value to future conservation efforts.
-
-Among Bd-persistent mountain yellow-legged frog populations, little is known about the ecology and demographics of those in the northern part of their range [@interagency2018][@brown2019].
-Compared to their relatively well-studied counterparts in the central and southern Sierra Nevada, northern populations are fewer in number, have lower abundances, utilize a more diverse range of habitat, and occupy lower elevations [@interagency2018].
-To implement effective conservation measures for mountain yellow-legged frogs in the northern portion of their range, it is vital to better understand the demographics and dynamics of these rare and understudied disease-persistent populations [@brown2020].
-
-In this study, we describe one such population, located in the Mossy Pond area of the Tahoe National Forest, California...
+Understanding demographic patterns in recovering MYL frog populations has significant conservation implications.
+Recent research demonstrates these populations can serve as effective founders in reintroduction programs, establishing self-sustaining populations when translocated to suitable but currently unoccupied habitats [@knapp2024] (Knapp et al.
+2024).
+Leveraging evolved disease resistance from naturally recovering populations offers a promising conservation strategy.
+Our study provides critical demographic data to support these conservation efforts, particularly for the less-studied northern Sierra Nevada populations.
 
@@
 
-The Mossy Pond Complex is located in the Tahoe National Forest, north of highway 80 in Nevada County, California.
-It consists of approximately 60 lakes, ponds, and streams set on granite benches southeast of Fordyce Lake.
-Elevation ranges from 6,400 feet to 7,100 feet.
-The study area is a one-square-mile section of the complex, containing the namesake Mossy Pond, its seasonally-flowing outlet stream, and 12 ephemeral ponds.
-**\*\*Rationale for study area selection\*\*.** The United States Geological Survey (USGS) first detected SNYLF in the complex in 1998, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.
+The Mossy Pond study area is a one square mile section of the Tahoe National Forest, north of Highway 80 in Nevada County, California.
+It is characterized by lakes, ponds, and streams set on granite benches, ranging from 6,400 feet to 7,100 feet in elevation.
+The study area includes the namesake Mossy Pond (6 hectare surface area, maximum depth 2.5 meters), its seasonally flowing outlet stream, and 12 ephemeral ponds.
+The absence of deep-water habitat is unusual for areas supporting MYL frog populations, given their reliance on water bodies deeper than 4 meters for overwinter survival [@bradford1983] [@knapp2003] (Bradford 1983; Knapp et al.
+2003).
 
-### Capture-mark-recapture surveys
+*R.
+sierrae* is a federally endangered, mid-sized ranid frog that occupies montane waterbodies in California's Sierra Nevada (U.S.
+Fish and Wildlife Service 2014).
+The US Geological Survey first detected *R.
+sierrae* in the Mossy Pond area in 1998, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.
+**More?**
 
-Capture-mark-recapture (CMR) sampling was conducted according to Pollock's [@pollock1982] robust design.
-Under the robust design, three primary survey periods occurred each summer from 2014 to 2018, each consisting of three consecutive survey days (secondary periods).
-**More needed?**
+### Bd status
 
-Field crews captured frogs by hand or dip net, and processed frogs at the point of capture.
-Crews first scanned each captured frog with a BioMark 601 passive integrated transponder (PIT) tag reader to detect if the frog was marked (i.e., a recapture).
-Crews then used calipers to measure the snout to urostyle (SUL) length of each captured frog, and released frogs \<40 mm SUL (which are considered subadults) without further processing.
-Crews continued collecting data on larger (\>40 mm SUL) frogs, which are considered adults.
-For new adult captures, crews inserted an 8 X 1.4 mm PIT tag under the dorsal skin using methods recommended by McAllister et al.
-(2004) [@mcallister2004].
-Crews collected a GPS point (estimated precision error \~3 m) for each capture using a handheld Garmin GPS unit.
-Crews weighed each adult frog inside a tared plastic bag to the nearest 0.1 g using a Pesola spring scale.
-For each captured adult individual, crews only collected length and weight data during the first capture event within a three-day primary period.
+To assess the Mossy Pond *R.
+sierrae* population's Bd status, we collected epithelial swabs from 24 adult frogs over three years (2008, 2010, and 2021).
+Partner scientists quantified the amount of Bd DNA on each swab using real-time quantitative polymerase chain reaction (qPCR) analysis (Boyle et al.
+2004).
+We report infection loads as the number of ITS1 (Internal Transcribed Spacer 1) gene copies detected [@longo2013] [@joseph2018] (Longo et al.
+2013; Joseph and Knapp 2018).
+Quantifying Bd loads rather than simply testing for Bd presence allowed us to compare our study population with other known recovering populations.
+Partner scientists provided *R.
+sierrae* Bd load data from three known naturally recovering populations in Yosemite National Park [@knapp2024] (Knapp et al.
+2024).
+
+### Frog population surveys
+
+We studied *R.
+sierrae* with capture-mark-recapture (CMR) methods, which are widely used for estimating wildlife population parameters such as abundance, survival, and recruitment (Williams et al.
+2001).
+We conducted CMR sampling according to Pollock's robust design [@pollock1982], which involves sampling at two temporal scales: primary periods (between which the population is open to demographic change) and secondary periods, during which the population is assumed closed.
+Each summer from 2014 to 2018, we visited the study site 3 or 4 times (primary periods) for three consecutive days of surveys (secondary periods), for a total of 47 individual surveys.
+
+Surveys occurred between 0800 and 1800, when *R.
+sierrae* are most active.
+We systematically searched all water bodies in the study area, capturing frogs by hand or dip net for immediate processing.
+We first scanned each captured frog with a BioMark 601 passive integrated transponder (PIT) tag reader (BioMark, Boise, Idaho, USA) to detect if the frog was marked (i.e., a recapture).
+We then used calipers to measure the snout to urostyle (SUL) length of each captured frog, and released frogs \<40 mm SUL (which are considered subadults) without further processing.
+Frogs larger than 40mm (adults) required further processing.
+New adult captures were PIT-tagged dorsally following McAlister et al.
+2004 [@mcallister2004].
+Next, we recorded the frog's sex and weighed it inside a tared plastic bag to the nearest 0.1 g using a Pesola spring scale.
+Finally, we collected a GPS point (estimated precision error \~3 m) for each capture using a handheld GPS unit.
+We only collected length and weight data during the first capture event within a three-day primary period.
 Afterward, if crews captured the same individual on a subsequent day during the same primary period, crews only recorded PIT tag, sex, and location data.
 
-### Population modeling
+Over the 5-year period of our study, we recorded 3,631 captures of 657 individual frogs.
 
-We estimated survival probability, detection probability, recruitment of new frogs into the adult population, and adult population size for the Mossy Pond *R.
-sierrae* population using a site-specific Jolly-Seber CMR model implemented in a Bayesian framework with R statistical software [\@R](v4.4.2).
-Our model was adapted from the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), which was developed for the analysis of robust design CMR data collected from *R.
-sierrae* populations.
-The model represents a superpopulation of M individuals whose states (not recruited, alive, dead) and observations (detected, not detected) evolve through time, with any new frogs assumed to arise from local reproduction [@knapp2024].
-We assessed model convergence via trace plots and Gelman–Rubin statistics (rhat).
+### Statistical Analysis
 
-Models copmpared with LOO cross-validation.
+We estimated survival, detection, recruitment, and adult population size for the Mossy Pond *R.
+sierrae* population using a site‐specific open‐population Jolly–Seber CMR model in a Bayesian framework with R statistical software \[[\@R](v4.4.2)\].
+Our implementation builds on the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), which tracks a superpopulation of M individuals through not‐recruited, alive, or dead states and models detection and demographic processes across primary and secondary periods [@knapp2024].
+We assessed convergence with trace‐plots and Gelman–Rubin (R̂) diagnostics and compared model alternatives using leave‐one‐out cross‐validation.
 
-### Winter severity calculations
+In our model φₜ represents the probability that an individual alive at primary period t survives through to t + 1.
+Because the interval length Δₜ (in months) varies, we standardized each φₜ to a per‐month rate by taking the Δₜth root of every posterior draw, producing monthly survival estimates (with 95% credible bounds) that are directly comparable across short summer and long overwinter intervals.
 
-\
+To test for sex differences in survival, we added a fixed “male” coefficient to the logit‐scale survival predictor alongside the baseline intercept and period‐specific random effects.
+Pseudo-individuals were assigned sex by sampling from the observed sex ratio, assuming equal detectability.
+For each draw and period we computed the female logit (intercept + period effect), added the male coefficient to get the male logit, transformed both via the logistic function, and recorded their difference.
+We summarize those interval‐specific male–female differences by posterior mean, standard deviation, and 95 % credible interval.
+
+To capture unmeasured heterogeneity in detection across survey occasions, we introduced survey‐specific random effects on logit‐scale detection.
+This parallels the hierarchical structure used for survival and recruitment in the original mrmr model and acknowledges that factors like temperature, effort, observer, and water clarity can affect detectability even when fixed covariate data are incomplete.
+
+As a final analytical step, we used the posterior samples from our fitted CMR model to test for climate–demography links.
+For each posterior draw, we calculated Pearson correlation coefficients between winter severity (percent of average snow-water equivalent remaining on 1 April) and each demographic parameter—monthly survival, recruitment, and adult abundance—thereby generating full posterior distributions of those correlations.
+To capture potential delayed effects given the frog’s multi-year tadpole overwintering, we repeated these analyses using winter severity at lag 0, 1, 2, and 3 years as well as a 3-year rolling average.
+This approach quantifies both the strength and uncertainty of climate influences on population dynamics while preserving the hierarchical uncertainty structure of our Bayesian framework.
+
+### Environmental Covariates
+
+We defined winter severity as the percentage of historical average snow-water equivalent (SWE) remaining on April 1 each year (100% indicating average severity).
+We obtained SWE data from ten regional snow-survey stations (4.5–19.5 km from Mossy Pond) to capture the spatial variability in snowpack present in topographically complex mountainous environments [@revuelto2014] (Revuelto et al.
+2014).
+Annual SWE values (2011–2018) were interpolated across the study area using inverse-distance weighting in ArcGIS Pro 3.4.3 (Esri Inc.
+2025).
+From the resulting continuous SWE raster, we extracted the mean percent SWE within our one-square-mile study boundary.
+All raw SWE measurements were obtained from the California Data Exchange Center (CDEC; http://cdec.water.ca.gov).
 
@@
 
+### Bd status
+
+16 out of 24 samples tested positive for Bd, with infection intensities ranging from 16 to 43,318 ITS1 gene copies.
+These Bd loads are well below the \~600,000 gene copy threshold associated with severe chytridiomycosis and increased MYL frog mortality [@vredenburg2010] [@longo2013] (Vredenburg et al.
+2010; Longo et al.
+2013).
+The infection intensities observed in our study population are typical of those measured in other known naturally recovering *R.
+sierrae* populations [@knapp2024] (Figure 1; Knapp et al.
+2024).
+
+![**Figure 1.
+Batrachochytrium dendrobatidis (Bd) infection loads across naturally recovering mountain yellow-legged frog populations.** Box plots show the distribution of Bd infection intensity (ITS1 gene copies) for Mossy Pond and three recovering Rana sierrae populations in Yosemite National Park (identified by ID numbers 70459, 72996, and 70567).
+The dashed red line indicates the approximate threshold (\~600,000 gene copies) associated with severe chytridiomycosis and increased mortality in mountain yellow-legged frogs.
+Note the logarithmic scale on the y-axis.
+All recovering populations maintain Bd loads well below the mortality threshold, with Mossy Pond showing infection intensities comparable to those observed in the Yosemite populations.](images/Bd_load_comparison-01.png)
+
+### Abundance, Survival, Recruitment, and Detection
+
+*Abundance*
+
+We calculated the abundance of the Mossy Pond *R.
+sierrae* population for each of the 16 primary survey periods of the study.
+The lowest abundance was observed in September 2018 (117 individuals, 95% credible interval (95% CrI): 103-132) and a maximum of 245 individuals (95% CrI: 237-256) was observed during the fifth primary period, in July 2015.
+Across all primary periods, the mean abundance of the population was 168 individuals.
+
+*Recruitment*
+
+A total of 629 frogs recruited into the adult population over the course of the study.
+Annual recruitment ranged from 60 individuals (95% CrI: 45-78) in 2018 to 166 individuals (95% CrI: 143-188) in 2015.
+
+*Survival*
+
+Monthly survival during the active (summer) season was markedly lower and more variable than during overwintering.
+Across 11 summer intervals, mean monthly survival was 0.803 (range: 0.673–0.904), whereas across 4 overwinter intervals, mean monthly survival was 0.966 (range: 0.945–0.974).
+
+In every interval, mean male survival exceeded mean female survival, with differences ranging from 0.015 (95% CrI 0.004-0.029) between the first and second primary periods to 0.069 between the sixth and seventh primary periods (Figure 2).
+The 95% credible interval for each period lay entirely above zero, indicating consistently higher male survival across both summer and overwinter intervals.
+These results suggest a modest but statistically robust male survival advantage throughout the study.
+
+![**Figure 2.
+Sex-specific survival differences in the Mossy Pond *Rana sierrae* population, 2014-2018.** Points represent the mean difference in monthly survival probability between males and females (M-F) across 15 primary period transitions.
+Vertical lines show 95% credible intervals.
+All intervals are entirely above zero, indicating consistently higher male survival across both summer and overwinter intervals throughout the study period.
+Overwinter periods occur during the 3-4, 6-7, 9-10, and 12-13 interval transitions.](images/survival_by_sex.png)
+
+*Detection*
+
+For each survey (secondary period; n=47) of the study, we calculated the probability that an individual frog that is alive and available for detection is actually detected.
+Detection probability was lowest (median 0.26, 95% CrI: 0.2-0.31) during the sixteenth survey (09 September, 2015) and highest (median 0.61, 95% CrI: 0.54-0.67) during the forty-third survey (07 August, 2018).
+The average detection probability over 47 individual surveys was 0.45.
+
+### Winter severity effects on population dynamics
+
+Our study period coincided with high variation in winter severity at the Mossy Pond study area, with April 1 SWE ranging from 3.5% of historical average in 2015 to 150% in 2017.
+Our analyses reveal a strong negative correlation between winter severity lagged by one year and both the number of frogs recruited into the adult population (median correlation coefficient: -0.92; 95% CrI: –0.79 to -0.98; Figure 3) and the abundance of the population (median correlation coefficient: -0.84, 95% CrI: -0.8 to -0.89).
+Winter severity showed no notable relationship with overwinter survival probabilities.
+
+![**Figure 3.
+Relationship between adult Rana sierrae recruitment and prior year winter severity at Mossy Pond, 2014-2018.** The solid line represents the number of frogs recruited into the adult population each year (left y-axis), while the dashed line shows winter severity as percent of average snow-water equivalent (SWE) measured on April 1 of the prior year (right y-axis).
+Note the inverse relationship, with high recruitment following mild winters (low SWE) and reduced recruitment following severe winters (high SWE).](images/WinterSeverity_Recruitment-02.png)
+
+![**Figure 4.
+Negative correlation between adult *Rana sierrae* recruitment and prior year winter severity at Mossy Pond.** Scatterplot showing the strong negative relationship (r = -0.92, 95% credible interval: -0.79 to -0.98) between annual adult frog recruitment and the prior year's winter severity (percent of average snow-water equivalent on April 1).
+Each point is labeled with the recruitment year (2014-2018).
+The gray line represents the linear regression of the posterior median number of frogs recruited against winter severity, illustrating the strong negative correlation between these variables.
+Years with mild prior winters (left side of x-axis) yielded substantially higher recruitment compared to years following severe winters (right side of x-axis).](images/WSB_corr.png)
+
+## Discussion
+
+#### **Bd infection dynamics and evidence for resistance**
+
+The Mossy Pond frogs carried Bd at moderate prevalence (67%) but at infection intensities that were two to three orders of magnitude lower than levels associated with increased mortality in MYL frogs.
+Such low loads mirror those reported for other naturally recovering populations in Yosemite National Park and stand in stark contrast to the uniformly high loads (\>600,000 ITS1 copies) and near-certain mortality that typify collapsing populations [@vredenburg2010] (Vredenburg et al.
+2010).
+Taken together with the population’s stable abundance over the study period, these data support the hypothesis that the Mossy Pond frogs have undergone an evolutionary rescue process in which genotypes conferring resistance have been favoured and now dominate the population.
+This expands the geographic scope of Bd-resistant *R.
+sierrae* beyond the southern Sierra Nevada, indicating that resistance is not confined to a single environmental setting or genetic lineage.
+**Too much?**
+
+#### **Winter severity as a driver of recruitment and abundance**
+
+Our Bayesian correlation analysis revealed a strong negative association between winter severity in the previous year and both adult recruitment and population size.
+Mild, low-snow winters may enhance juvenile overwinter survival and accelerate growth such that more animals reach the 40 mm adult threshold before the next breeding season.
+This mechanism has been invoked for long-lived amphibians in other montane systems [@mccaffery2010] (McCaffery & Maxell 2010) and is consistent with the lag-1 pattern we observe.
+The absence of any clear winter-severity signal in overwinter adult survival suggests that once frogs surpass a critical body size, they can tolerate a broad range of cold-season conditions.
+
+#### **Seasonal survival patterns**
+
+Monthly survival was consistently high during the four overwinter intervals (μ = 0.97) but dropped and became more variable in summer (μ = 0.80).
+Elevated summer mortality is probably multifactorial: frogs are active and conspicuous, water levels drop, and thermal and desiccation stress peak—all of which increase predation risk and physiological costs.
+Conversely, winter conditions at Mossy Pond may be unusually benign.
+Although the study area lacks the deep-water refuges (\>4m) traditionally thought essential, our capture records show adults concentrated in the rocky outlet stream and mud-bottomed main pond—habitats that can stay unfrozen and oxygenated thanks to possible groundwater inputs and insulating effects.
+**More from Isaac here?**
+
+The pronounced male survival advantage across all intervals (Δφ = 0.02–0.07) contrasts with Fellers et al.
+(2013), who reported nearly identical annual survival of male and female R.
+sierrae at Summit Meadow, with females sometimes faring slightly better.
+Our pattern instead mirrors Rocky Mountain tailed frogs (Ascaphus montanus), where males have shown modestly higher survival [@honeycutt2019] (Honeycutt et al.
+2019).
+
+Several non-exclusive factors could explain why females at Mossy Pond survived slightly less well.
+Most obviously, the energetic burden of reproduction falls almost entirely on females: producing and depositing large gelatinous egg masses—often while bearing the added weight of one or more males in amplexus—depletes energy reserves and can delay post-breeding foraging.
+Post-oviposition condition declines have been linked to reduced annual survival in other anurans (e.g., common toads; [@madsen2010] Loman & Madsen 2010).
+Additionally, movement studies of *A.
+montanus* show that females travel farther than males to locate suitable oviposition sites, increasing exposure to predators and desiccating microhabitats [@honeycutt2019] (Honeycutt et al.
+2019).
+Together, these reproductive and behavioural costs may magnify female summer mortality in the Mossy Pond *R.
+sierrae* population, driving the male-biased survival pattern we observed.
+
+#### **Climate change context**
+
+Across the western United States, April-1 SWE has declined by about 18% on average since 1955, with particularly steep losses in northern California (US EPA 2024 indicator).
+Regional climate projections point to continued warming, more precipitation falling as rain rather than snow, and further reductions in spring snowpack over the coming decades [@gottlieb2024] (Gottlieb and Mankin 2024).
+In the short term, such trends could benefit Mossy Pond frogs by sustaining the sequence of “light” winters that boost recruitment.
+Over longer horizons, however, chronic snowpack loss and more frequent summer droughts are likely to lower water levels, shorten hydroperiods, and shrink the already shallow overwintering habitats available in the basin.
+The drop in visual-encounter counts recorded after the 2018–2021 drought hints that these longer-term drying forces may eventually override the positive recruitment signal.
+
+### Implications for conservation management
+
+Our results place the Mossy Pond frogs squarely within the “resistant-donor” framework that Knapp et al.
+[@knapp2024] (2024) used to successfully facilitate landscape-scale *R.
+sierrae* recovery in Yosemite National Park.
+In that 15-year experiment, reintroductions sourced from naturally recovering, Bd-resistant *R.
+sierrae* populations established self-sustaining colonies at 75% of recipient sites despite the pathogen’s continued presence, and half of those new populations were projected to remain viable over 50 years.
+The Mossy Pond population’s large size, stable demographics, and demonstrated resistance make it a valuable additional conservation resource, especially for northern Sierra reintroductions where local adaptation may confer an advantage for long-term population viability.
+
 ## References
 
+California Department of Water Resources. 2025. California Data Exchange Center (CDEC). California Department of Water Resources. <https://cdec.water.ca.gov>
+
+Esri Inc. 2025. ArcGIS Pro (Version 3.4.3). Esri Inc. <https://www.esri.com/en-us/arcgis/products/arcgis-pro/overview>
+
+MYLF ITT.
+2018.
+USDA Forest Service, U.S.
+Fish and Wildlife Service, National Park Service, and California Department of Fish and Wildlife.
+*Interagency Conservation Strategy for Mountain Yellow-Legged Frogs in the Sierra Nevada*.
+U.S.
+Department of Agriculture, Forest Service, Pacific Southwest Region.
+
+U.S. Environmental Protection Agency (EPA). (2024). Climate Change Indicators: Snowpack. Retrieved from <https://www.epa.gov/climate-indicators/climate-change-indicators-snowpack>
+
+U.S.
+Fish and Wildlife Service.
+2014.
+Endangered and threatened wildlife and plants; endangered species status for Sierra Nevada yellow-legged frog and northern distinct population segment of the mountain yellow-legged frog, and threatened species status for Yosemite toad.
+*Federal Register* 79(82):24256–24310.
+
+Williams, B. K., Nichols, J. D., and Conroy, M. J. 2002. *Analysis and management of animal populations*. Academic Press, San Diego, California.
+
 ## Figures

### 2025-04-29 doc/Manuscript.qmd
draft manuscript for circulation 04/26/2025

@@
     number-sections: true
-    include-in-header: draft-options.tex
-bibliography: references.bib
+    geometry: margin=1in
+    include-in-header: |
+      \usepackage{lineno}
+      \AtBeginDocument{\linenumbers}
+bibliography: 
+  - references.bib
+  - manual.bib
 execute:
-  echo: true
+  echo: false
 ---
@@
 
-John Imperato^1^
+## Abstract
 
-^1^Yale School of the Environment
-
-## Abstract
+Mountain yellow-legged frogs (*Rana sierrae* and *R.
+muscosa*) have disappeared from \>90% of their historic range due to *Batrachochytrium dendrobatidis* (Bd) infection, yet a handful of populations now persist and recover in the pathogen's presence.
+We studied one such recovering population at Mossy Pond, Tahoe National Forest, for five summers (2014–2018) using robust-design capture–mark–recapture methods and a Bayesian Jolly–Seber model.
+From 3,631 captures of 657 uniquely marked adults, estimated abundance varied from 117 to 245 individuals, while 629 new adults recruited over the study.
+Monthly survival was strongly seasonal—mean 0.97 in winter versus 0.80 in summer—with male survival exceeding female survival by 2–7% in each interval.
+Survey-specific detection probability averaged 0.45.
+Bd was present in 67% of skin-swab samples and infection loads remained two–three orders of magnitude below lethal thresholds.
+Recruitment and abundance declined sharply after severe winters, yet overwinter adult survival showed no snow effect.
+These findings confirm functional Bd resistance, reveal contrasting seasonal pressures that shape population dynamics, and identify this robust northern Sierra Nevada population as a prime donor for translocation-based recovery efforts.
 
@@
 
-Amphibians continue to experience global declines, with infectious diseases recognized as primary drivers of biodiversity loss [@luedtke2023a] (Luedtke et al.
-2023).
-Chytridiomycosis, caused by the fungal pathogen *Batrachochytrium dendrobatidis* (Bd), has driven declines or extinctions of approximately 200 amphibian species globally [@skerratt2007] (Skerratt et al.
-2007).
-Amphibian responses to Bd infection vary dramatically; some populations show minimal impact, others rapidly go extinct, and a rare few exhibit persistent recovery following severe declines [@lips2016] (Lips 2016).
-The mechanisms supporting persistence or recovery remain poorly understood but likely involve interactions among ecological, evolutionary, immunological, and demographic factors [@scheele2017] (Scheele et al.
-2017).
+Amphibians continue to experience global declines, with infectious diseases recognized as primary drivers of biodiversity loss [@luedtke2023a].
+Chytridiomycosis, caused by the fungal pathogen *Batrachochytrium dendrobatidis* (Bd), has driven declines or extinctions of approximately 200 amphibian species globally [@skerratt2007].
+Amphibian responses to Bd infection vary dramatically; some populations show minimal impact, others rapidly go extinct, and a rare few exhibit persistent recovery following severe declines [@lips2016].
+The mechanisms supporting persistence or recovery remain poorly understood but likely involve interactions among ecological, evolutionary, immunological, and demographic factors [@scheele2017].
 
@@
 muscosa*), endemic to California's Sierra Nevada and Transverse mountain ranges, provides a particularly well-documented case of amphibian decline and recovery from Bd infection.
-Starting in the 1970s, Bd spread through the Sierra Nevada, causing widespread and rapid extirpations of MYL frog populations [@vredenburg2010] [@vredenburg2019] (Vredenburg et al.
-2010, 2019).
-By the early 2000s, over 90% of their historical range had been lost, due to combined impacts from Bd and habitat degradation caused by introduced trout [@knapp2000] (Knapp and Matthews 2000).
-Despite these losses, a few populations survived initial Bd outbreaks, eventually exhibiting increased abundance and apparent disease resistance [\@knapp2016](Knapp%20et%20al.%202016).
-These dynamics exemplify patterns of evolutionary rescue [@carlson2014] (Carlson et al.
-2014), in which populations rapidly acquire adaptive genetic variation that restores demographic viability under persistent threats [@knapp2024] (Knapp et al.
-2024).
+Starting in the 1970s, Bd spread through the Sierra Nevada, causing widespread and rapid extirpations of MYL frog populations [@vredenburg2010; @vredenburg2019].
+By the early 2000s, over 90% of their historical range had been lost, due to combined impacts from Bd and habitat degradation caused by introduced trout [@knapp2000].
+Despite these losses, a few populations survived initial Bd outbreaks, eventually exhibiting increased abundance and apparent disease resistance [@knapp2016].
+These dynamics exemplify patterns of evolutionary rescue [@carlson2014], in which populations rapidly acquire adaptive genetic variation that restores demographic viability under persistent threats [@knapp2024].
 
 While broad-scale studies have described these declines and recoveries, detailed demographic analyses at the individual population level remain scarce, primarily due to logistical challenges.
-Most fine-scale demographic research has focused on southern populations, particularly within Yosemite National Park [@fellers2013] [@joseph2018] [\@knapp2024](Fellers%20et%20al.%202013;%20Joseph%20and%20Knapp%202018;%20Knapp%20et%20al.%202024).
-In contrast, the population dynamics of frogs in the northern Sierra Nevada—where habitats are more heterogeneous and populations are smaller and lower in elevation—remain understudied [@brown2019] (MYLF ITT 2018; Brown et al.
-2020).
+Most fine-scale demographic research has focused on southern populations, particularly within Yosemite National Park [@fellers2013; @joseph2018; @knapp2024] .
+In contrast, the population dynamics of frogs in the northern Sierra Nevada—where habitats are more heterogeneous and populations are smaller and lower in elevation—remain understudied (MYLF ITT 2018) [@brown2019].
 
@@
 Our objectives were to estimate key demographic parameters—abundance, survival, recruitment, and detection probability—to provide insights into the ecological processes and factors that contribute to population viability under endemic Bd infection.
-We also examined the role of climate variability in shaping these demographic parameters, as climate change is increasingly recognized as a primary driver of amphibian declines worldwide, comparable in impact to disease and habitat loss [\@luedtke2023a](Luedtke%20et%20al.%202023).
+We also examined the role of climate variability in shaping these demographic parameters, as climate change is increasingly recognized as a primary driver of amphibian declines worldwide, comparable in impact to disease and habitat loss [@luedtke2023a].
 Understanding how climate influences abundance, survival, and recruitment is particularly important for high-elevation amphibian populations living with endemic Bd infection.
@@
 Understanding demographic patterns in recovering MYL frog populations has significant conservation implications.
-Recent research demonstrates these populations can serve as effective founders in reintroduction programs, establishing self-sustaining populations when translocated to suitable but currently unoccupied habitats [@knapp2024] (Knapp et al.
-2024).
+Recent research demonstrates these populations can serve as effective founders in reintroduction programs, establishing self-sustaining populations when translocated to suitable but currently unoccupied habitats [@knapp2024].
 Leveraging evolved disease resistance from naturally recovering populations offers a promising conservation strategy.
@@
 The study area includes the namesake Mossy Pond (6 hectare surface area, maximum depth 2.5 meters), its seasonally flowing outlet stream, and 12 ephemeral ponds.
-The absence of deep-water habitat is unusual for areas supporting MYL frog populations, given their reliance on water bodies deeper than 4 meters for overwinter survival [@bradford1983] [@knapp2003] (Bradford 1983; Knapp et al.
-2003).
+The absence of deep-water habitat is unusual for areas supporting MYL frog populations, given their reliance on water bodies deeper than 4 meters for overwinter survival [@bradford1983; @knapp2003].
 
 *R.
-sierrae* is a federally endangered, mid-sized ranid frog that occupies montane waterbodies in California's Sierra Nevada (U.S.
-Fish and Wildlife Service 2014).
+sierrae* is a federally endangered, mid-sized ranid frog that occupies montane waterbodies in California's Sierra Nevada [@u.s.fishandwildlifeservice2014].
 The US Geological Survey first detected *R.
 sierrae* in the Mossy Pond area in 1998, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.
-**More?**
 
@@
 sierrae* population's Bd status, we collected epithelial swabs from 24 adult frogs over three years (2008, 2010, and 2021).
-Partner scientists quantified the amount of Bd DNA on each swab using real-time quantitative polymerase chain reaction (qPCR) analysis (Boyle et al.
-2004).
-We report infection loads as the number of ITS1 (Internal Transcribed Spacer 1) gene copies detected [@longo2013] [@joseph2018] (Longo et al.
-2013; Joseph and Knapp 2018).
+Partner scientists quantified the amount of Bd DNA on each swab using real-time quantitative polymerase chain reaction (qPCR) analysis [@boyle2007].
+We report infection loads as the number of ITS1 (Internal Transcribed Spacer 1) gene copies detected [@longo2013; @joseph2018].
 Quantifying Bd loads rather than simply testing for Bd presence allowed us to compare our study population with other known recovering populations.
 Partner scientists provided *R.
-sierrae* Bd load data from three known naturally recovering populations in Yosemite National Park [@knapp2024] (Knapp et al.
-2024).
+sierrae* Bd load data from three known naturally recovering populations in Yosemite National Park [@knapp2024].
 
@@
 We studied *R.
-sierrae* with capture-mark-recapture (CMR) methods, which are widely used for estimating wildlife population parameters such as abundance, survival, and recruitment (Williams et al.
-2001).
+sierrae* with capture-mark-recapture (CMR) methods, which are widely used for estimating wildlife population parameters such as abundance, survival, and recruitment [@williams2001].
 We conducted CMR sampling according to Pollock's robust design [@pollock1982], which involves sampling at two temporal scales: primary periods (between which the population is open to demographic change) and secondary periods, during which the population is assumed closed.
@@
 Frogs larger than 40mm (adults) required further processing.
-New adult captures were PIT-tagged dorsally following McAlister et al.
-2004 [@mcallister2004].
+New adult captures were PIT-tagged dorsally following @mcallister2004.
 Next, we recorded the frog's sex and weighed it inside a tared plastic bag to the nearest 0.1 g using a Pesola spring scale.
@@
 We estimated survival, detection, recruitment, and adult population size for the Mossy Pond *R.
-sierrae* population using a site‐specific open‐population Jolly–Seber CMR model in a Bayesian framework with R statistical software \[[\@R](v4.4.2)\].
+sierrae* population using a site‐specific open‐population Jolly–Seber CMR model in a Bayesian framework with R statistical software [@R].
 Our implementation builds on the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), which tracks a superpopulation of M individuals through not‐recruited, alive, or dead states and models detection and demographic processes across primary and secondary periods [@knapp2024].
@@
 We defined winter severity as the percentage of historical average snow-water equivalent (SWE) remaining on April 1 each year (100% indicating average severity).
-We obtained SWE data from ten regional snow-survey stations (4.5–19.5 km from Mossy Pond) to capture the spatial variability in snowpack present in topographically complex mountainous environments [@revuelto2014] (Revuelto et al.
-2014).
-Annual SWE values (2011–2018) were interpolated across the study area using inverse-distance weighting in ArcGIS Pro 3.4.3 (Esri Inc.
-2025).
+We obtained SWE data from ten regional snow-survey stations (4.5–19.5 km from Mossy Pond) to capture the spatial variability in snowpack present in topographically complex mountainous environments [@revuelto2014].
+Annual SWE values (2011–2018) were interpolated across the study area using inverse-distance weighting in ArcGIS Pro 3.4.3.
 From the resulting continuous SWE raster, we extracted the mean percent SWE within our one-square-mile study boundary.
@@
 16 out of 24 samples tested positive for Bd, with infection intensities ranging from 16 to 43,318 ITS1 gene copies.
-These Bd loads are well below the \~600,000 gene copy threshold associated with severe chytridiomycosis and increased MYL frog mortality [@vredenburg2010] [@longo2013] (Vredenburg et al.
-2010; Longo et al.
-2013).
+These Bd loads are well below the \~600,000 gene copy threshold associated with severe chytridiomycosis and increased MYL frog mortality [@vredenburg2010; @longo2013].
 The infection intensities observed in our study population are typical of those measured in other known naturally recovering *R.
-sierrae* populations [@knapp2024] (Figure 1; Knapp et al.
-2024).
+sierrae* populations [Figure 1; @knapp2024].
 
@@
 The Mossy Pond frogs carried Bd at moderate prevalence (67%) but at infection intensities that were two to three orders of magnitude lower than levels associated with increased mortality in MYL frogs.
-Such low loads mirror those reported for other naturally recovering populations in Yosemite National Park and stand in stark contrast to the uniformly high loads (\>600,000 ITS1 copies) and near-certain mortality that typify collapsing populations [@vredenburg2010] (Vredenburg et al.
-2010).
-Taken together with the population’s stable abundance over the study period, these data support the hypothesis that the Mossy Pond frogs have undergone an evolutionary rescue process in which genotypes conferring resistance have been favoured and now dominate the population.
-This expands the geographic scope of Bd-resistant *R.
-sierrae* beyond the southern Sierra Nevada, indicating that resistance is not confined to a single environmental setting or genetic lineage.
-**Too much?**
+Such low loads mirror those reported for other naturally recovering populations in Yosemite National Park and stand in stark contrast to the uniformly high loads (\>600,000 ITS1 copies) and near-certain mortality that typify collapsing populations [@vredenburg2010].
+Coupled with the population’s stable abundance over the study period, these patterns are consistent with the idea that some degree of Bd resistance as arisen in this population.
+While such dynamics could reflect an evolutionary rescue process in which genotypes conferring lower susceptibility have increased in frequency, additional genetic and experimental evidence would be required to confirm that mechanism.
+This expands the geographic scope of Bd-resistant R.
+sierrae beyond the southern Sierra Nevada, indicating that resistance is not confined to a single environmental setting or genetic lineage.
 
@@
 Mild, low-snow winters may enhance juvenile overwinter survival and accelerate growth such that more animals reach the 40 mm adult threshold before the next breeding season.
-This mechanism has been invoked for long-lived amphibians in other montane systems [@mccaffery2010] (McCaffery & Maxell 2010) and is consistent with the lag-1 pattern we observe.
+This mechanism has been invoked for long-lived amphibians in other montane systems [@mccaffery2010] and is consistent with the lag-1 pattern we observe.
 The absence of any clear winter-severity signal in overwinter adult survival suggests that once frogs surpass a critical body size, they can tolerate a broad range of cold-season conditions.
@@
 Although the study area lacks the deep-water refuges (\>4m) traditionally thought essential, our capture records show adults concentrated in the rocky outlet stream and mud-bottomed main pond—habitats that can stay unfrozen and oxygenated thanks to possible groundwater inputs and insulating effects.
-**More from Isaac here?**
 
@@
 sierrae at Summit Meadow, with females sometimes faring slightly better.
-Our pattern instead mirrors Rocky Mountain tailed frogs (Ascaphus montanus), where males have shown modestly higher survival [@honeycutt2019] (Honeycutt et al.
-2019).
+Our pattern instead mirrors Rocky Mountain tailed frogs (*Ascaphus montanus*), where males have shown modestly higher survival [@honeycutt2019].
 
@@
 Most obviously, the energetic burden of reproduction falls almost entirely on females: producing and depositing large gelatinous egg masses—often while bearing the added weight of one or more males in amplexus—depletes energy reserves and can delay post-breeding foraging.
-Post-oviposition condition declines have been linked to reduced annual survival in other anurans (e.g., common toads; [@madsen2010] Loman & Madsen 2010).
+Post-oviposition condition declines have been linked to reduced annual survival in other anurans [e.g., common toads; @madsen2010].
 Additionally, movement studies of *A.
-montanus* show that females travel farther than males to locate suitable oviposition sites, increasing exposure to predators and desiccating microhabitats [@honeycutt2019] (Honeycutt et al.
-2019).
+montanus* show that females travel farther than males to locate suitable oviposition sites, increasing exposure to predators and desiccating microhabitats [@honeycutt2019].
 Together, these reproductive and behavioural costs may magnify female summer mortality in the Mossy Pond *R.
@@
 
-Across the western United States, April-1 SWE has declined by about 18% on average since 1955, with particularly steep losses in northern California (US EPA 2024 indicator).
-Regional climate projections point to continued warming, more precipitation falling as rain rather than snow, and further reductions in spring snowpack over the coming decades [@gottlieb2024] (Gottlieb and Mankin 2024).
+Across the western United States, April-1 SWE has declined by about 18% on average since 1955, with particularly steep losses in northern California [@usepa2016].
+Regional climate projections point to continued warming, more precipitation falling as rain rather than snow, and further reductions in spring snowpack over the coming decades [@gottlieb2024].
 In the short term, such trends could benefit Mossy Pond frogs by sustaining the sequence of “light” winters that boost recruitment.
@@
 
-Our results place the Mossy Pond frogs squarely within the “resistant-donor” framework that Knapp et al.
-[@knapp2024] (2024) used to successfully facilitate landscape-scale *R.
+Our results place the Mossy Pond frogs squarely within the “resistant-donor” framework that @knapp2024 used to successfully facilitate landscape-scale *R.
 sierrae* recovery in Yosemite National Park.
@@
 ## References
-
-California Department of Water Resources. 2025. California Data Exchange Center (CDEC). California Department of Water Resources. <https://cdec.water.ca.gov>
-
-Esri Inc. 2025. ArcGIS Pro (Version 3.4.3). Esri Inc. <https://www.esri.com/en-us/arcgis/products/arcgis-pro/overview>
-
-MYLF ITT.
-2018.
-USDA Forest Service, U.S.
-Fish and Wildlife Service, National Park Service, and California Department of Fish and Wildlife.
-*Interagency Conservation Strategy for Mountain Yellow-Legged Frogs in the Sierra Nevada*.
-U.S.
-Department of Agriculture, Forest Service, Pacific Southwest Region.
-
-U.S. Environmental Protection Agency (EPA). (2024). Climate Change Indicators: Snowpack. Retrieved from <https://www.epa.gov/climate-indicators/climate-change-indicators-snowpack>
-
-U.S.
-Fish and Wildlife Service.
-2014.
-Endangered and threatened wildlife and plants; endangered species status for Sierra Nevada yellow-legged frog and northern distinct population segment of the mountain yellow-legged frog, and threatened species status for Yosemite toad.
-*Federal Register* 79(82):24256–24310.
-
-Williams, B. K., Nichols, J. D., and Conroy, M. J. 2002. *Analysis and management of animal populations*. Academic Press, San Diego, California.
-
-## Figures

### 2025-05-01 doc/Manuscript.qmd
minor updates
add line numbers to rendered .pdf

@@
 ---
+title: Dynamics of an Endangered Frog Population Recovering from Bd-Driven Decline
 format:
@@
     linestretch: 2
-    number-sections: true
+    number-sections: False
     geometry: margin=1in
-    include-in-header: |
+    include-in-header:
+      text: |
         \usepackage{lineno}
-      \AtBeginDocument{\linenumbers}
+        \linenumbers
 bibliography: 
@@
 ---
-
-# Dynamics of an Endangered Frog Population Recovering from Bd-Driven Decline
 
@@
 Afterward, if crews captured the same individual on a subsequent day during the same primary period, crews only recorded PIT tag, sex, and location data.
-
 Over the 5-year period of our study, we recorded 3,631 captures of 657 individual frogs.
@@
 Such low loads mirror those reported for other naturally recovering populations in Yosemite National Park and stand in stark contrast to the uniformly high loads (\>600,000 ITS1 copies) and near-certain mortality that typify collapsing populations [@vredenburg2010].
-Coupled with the population’s stable abundance over the study period, these patterns are consistent with the idea that some degree of Bd resistance as arisen in this population.
+Coupled with the population’s stable abundance over the study period, these patterns are consistent with the idea that some degree of Bd resistance has arisen in this population.
 While such dynamics could reflect an evolutionary rescue process in which genotypes conferring lower susceptibility have increased in frequency, additional genetic and experimental evidence would be required to confirm that mechanism.

### 2025-05-07 code/R/Survival_by_sex.Rmd
organize R code files

@@
 
+write_csv(time_summaries_df, "/Users/johnimperato/Downloads/MF_survival.csv")
+
 ```
@@
   labs(
-    title = "Difference in Survival Probabilities (M - F) Between Primary Periods\nsurvival_fill_value = 'M'",
     x = "Primary Period",
-    y = "Difference in Survival Probability"
+    y = "Difference in Survival Probability (M-F)"
   ) +
@@
 # Save the plot to a file
-ggsave("/Users/johnimperato/Desktop/Mossy-Pond_companion/difference_in_survival_probabilities_M_fill.png", 
+ggsave("/Users/johnimperato/Desktop/Mossy-Pond_companion/difference_in_survival_probabilities_ratio_fill.png", 
        plot = difference_plot, width = 10, height = 6, dpi = 300, bg = "white")
 ```
+
+```{r}
+
+library(ggplot2)
+library(dplyr)
+
+# 1. Prepare transition labels
+time_summaries_df <- time_summaries_df %>%
+  arrange(time_index) %>%
+  mutate(
+    transition = factor(
+      paste0(time_index, "-", time_index + 1),
+      levels = paste0(time_index, "-", time_index + 1)
+    )
+  )
+
+# 2. Compute y-axis lower bound so that zero sits above the bottom 
+y_lower <- min(time_summaries_df$lower95, 0) - 0.1
+y_upper <- max(time_summaries_df$upper95) + 0.01
+
+# 3. Plot
+difference_plot <- ggplot(time_summaries_df, aes(x = transition, y = mean)) +
+  geom_point(size = 2, color = "black") +
+  geom_errorbar(aes(ymin = lower95, ymax = upper95),
+                width = 0.2, color = "black") +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "#B2182B") +
+  scale_y_continuous(
+    limits = c(y_lower, y_upper),
+    expand = c(0, 0)
+  ) +
+  labs(
+    x = "Primary Period Transitions",
+    y = "Difference in Survival Probability (M–F)"
+  ) +
+  theme_minimal(base_size = 12) +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    axis.title    = element_text(face = "bold"),
+    panel.grid.minor = element_blank()
+  )
+
+# 4. Display
+print(difference_plot)
+
+```

### 2025-05-07 code/R/WinterSeverity_analysis_MossyPond.Rmd
organize R code files

@@
 avg_swe <- avg_swe %>%
-  rename(Winter_Severity = Average.Pixel.Value) %>%
   mutate(Winter = c("2010-2011", "2011-2012", "2012-2013", "2013-2014", "2014-2015", "2015-2016", "2016-2017", "2017-2018"))
@@
 
-## Population Vital Rates/Numbers and Growth Rate
+## Population Vital Rates/Numbers
 
@@
 
-model_summary <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_custom_mod_sumary.csv')
+model_summary <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_custom_mod_summary.csv')
 
@@
 N <- model_summary[33659:33674, ]
+
+View(N)
 
@@
 # extract overwinter survival values from model summary
-phi <- B <- model_summary[c(33738:33752), ]
+phi <- model_summary[c(33738:33752), ]
+View(phi)
 
@@
   select(Winter, Overwinter_Survival_Probability)
-
-print(N)
-print(B)
 
@@
   theme(plot.title = element_text(hjust = 0.5))
+
+abundance_WinterSeverity_2yr
+correlation_plot_2_N
 
@@
 ggsave("WinterSeverityB1.png", plot = WinterSeverity_B1, bg = "white")
-ggsave("CorrelationB1.png", plot = correlation_plot_B1, bg = "white")
+ggsave("CorrelationB1_narrow.png", plot = correlation_plot_B1, bg = "white")
 
@@
     x = "Winter Severity (% of Historical Average 1 April SWE)",
-    y = "Yearly Mean Abundance"
+    y = "Yearly Recruitment"
   ) +

### 2025-05-07 doc/Manuscript.qmd
organize R code files

@@
 ---
-title: Dynamics of an Endangered Frog Population Recovering from Bd-Driven Decline
+title: Popualtion Dynamics of an Endangered Frog Recovering from Disease-Driven Decline
 format:
@@
 
-Mountain yellow-legged frogs (*Rana sierrae* and *R.
-muscosa*) have disappeared from \>90% of their historic range due to *Batrachochytrium dendrobatidis* (Bd) infection, yet a handful of populations now persist and recover in the pathogen's presence.
-We studied one such recovering population at Mossy Pond, Tahoe National Forest, for five summers (2014–2018) using robust-design capture–mark–recapture methods and a Bayesian Jolly–Seber model.
-From 3,631 captures of 657 uniquely marked adults, estimated abundance varied from 117 to 245 individuals, while 629 new adults recruited over the study.
-Monthly survival was strongly seasonal—mean 0.97 in winter versus 0.80 in summer—with male survival exceeding female survival by 2–7% in each interval.
+Mountain yellow-legged frogs (*Rana sierrae* and *Rana muscosa*) have disappeared from \>90% of their historic range due to the virulent fungal pathogen *Batrachochytrium dendrobatidis* (Bd), yet a handful of populations now persist and recover in the pathogen's presence.
+We studied one such recovering population at Mossy Pond, Tahoe National Forest, for five summers (2014–2018) using robust-design capture–mark–recapture methods and analyzed the capture histories with a Bayesian Jolly–Seber model.
+From 3,631 captures of 657 uniquely marked adults, estimated abundance varied from 117 to 245 individuals, and a total of 629 new adults recruited over the study.
+Monthly survival was higher in winter than in summer (mean and credible interval; winter: 0.97 (xx-xx), summer: 0.8 (xx-xx) and male survival exceeded female survival by 2–7% in each interval.
 Survey-specific detection probability averaged 0.45.
-Bd was present in 67% of skin-swab samples and infection loads remained two–three orders of magnitude below lethal thresholds.
+Bd was present in 67% of skin-swab samples and infection loads were two–three orders of magnitude below lethal thresholds.
 Recruitment and abundance declined sharply after severe winters, yet overwinter adult survival showed no snow effect.
-These findings confirm functional Bd resistance, reveal contrasting seasonal pressures that shape population dynamics, and identify this robust northern Sierra Nevada population as a prime donor for translocation-based recovery efforts.
+These findings suggest this population is persisting despite ongoing Bd infection, perhaps as a result of some degree of infection resistance.
+.................
+identify this robust northern Sierra Nevada population as a prime donor for reintroduction-based recovery efforts.
 
@@
 Chytridiomycosis, caused by the fungal pathogen *Batrachochytrium dendrobatidis* (Bd), has driven declines or extinctions of approximately 200 amphibian species globally [@skerratt2007].
-Amphibian responses to Bd infection vary dramatically; some populations show minimal impact, others rapidly go extinct, and a rare few exhibit persistent recovery following severe declines [@lips2016].
+Amphibian responses to Bd infection vary dramatically; some populations show minimal impact, others rapidly go extinct, and a rare few exhibit persistence or even recovery following severe declines [@lips2016].
 The mechanisms supporting persistence or recovery remain poorly understood but likely involve interactions among ecological, evolutionary, immunological, and demographic factors [@scheele2017].
 
-The mountain yellow-legged (MYL) frog species complex (*Rana sierrae* and *R.
-muscosa*), endemic to California's Sierra Nevada and Transverse mountain ranges, provides a particularly well-documented case of amphibian decline and recovery from Bd infection.
+The mountain yellow-legged (MYL) frog species complex (*Rana sierrae* and *Rana muscosa*), endemic to California's Sierra Nevada and Transverse ranges, provides a particularly well-documented case of amphibian decline and recovery from Bd infection.
 Starting in the 1970s, Bd spread through the Sierra Nevada, causing widespread and rapid extirpations of MYL frog populations [@vredenburg2010; @vredenburg2019].
-By the early 2000s, over 90% of their historical range had been lost, due to combined impacts from Bd and habitat degradation caused by introduced trout [@knapp2000].
+By the early 2000s, over 90% of historical populations had been lost, due to combined impacts from Bd and introduced predatory trout [@knapp2000].
 Despite these losses, a few populations survived initial Bd outbreaks, eventually exhibiting increased abundance and apparent disease resistance [@knapp2016].
-These dynamics exemplify patterns of evolutionary rescue [@carlson2014], in which populations rapidly acquire adaptive genetic variation that restores demographic viability under persistent threats [@knapp2024].
+These dynamics are consistent with patterns of evolutionary rescue [@carlson2014], in which populations rapidly acquire adaptive genetic variation that restores demographic viability under persistent threats [e.g.
+@knapp2024].
 
-While broad-scale studies have described these declines and recoveries, detailed demographic analyses at the individual population level remain scarce, primarily due to logistical challenges.
+Although broad-scale studies have described these declines and recoveries, detailed demographic analyses at the individual population level remain scarce, primarily due to logistical challenges.
 Most fine-scale demographic research has focused on southern populations, particularly within Yosemite National Park [@fellers2013; @joseph2018; @knapp2024] .
-In contrast, the population dynamics of frogs in the northern Sierra Nevada—where habitats are more heterogeneous and populations are smaller and lower in elevation—remain understudied (MYLF ITT 2018) [@brown2019].
+In contrast, the population dynamics of *R.
+sierrae* in the northern Sierra Nevada—where habitats are more heterogeneous and populations are smaller and lower in elevation—remain understudied (MYLF ITT 2018) [@brown2019].
 
-To address this gap, we conducted a five-year capture-mark-recapture (CMR) study of a naturally recovering, Bd-persistent *R.
+To address this gap, we conducted a five-year capture-mark-recapture (CMR) study of a naturally recovering, Bd-positive *R.
 sierrae* population in Tahoe National Forest.
@@
 
-The Mossy Pond study area is a one square mile section of the Tahoe National Forest, north of Highway 80 in Nevada County, California.
-It is characterized by lakes, ponds, and streams set on granite benches, ranging from 6,400 feet to 7,100 feet in elevation.
+The Mossy Pond study area is a 2.6 square km section of the Tahoe National Forest, north of Highway 80 in Nevada County, California.
+It is characterized by lakes, ponds, and streams set on granite benches, ranging from 1,950 m to 2,164 m in elevation.
 The study area includes the namesake Mossy Pond (6 hectare surface area, maximum depth 2.5 meters), its seasonally flowing outlet stream, and 12 ephemeral ponds.
-The absence of deep-water habitat is unusual for areas supporting MYL frog populations, given their reliance on water bodies deeper than 4 meters for overwinter survival [@bradford1983; @knapp2003].
+The absence of deep-water habitat is unusual for areas supporting MYL frog populations, given their general reliance on water bodies deeper than 3-4 meters for overwinter survival [@bradford1983; @knapp2003].
 
@@
 sierrae* is a federally endangered, mid-sized ranid frog that occupies montane waterbodies in California's Sierra Nevada [@u.s.fishandwildlifeservice2014].
-The US Geological Survey first detected *R.
-sierrae* in the Mossy Pond area in 1998, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.
+*R.
+sierrae* was first detected in the Mossy Pond area in 1998 **(citation)**, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.
 
@@
 To assess the Mossy Pond *R.
-sierrae* population's Bd status, we collected epithelial swabs from 24 adult frogs over three years (2008, 2010, and 2021).
-Partner scientists quantified the amount of Bd DNA on each swab using real-time quantitative polymerase chain reaction (qPCR) analysis [@boyle2007].
+sierrae* population's Bd status, we collected skin swabs from 24 adult frogs over three years (2008, 2010, and 2021).
+The amount of Bd DNA on each swab was quantified using real-time quantitative polymerase chain reaction (qPCR) analysis [@boyle2007].
 We report infection loads as the number of ITS1 (Internal Transcribed Spacer 1) gene copies detected [@longo2013; @joseph2018].
-Quantifying Bd loads rather than simply testing for Bd presence allowed us to compare our study population with other known recovering populations.
-Partner scientists provided *R.
-sierrae* Bd load data from three known naturally recovering populations in Yosemite National Park [@knapp2024].
+To compare the Bd loads of Mossy Pond frogs to those of other known recovering *R.
+sierrae* populations, we compared Bd loads from Mossy Pond to those from three known naturally recovering populations in Yosemite National Park [@knapp2024].
 
@@
 
-We studied *R.
-sierrae* with capture-mark-recapture (CMR) methods, which are widely used for estimating wildlife population parameters such as abundance, survival, and recruitment [@williams2001].
-We conducted CMR sampling according to Pollock's robust design [@pollock1982], which involves sampling at two temporal scales: primary periods (between which the population is open to demographic change) and secondary periods, during which the population is assumed closed.
+We surveyed the Mossy Pond *R.
+sierrae* population using capture-mark-recapture (CMR) methods, which are widely used for estimating wildlife population parameters such as abundance, survival, and recruitment [@williams2001].
+We conducted CMR sampling according to Pollock's robust design [@pollock1982], which involves sampling at two temporal scales: primary periods, between which the population is open to demographic change, and secondary periods, during which the population is assumed closed.
 Each summer from 2014 to 2018, we visited the study site 3 or 4 times (primary periods) for three consecutive days of surveys (secondary periods), for a total of 47 individual surveys.

### 2025-05-07 code/R/MP_captures_formatting.R

@@ -1,33 +1,0 @@
-# Load capture data and format 'captures' data frame for mrmr model
-
-# Load .csv of all captures from the Mossy Pond CMR study period
-captures <- read.csv("/Users/johnimperato/Documents/Capture_Data.csv")
-
-# Remove "00:00:00 EDT" from Date column for compatibility with mrmr 
-captures$Date <- gsub("00:00:00 EDT","", captures$Date)
-
-# Convert 'Date' variable to date object class 
-captures$Date <- as.Date(captures$Date, format = "%a %b %d  %Y")
-
-# Change variable names for compatibility with mrmr 
-names(captures)[names(captures) == "Date"] <- "survey_date"
-names(captures)[names(captures) == "PIT_TagCode"] <- "pit_tag_id"
-
-# Change data types for compatibility with mrmr
-captures$survey_date <- as.character(captures$survey_date)
-captures$pit_tag_id <- as.character(captures$pit_tag_id)
-
-# Remove the single dead frog observation (instead, use it to create a 'removals' data frame)
-captures <- captures[captures$frog_state != "dead", ]
-
-# ATTEMPT TO FIGURE THE PROBLEM OUT
-# Include only a subset of columns from captures. This seems to have worked in the past. 
-### captures <- unique(captures[ , c(2,3,5,7)])
-# when I create this subset, 5 observations disappear
-# This subset didn't work. Lets try getting rid of CA lake id column
-captures <- unique(captures[ , c(3,5)])
-# another 2 observations disappeared
-# I have not yet run this model
-
-write.csv( captures , '/Users/johnimperato/Desktop/Mossy-Pond_companion/MP_captures.csv' )
-

### 2025-05-07 code/R/classic_fit_model_function.R

@@ -1,78 +1,0 @@
-fit_model_classic <- function(
-    data,
-    seed = NULL,
-    refresh = NULL,
-    init = NULL,
-    save_latent_dynamics = FALSE,
-    output_dir = NULL,
-    output_basename = NULL,
-    sig_figs = NULL,
-    chains = 4,
-    parallel_chains = getOption("mc.cores", 1),
-    chain_ids = seq_len(chains),
-    threads_per_chain = NULL,
-    opencl_ids = NULL,
-    iter_warmup = 1000,
-    iter_sampling = 1000,
-    save_warmup = FALSE,
-    thin = NULL,
-    max_treedepth = NULL,
-    adapt_engaged = TRUE,
-    adapt_delta = NULL,
-    step_size = NULL,
-    metric = NULL,
-    metric_file = NULL,
-    inv_metric = NULL,
-    init_buffer = NULL,
-    term_buffer = NULL,
-    window = NULL,
-    fixed_param = FALSE,
-    show_messages = TRUE,
-    compile = TRUE,
-    ...
-) {
-  # updated to use my custom stan model
-  stan_file <- "/Users/johnimperato/Desktop/Mossy-Pond/code/stan/JI_classic_twostate.stan" # specify local path
-  
-  model <- cmdstan_model(stan_file, ...)
-  
-  m_fit <- model$sample(
-    data = data$stan_d,
-    seed = seed,
-    refresh = refresh,
-    init = init,
-    save_latent_dynamics = save_latent_dynamics,
-    output_dir = output_dir,
-    output_basename = output_basename,
-    sig_figs = sig_figs,
-    chains = chains,
-    parallel_chains = parallel_chains,
-    chain_ids = chain_ids,
-    threads_per_chain = threads_per_chain,
-    opencl_ids = opencl_ids,
-    iter_warmup = iter_warmup,
-    iter_sampling = iter_sampling,
-    save_warmup = save_warmup,
-    thin = thin,
-    max_treedepth = max_treedepth,
-    adapt_engaged = adapt_engaged,
-    adapt_delta = adapt_delta,
-    step_size = step_size,
-    metric = metric,
-    metric_file = metric_file,
-    inv_metric = inv_metric,
-    init_buffer = init_buffer,
-    term_buffer = term_buffer,
-    window = window,
-    fixed_param = fixed_param,
-    show_messages = show_messages
-  )
-  
[…]

### 2025-05-07 code/R/custom_fit_model_function.R

@@ -1,80 +1,0 @@
-fit_model_custom <- function(
-    data,
-    seed = NULL,
-    refresh = NULL,
-    init = NULL,
-    save_latent_dynamics = FALSE,
-    output_dir = NULL,
-    output_basename = NULL,
-    sig_figs = NULL,
-    chains = 4,
-    parallel_chains = getOption("mc.cores", 1),
-    chain_ids = seq_len(chains),
-    threads_per_chain = NULL,
-    opencl_ids = NULL,
-    iter_warmup = 1000,
-    iter_sampling = 1000,
-    save_warmup = FALSE,
-    thin = NULL,
-    max_treedepth = NULL,
-    adapt_engaged = TRUE,
-    adapt_delta = NULL,
-    step_size = NULL,
-    metric = NULL,
-    metric_file = NULL,
-    inv_metric = NULL,
-    init_buffer = NULL,
-    term_buffer = NULL,
-    window = NULL,
-    fixed_param = FALSE,
-    show_messages = TRUE,
-    compile = TRUE,
-    ...
-) {
-  # updated to use my custom stan model
-  stan_file <- "/Users/johnimperato/Desktop/Mossy-Pond/code/stan/JI_custom_twostate.stan" # specify local path
-  
-  model <- cmdstan_model(stan_file, ...)
-  
-  m_fit <- model$sample(
-    data = data$stan_d,
-    seed = seed,
-    refresh = refresh,
-    init = init,
-    save_latent_dynamics = save_latent_dynamics,
-    output_dir = output_dir,
-    output_basename = output_basename,
-    sig_figs = sig_figs,
-    chains = chains,
-    parallel_chains = parallel_chains,
-    chain_ids = chain_ids,
-    threads_per_chain = threads_per_chain,
-    opencl_ids = opencl_ids,
-    iter_warmup = iter_warmup,
-    iter_sampling = iter_sampling,
-    save_warmup = save_warmup,
-    thin = thin,
-    max_treedepth = max_treedepth,
-    adapt_engaged = adapt_engaged,
-    adapt_delta = adapt_delta,
-    step_size = step_size,
-    metric = metric,
-    metric_file = metric_file,
-    inv_metric = inv_metric,
-    init_buffer = init_buffer,
-    term_buffer = term_buffer,
-    window = window,
-    fixed_param = fixed_param,
-    show_messages = show_messages
-  )
-  
[…]

### 2025-05-07 code/R/B.Rmd
initial commits of individual parameter extraction scripts

---
title: "B"
output: html_document
date: "2025-05-07"
---

```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Recruitment (B) summaries by primary period & annual totals
# ─────────────────────────────────────────────────────────────────────────────

# 0) Load packages
# install.packages(c("dplyr","tidyr","stringr","readr"))
library(dplyr)
library(tidyr)
library(stringr)
library(readr)

# 1) Load model & survey data
mod     <- readRDS("/Users/johnimperato/Desktop/Mossy-Pond_companion/Model_2.0.rds")
surveys <- read_csv("/Users/johnimperato/Desktop/Mossy-Pond/data/survey_MossyPond.csv") %>%
             mutate(survey_date = as.Date(survey_date))

# 2) Build table of primary period dates
period_dates <- surveys %>%
  group_by(primary_period) %>%
  summarize(
    start_date = min(survey_date),
    end_date   = max(survey_date),
    .groups    = "drop"
  )

# 3) Extract all B draws (B[1] = recruitment 1→2, etc.)
b_draws_raw <- mod$m_fit$draws("B", format = "draws_df")

# 4) Pivot long & attach dates
b_draws <- b_draws_raw %>%
  pivot_longer(
    cols      = starts_with("B["),
    names_to  = "name",
    values_to = "B_raw"
  ) %>%
  mutate(
    # interval index → primary_period
    t              = as.integer(str_extract(name, "(?<=\\[)\\d+(?=\\])")),
    primary_period = t
  ) %>%
  left_join(period_dates, by = "primary_period")

# 5) Map primary_period → calendar Year
Year_map <- tibble(
  primary_period = 1:16,
  Year = c(
    rep(2014, 3),  # periods 1,2,3 → 2014
    rep(2015, 3),  # 4,5,6 → 2015
    rep(2016, 3),  # 7,8,9 → 2016
    rep(2017, 3),  # 10,11,12 → 2017
    rep(2018, 4)   # 13,14,15,16 → 2018
  )
)

b_draws <- b_draws %>%
  left_join(Year_map, by = "primary_period")

# 6) Interval‐specific summary (as before)
Recruitment_summary <- b_draws %>%
  group_by(primary_period, start_date, end_date) %>%
  summarize(
    median     = median(B_raw),
    lo025    = quantile(B_raw, 0.025),
    hi975    = quantile(B_raw, 0.975),
    cri_width = hi975 - lo025,
    .groups  = "drop"
  ) %>%
  arrange(primary_period)

# 7) Annual recruitment: sum within each posterior draw, then summarize
#    We group on the internal .draw index so covariance is preserved.
annual_B_draws <- b_draws %>%
  group_by(.draw, Year) %>%
  summarise(
    B_year = sum(B_raw),
    .groups = "drop"
  )

Yearly_Recruitment_summary <- annual_B_draws %>%
  group_by(Year) %>%
  summarise(
    median     = median(B_year),
    lo025    = quantile(B_year, 0.025),
    hi975    = quantile(B_year, 0.975),
    cri_width = hi975 - lo025,
    .groups  = "drop"
  ) %>%
  arrange(Year)

# 8) Print or export
print(Recruitment_summary)
print(Yearly_Recruitment_summary)

# Optionally save to CSV:
# write_csv(Recruitment_summary,        "Recruitment_summary.csv")
# write_csv(Yearly_Recruitment_summary, "Yearly_Recruitment_summary.csv")

```

### 2025-05-07 code/R/N.Rmd
initial commits of individual parameter extraction scripts

---
title: 'N'
output: html_document
date: "2025-05-07"
---

```{r}

# ─────────────────────────────────────────────────────────────────────────────
# Abundance (N) summaries by primary period
# ─────────────────────────────────────────────────────────────────────────────

# 0) Load packages
# install.packages(c("dplyr","tidyr","stringr","readr"))
library(dplyr)
library(tidyr)
library(stringr)
library(readr)

# 1) Load model & surveys
mod     <- readRDS("/Users/johnimperato/Desktop/Mossy-Pond_companion/Model_2.0.rds")
surveys <- read_csv("/Users/johnimperato/Desktop/Mossy-Pond/data/survey_MossyPond.csv") %>%
             mutate(survey_date = as.Date(survey_date))

# 2) Build table of primary period dates
period_dates <- surveys %>%
  group_by(primary_period) %>%
  summarize(
    start_date = min(survey_date),
    end_date   = max(survey_date),
    .groups    = "drop"
  )

# 3) Extract all N draws (N[1] = period 2, N[2] = period 3, …)
n_draws_raw <- mod$m_fit$draws("N", format = "draws_df")

# 4) Pivot long & attach dates
n_draws <- n_draws_raw %>%
  pivot_longer(
    cols      = starts_with("N["),
    names_to  = "name",
    values_to = "N_raw"
  ) %>%
  mutate(
    t             = as.integer(str_extract(name, "(?<=\\[)\\d+(?=\\])")),
    primary_period = t
  ) %>%
  left_join(period_dates, by = "primary_period")

# 5) Summarize
Abundance_summary <- n_draws %>%
  group_by(primary_period, start_date, end_date) %>%
  summarize(
    median   = median(N_raw),
    lo025  = quantile(N_raw, 0.025),
    hi975  = quantile(N_raw, 0.975),
    ci_width = hi975 - lo025,
    .groups  = "drop"
  ) %>%
  arrange(primary_period)

# 6) Output
print(Abundance_summary)
# write_csv(Abundance_summary, "Abundance_summary.csv")

```

### 2025-05-07 code/R/p.Rmd
initial commits of individual parameter extraction scripts

---
title: "p"
output: html_document
date: "2025-05-07"
---

```{r}

# ─────────────────────────────────────────────────────────────────────────────
# Detection (p) summaries by survey
# ─────────────────────────────────────────────────────────────────────────────

# 0) Load packages
# install.packages(c("dplyr","tidyr","stringr","readr"))
library(dplyr)
library(tidyr)
library(stringr)
library(readr)

# 1) Load model & survey data (add survey_idx for joining)
mod     <- readRDS("/Users/johnimperato/Desktop/Mossy-Pond_companion/Model_2.0.rds")
surveys <- read_csv("/Users/johnimperato/Desktop/Mossy-Pond/data/survey_MossyPond.csv") %>%
  mutate(
    survey_date = as.Date(survey_date),
    survey_idx  = row_number()
  )

# 2) Extract all p draws (p[1] … p[Jtot])
p_draws_raw <- mod$m_fit$draws("p", format = "draws_df")

# 3) Pivot to long form and attach survey metadata
p_draws <- p_draws_raw %>%
  pivot_longer(
    cols      = starts_with("p["),
    names_to  = "name",
    values_to = "p_raw"
  ) %>%
  mutate(
    survey_idx = as.integer(str_extract(name, "(?<=\\[)\\d+(?=\\])"))
  ) %>%
  left_join(surveys, by = "survey_idx")

# 4) Summarize detection probability by survey
Detection_summary <- p_draws %>%
  group_by(survey_idx, primary_period, survey_date) %>%
  summarize(
    median    = median(p_raw),
    lo025     = quantile(p_raw, 0.025),
    hi975     = quantile(p_raw, 0.975),
    ci_width  = hi975 - lo025,
    .groups   = "drop"
  ) %>%
  arrange(survey_date)

# 5) Print or export results
print(Detection_summary)
# write_csv(Detection_summary, "Detection_summary.csv")

```

### 2025-05-07 code/R/phi.Rmd
initial commits of individual parameter extraction scripts

---
title: "Phi"
output: html_document
date: "2025-05-07"
---

```{r}

# ─────────────────────────────────────────────────────────────────────────────
# Interval & Seasonal Monthly Survival with Dates, CI width & Season
# ─────────────────────────────────────────────────────────────────────────────

# 0) install / load required packages
# install.packages(c("dplyr","tidyr","stringr","readr"))
library(dplyr)
library(tidyr)
library(stringr)
library(readr)

# 1) Load your model and survey data
mod <- readRDS("/Users/johnimperato/Desktop/Mossy-Pond_companion/Model_2.0.rds")

surveys <- read_csv("/Users/johnimperato/Desktop/Mossy-Pond/data/survey_MossyPond.csv") %>%
  mutate(survey_date = as.Date(survey_date))

# 2) Compute first & last survey date for each primary period
period_dates <- surveys %>%
  group_by(primary_period) %>%
  summarise(
    start_date = min(survey_date),
    end_date   = max(survey_date),
    .groups    = "drop"
  )

# 3) Build the intervals table: transition index, from→to, start/end dates, length
intervals <- period_dates %>%
  mutate(
    from       = primary_period,
    to         = lead(primary_period),
    next_start = lead(start_date)
  ) %>%
  filter(!is.na(to)) %>%
  mutate(
    delta_days   = as.numeric(difftime(next_start, end_date, units = "days")),
    delta_months = delta_days / 30.44,
    transition   = from
  ) %>%
  select(
    transition,
    from, 
    to,
    start_date,
    end_date,
    delta_days,
    delta_months
  )

# 4) Extract all raw posterior draws of overall_phi
phi_draws_raw <- mod$m_fit$draws("overall_phi", format = "draws_df")

# 5) Pivot to long form, parse the Stan index, join intervals, and standardize to monthly
phi_draws <- phi_draws_raw %>%
  pivot_longer(
    cols      = starts_with("overall_phi"),
    names_to  = "name",
    values_to = "phi_raw"
  ) %>%
  mutate(
    transition = as.integer(str_extract(name, "(?<=\\[)\\d+(?=\\])"))
  ) %>%
  left_join(intervals, by = "transition") %>%
  mutate(
    phi_monthly = phi_raw^(1 / delta_months)
  )

# 6) Interval‐specific summary: collapse down, then add CI width & season
interval_summary <- phi_draws %>%
  group_by(transition, from, to, start_date, end_date, delta_days, delta_months) %>%
  summarise(
    median   = median(phi_monthly),
    lo025  = quantile(phi_monthly, 0.025),
    hi975  = quantile(phi_monthly, 0.975),
    .groups = "drop"
  ) %>%
  # now that you have one row per transition, compute:
  mutate(
    cri_width = hi975 - lo025,
    season   = if_else(delta_months > 2, "winter", "summer")
  ) %>%
  arrange(transition)

# 7) Seasonal summaries (winter >2 mo vs. summer ≤2 mo)
season_summary <- phi_draws %>%
  mutate(
    season = if_else(delta_months > 2, "winter", "summer")
  ) %>%
  group_by(season) %>%
  summarise(
    median     = median(phi_monthly),
    lo025    = quantile(phi_monthly, 0.025),
    hi975    = quantile(phi_monthly, 0.975),
    ci_width = hi975 - lo025,
    .groups  = "drop"
  )

# 8) Print or export results
print(interval_summary)
print(season_summary)

# Optionally write tables for your paper:
# write_csv(interval_summary, "interval_survival_summary.csv")
# write_csv(season_summary,   "season_survival_summary.csv")

```

### 2025-05-08 doc/Manuscript.qmd
add figures to results

@@
   echo: false
+bibliography: references.bib
 ---
@@
 
-Mountain yellow-legged frogs (*Rana sierrae* and *Rana muscosa*) have disappeared from \>90% of their historic range due to the virulent fungal pathogen *Batrachochytrium dendrobatidis* (Bd), yet a handful of populations now persist and recover in the pathogen's presence.
-We studied one such recovering population at Mossy Pond, Tahoe National Forest, for five summers (2014–2018) using robust-design capture–mark–recapture methods and analyzed the capture histories with a Bayesian Jolly–Seber model.
+Mountain yellow-legged frogs (*Rana sierrae* and *Rana muscosa*) have disappeared from \>90% of their historic range due to the virulent fungal pathogen *Batrachochytrium dendrobatidis* (Bd).
+Despite the severe impacts of Bd on naive populations, a small fraction of affected populations now persist and recover in the pathogen's presence.
+We studied one such recovering *R.
+sierrae* population at Mossy Pond, Tahoe National Forest, for five summers (2014–2018) using robust-design capture–mark–recapture methods and analyzed capture histories with a Bayesian Jolly–Seber model.
 From 3,631 captures of 657 uniquely marked adults, estimated abundance varied from 117 to 245 individuals, and a total of 629 new adults recruited over the study.
-Monthly survival was higher in winter than in summer (mean and credible interval; winter: 0.97 (xx-xx), summer: 0.8 (xx-xx) and male survival exceeded female survival by 2–7% in each interval.
+Monthly survival was higher in winter than in summer (posterior median and credible interval; winter: 0.97 (0.94-0.98), summer: 0.82 (0.63-0.92) and male survival exceeded female survival by 2–7% in each interval.
 Survey-specific detection probability averaged 0.45.
-Bd was present in 67% of skin-swab samples and infection loads were two–three orders of magnitude below lethal thresholds.
+Bd was present in 67% of skin swab samples and infection loads were two–three orders of magnitude below lethal thresholds.
 Recruitment and abundance declined sharply after severe winters, yet overwinter adult survival showed no snow effect.
 These findings suggest this population is persisting despite ongoing Bd infection, perhaps as a result of some degree of infection resistance.
-.................
-identify this robust northern Sierra Nevada population as a prime donor for reintroduction-based recovery efforts.
+Our results indicate this *R. sierrae* population may serve as a potential donor for reintroduction-based recovery efforts in the northern Sierra Nevada.
 
@@
 
-Amphibians continue to experience global declines, with infectious diseases recognized as primary drivers of biodiversity loss [@luedtke2023a].
-Chytridiomycosis, caused by the fungal pathogen *Batrachochytrium dendrobatidis* (Bd), has driven declines or extinctions of approximately 200 amphibian species globally [@skerratt2007].
-Amphibian responses to Bd infection vary dramatically; some populations show minimal impact, others rapidly go extinct, and a rare few exhibit persistence or even recovery following severe declines [@lips2016].
-The mechanisms supporting persistence or recovery remain poorly understood but likely involve interactions among ecological, evolutionary, immunological, and demographic factors [@scheele2017].
+Amphibians are declining worldwide, with infectious diseases recognized as primary drivers of biodiversity loss [@luedtke2023a].
+Chytridiomycosis, caused by the virulent fungal pathogen *Batrachochytrium dendrobatidis* (Bd), has contributed to the decline of at least 501 amphibian species over the past half-century, including 90 presumed extinctions, representing the greatest documented loss of biodiversity attributable to a disease [@scheele2019].
 
-The mountain yellow-legged (MYL) frog species complex (*Rana sierrae* and *Rana muscosa*), endemic to California's Sierra Nevada and Transverse ranges, provides a particularly well-documented case of amphibian decline and recovery from Bd infection.
+Among amphibian populations susceptible to Bd, disease dynamics often unfold in two phases: an initial epizootic phase marked by rapid population decline and widespread mortality, followed, in some cases, by an enzootic phase where the pathogen persists at lower host densities and reduced infection intensities [@briggs2010].
+These trajectories can end in extinction, long-term persistence at reduced densities, or, in rare cases, substantial population recovery after severe initial declines [@lips2016].
+The ecological, evolutionary, immunological, and demographic processes underlying persistence and recovery remain poorly understood, yet identifying these mechanisms is crucial for amphibian conservation and management [@scheele2017].
+
+The mountain yellow-legged (MYL) frog species complex (*Rana sierrae* and *Rana muscosa*), endemic to California's Sierra Nevada and Transverse ranges, provides a particularly well-documented case of amphibian decline, persistence, and eventual recovery from Bd infection.
 Starting in the 1970s, Bd spread through the Sierra Nevada, causing widespread and rapid extirpations of MYL frog populations [@vredenburg2010; @vredenburg2019].
@@
 In contrast, the population dynamics of *R.
-sierrae* in the northern Sierra Nevada—where habitats are more heterogeneous and populations are smaller and lower in elevation—remain understudied (MYLF ITT 2018) [@brown2019].
+sierrae* in the northern Sierra Nevada—where habitats are more heterogeneous and populations are smaller and lower in elevation—remain understudied [MYLF ITT 2018; @brown2019].
 
-To address this gap, we conducted a five-year capture-mark-recapture (CMR) study of a naturally recovering, Bd-positive *R.
-sierrae* population in Tahoe National Forest.
-Our objectives were to estimate key demographic parameters—abundance, survival, recruitment, and detection probability—to provide insights into the ecological processes and factors that contribute to population viability under endemic Bd infection.
+To address this gap, we conducted a five-year capture-mark-recapture (CMR) study of a naturally recovering, Bd-positive *R. sierrae* population in Tahoe National Forest.
+Our objectives were to estimate key demographic parameters—abundance, survival, recruitment, and detection probability—to provide insights into the ecological processes and factors that contribute to population viability under enzootic Bd infection.
 We also examined the role of climate variability in shaping these demographic parameters, as climate change is increasingly recognized as a primary driver of amphibian declines worldwide, comparable in impact to disease and habitat loss [@luedtke2023a].
-Understanding how climate influences abundance, survival, and recruitment is particularly important for high-elevation amphibian populations living with endemic Bd infection.
+Understanding how climate influences abundance, survival, and recruitment is particularly important for high-elevation amphibian populations living with enzootic Bd infection.
 
@@
 
-The Mossy Pond study area is a 2.6 square km section of the Tahoe National Forest, north of Highway 80 in Nevada County, California.
+The Mossy Pond study area is a 2.6 square kilometer section of the Tahoe National Forest, north of Highway 80 in Nevada County, California.
 It is characterized by lakes, ponds, and streams set on granite benches, ranging from 1,950 m to 2,164 m in elevation.
@@
 
-*R.
-sierrae* is a federally endangered, mid-sized ranid frog that occupies montane waterbodies in California's Sierra Nevada [@u.s.fishandwildlifeservice2014].
-*R.
-sierrae* was first detected in the Mossy Pond area in 1998 **(citation)**, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.
+*R. sierrae* is a federally endangered, mid-sized ranid frog that occupies montane waterbodies in California's Sierra Nevada [@u.s.fishandwildlifeservice2014].
+*R. sierrae* was first detected in the Mossy Pond area in 1998, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001 (CDFW, unpublished data).
 
@@
 
-To assess the Mossy Pond *R.
-sierrae* population's Bd status, we collected skin swabs from 24 adult frogs over three years (2008, 2010, and 2021).
+To assess the Mossy Pond *R. sierrae* population's Bd status, we collected skin swabs from 24 adult frogs over three years (2008, 2010, and 2021).
 The amount of Bd DNA on each swab was quantified using real-time quantitative polymerase chain reaction (qPCR) analysis [@boyle2007].
 We report infection loads as the number of ITS1 (Internal Transcribed Spacer 1) gene copies detected [@longo2013; @joseph2018].
-To compare the Bd loads of Mossy Pond frogs to those of other known recovering *R.
-sierrae* populations, we compared Bd loads from Mossy Pond to those from three known naturally recovering populations in Yosemite National Park [@knapp2024].
+We compared Bd loads from Mossy Pond frogs to those reported from three naturally recovering *R. sierrae* populations in Yosemite National Park [@knapp2024].
 
-### Frog population surveys
+### Capture-Mark-Recapture Surveys
 
@@
 
-Surveys occurred between 0800 and 1800, when *R.
-sierrae* are most active.
-We systematically searched all water bodies in the study area, capturing frogs by hand or dip net for immediate processing.
-We first scanned each captured frog with a BioMark 601 passive integrated transponder (PIT) tag reader (BioMark, Boise, Idaho, USA) to detect if the frog was marked (i.e., a recapture).
-We then used calipers to measure the snout to urostyle (SUL) length of each captured frog, and released frogs \<40 mm SUL (which are considered subadults) without further processing.
-Frogs larger than 40mm (adults) required further processing.
-New adult captures were PIT-tagged dorsally following @mcallister2004.
-Next, we recorded the frog's sex and weighed it inside a tared plastic bag to the nearest 0.1 g using a Pesola spring scale.
+Surveys occurred between 0800 and 1800, when *R. sierrae* are most active.
+We systematically searched all water bodies in the study area, capturing adult frogs (\>40mm snout to urostyle length) by hand or dip net for immediate processing.
+We first scanned each captured frog for a passive integrated transponder (PIT) tag to detect if the frog was already marked (i.e., a recapture).
+Untagged adults were PIT-tagged dorsally following @mcallister2004.
+Next, we recorded the frog's sex and weighed it inside a plastic bag to the nearest 0.1 g using a Pesola spring scale.
 Finally, we collected a GPS point (estimated precision error \~3 m) for each capture using a handheld GPS unit.
@@
 Afterward, if crews captured the same individual on a subsequent day during the same primary period, crews only recorded PIT tag, sex, and location data.
-Over the 5-year period of our study, we recorded 3,631 captures of 657 individual frogs.
 
-### Statistical Analysis
+### Capture-Mark-Recapture Modeling & Statistical Analysis
 
 We estimated survival, detection, recruitment, and adult population size for the Mossy Pond *R.
-sierrae* population using a site‐specific open‐population Jolly–Seber CMR model in a Bayesian framework with R statistical software [@R].
-Our implementation builds on the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), which tracks a superpopulation of M individuals through not‐recruited, alive, or dead states and models detection and demographic processes across primary and secondary periods [@knapp2024].
-We assessed convergence with trace‐plots and Gelman–Rubin (R̂) diagnostics and compared model alternatives using leave‐one‐out cross‐validation.
+sierrae* population using an open‐population Jolly–Seber CMR model implemented in a Bayesian framework with R statistical software [@R].
+We analyzed the CMR data using the mrmr R package [@joseph2019] (https://snarl1.github.io/mrmr/index.html), but modified the package code to add additional functionality, as described below.
+Our implementation builds on the original package, which tracks a superpopulation of M individuals through not‐recruited, alive, or dead states and models detection and demographic processes across primary and secondary periods [@knapp2024].
+
+*We assessed convergence with trace‐plots and Gelman–Rubin (R̂) diagnostics and compared model alternatives using leave‐one‐out cross‐validation.*
 
@@
 For each draw and period we computed the female logit (intercept + period effect), added the male coefficient to get the male logit, transformed both via the logistic function, and recorded their difference.
-We summarize those interval‐specific male–female differences by posterior mean, standard deviation, and 95 % credible interval.
+We summarize those interval‐specific male–female differences by posterior mean, standard deviation, and 95% credible interval.
 
 To capture unmeasured heterogeneity in detection across survey occasions, we introduced survey‐specific random effects on logit‐scale detection.
-This parallels the hierarchical structure used for survival and recruitment in the original mrmr model and acknowledges that factors like temperature, effort, observer, and water clarity can affect detectability even when fixed covariate data are incomplete.
+This parallels the hierarchical structure used for survival and recruitment in the original mrmr model.
 
-As a final analytical step, we used the posterior samples from our fitted CMR model to test for climate–demography links.
-For each posterior draw, we calculated Pearson correlation coefficients between winter severity (percent of average snow-water equivalent remaining on 1 April) and each demographic parameter—monthly survival, recruitment, and adult abundance—thereby generating full posterior distributions of those correlations.
+As a final analytical step, we used the posterior samples from our fitted CMR model to evaluate the relationship between winter severity and adult survival, recruitment, and abundance.
+We calculated Pearson correlation coefficients between winter severity—measured as percent of average snow-water equivalent (SWE) on April 1, the date typically representing peak snowpack—and each demographic parameter, thereby generating full posterior distributions of those correlations.
 To capture potential delayed effects given the frog’s multi-year tadpole overwintering, we repeated these analyses using winter severity at lag 0, 1, 2, and 3 years as well as a 3-year rolling average.
-This approach quantifies both the strength and uncertainty of climate influences on population dynamics while preserving the hierarchical uncertainty structure of our Bayesian framework.
 
@@
 
-We defined winter severity as the percentage of historical average snow-water equivalent (SWE) remaining on April 1 each year (100% indicating average severity).
+We defined yearly winter severity as the percent of April 1 average snow-water equivalent (SWE) in the Mossy Pond study area, because this date typically coincides with peak snowpack in mountainous regions of the western United States [@bohr2001].
 We obtained SWE data from ten regional snow-survey stations (4.5–19.5 km from Mossy Pond) to capture the spatial variability in snowpack present in topographically complex mountainous environments [@revuelto2014].
-Annual SWE values (2011–2018) were interpolated across the study area using inverse-distance weighting in ArcGIS Pro 3.4.3.
-From the resulting continuous SWE raster, we extracted the mean percent SWE within our one-square-mile study boundary.
-All raw SWE measurements were obtained from the California Data Exchange Center (CDEC; http://cdec.water.ca.gov).
+Each station’s April 1 average is its 1991–2020 30-year normal, consistent with World Meteorological Organization standards [@naturalresourceconservationservice2025].
+For clarity, we label each winter by the calendar year in which it ends (e.g., “2011” refers to the 2010/2011 winter).
+Annual SWE values for winters 2011–2018 were interpolated across the study area with inverse-distance weighting in ArcGIS Pro 3.4.3, and we extracted the mean percent-of-average SWE within the one-square-mile study boundary.
+All raw SWE measurements were downloaded from the California Data Exchange Center (https://cdec.water.ca.gov)
 
@@
 
-16 out of 24 samples tested positive for Bd, with infection intensities ranging from 16 to 43,318 ITS1 gene copies.
-These Bd loads are well below the \~600,000 gene copy threshold associated with severe chytridiomycosis and increased MYL frog mortality [@vredenburg2010; @longo2013].
-The infection intensities observed in our study population are typical of those measured in other known naturally recovering *R.
-sierrae* populations [Figure 1; @knapp2024].
+16 out of 24 samples tested positive for Bd, and infection intensities of these Bd-positive swabs ranged from 16 to 43,318 ITS1 gene copies.
+These Bd loads are well below the \~600,000 gene copy threshold associated with severe chytridiomycosis and increased MYL frog mortality [@vredenburg2010; @joseph2018].
+The infection intensities observed in our study population are typical of those measured in other known naturally recovering *R. sierrae* populations [Figure 1; @knapp2024].
 
-![**Figure 1.
-Batrachochytrium dendrobatidis (Bd) infection loads across naturally recovering mountain yellow-legged frog populations.** Box plots show the distribution of Bd infection intensity (ITS1 gene copies) for Mossy Pond and three recovering Rana sierrae populations in Yosemite National Park (identified by ID numbers 70459, 72996, and 70567).
+![***Batrachochytrium dendrobatidis*** **(Bd) infection loads across naturally recovering *Rana sierrae* populations.** Box plots with points show the distribution of Bd infection intensity (ITS1 gene copies) for Mossy Pond and three recovering *R.
+sierrae* populations in Yosemite National Park (identified by ID numbers 70459, 72996, and 70567).
 The dashed red line indicates the approximate threshold (\~600,000 gene copies) associated with severe chytridiomycosis and increased mortality in mountain yellow-legged frogs.
-Note the logarithmic scale on the y-axis.
-All recovering populations maintain Bd loads well below the mortality threshold, with Mossy Pond showing infection intensities comparable to those observed in the Yosemite populations.](images/Bd_load_comparison-01.png)
+The y-axis is on a log10 scale.
+Individuals from recovering populations generally maintain Bd loads below the mortality threshold, with Mossy Pond showing infection intensities comparable to those observed in the Yosemite populations.](images/Bd_load_comparison-03.png)
 
-### Abundance, Survival, Recruitment, and Detection
+### Capture-Mark-Recapture Surveys
 
-*Abundance*
+Over 47 surveys spanning 5 years, we recorded 3,631 captures of 657 unique *R. sierrae* individuals.
+Captures were spread fairly evenly over the 5 years..................Or were highest in.......?
+**Figure**?
 
-We calculated the abundance of the Mossy Pond *R.
-sierrae* population for each of the 16 primary survey periods of the study.
-The lowest abundance was observed in September 2018 (117 individuals, 95% credible interval (95% CrI): 103-132) and a maximum of 245 individuals (95% CrI: 237-256) was observed during the fifth primary period, in July 2015.
+### Capture-Mark-Recapture Modeling
+
 Across all primary periods, the mean abundance of the population was 168 individuals.
+The lowest abundance was observed in September 2018 (117 individuals, 95% CrI: 103-132) and a maximum of 245 individuals (95% CrI: 237-256) was observed in July 2015 (**Figure Xa**).
+A total of 629 frogs recruited into the adult population over the course of the study.
+Annual recruitment ranged from 60 individuals (95% CrI: 45-78) in 2018 to 166 individuals (95% CrI: 143-188) in 2015 (**Figure Xb)**.
+The average detection probability over 47 individual surveys was 0.45, ranging from 0.26 to 0.61.
 
-*Recruitment*
+![Estimated adult abundance (N) of the Mossy Pond *Rana sierrae* population across 16 primary survey periods (June 2014–September 2018).
+Points show posterior median estimates; shaded ribbons represent 95 % Bayesian credible intervals.](images/N_plot.png)
 
-A total of 629 frogs recruited into the adult population over the course of the study.
-Annual recruitment ranged from 60 individuals (95% CrI: 45-78) in 2018 to 166 individuals (95% CrI: 143-188) in 2015.
+![Estimated annual recruitment (B) in the Mossy Pond *Rana sierrae* population from 2014 to 2018.
+For each year, recruitment values are calculated by summing posterior estimates of recruitment across all primary periods within that year, preserving posterior covariance.
+Points represent posterior median estimates of total annual recruitment; shaded ribbons show 95 % Bayesian credible intervals.](images/B_plot.png)
 
-*Survival*
+Monthly survival during the summer was markedly lower and more variable than during winter (**Figure X**).
+Across 11 summer intervals, median monthly survival probability was 0.82 (95% CrI 0.63–0.92), whereas across 4 overwinter intervals, mean monthly survival was 0.97 (95% CrI 0.94–0.98).
+In every interval, mean male survival exceeded mean female survival, with differences ranging from 0.015 (95% CrI 0.004-0.029) to 0.069 (Figure 2).
+The 95% credible interval for each period does not contain zero, indicating consistently higher male survival across both summer and overwinter intervals.
+These results suggest a modest and consistent male survival advantage throughout the study.
 
-Monthly survival during the active (summer) season was markedly lower and more variable than during overwintering.
-Across 11 summer intervals, mean monthly survival was 0.803 (range: 0.673–0.904), whereas across 4 overwinter intervals, mean monthly survival was 0.966 (range: 0.945–0.974).
+![Standardized monthly survival probabilities of of Rana sierrae at Mossy Pond, Tahoe National Forest, across 15 primary survey period transitions (June 2014 – August 2018).
+Points represent posterior medians; vertical shading shows 95% Bayesian credible intervals.
+Each point denotes survival for the interval beginning on the labelled primary period and ending at the next primary period.
+Red points are summer (active-season) transitions; blue points are overwinter transitions](images/phi_plot-03.png)
 
-In every interval, mean male survival exceeded mean female survival, with differences ranging from 0.015 (95% CrI 0.004-0.029) between the first and second primary periods to 0.069 between the sixth and seventh primary periods (Figure 2).
-The 95% credible interval for each period lay entirely above zero, indicating consistently higher male survival across both summer and overwinter intervals.
-These results suggest a modest but statistically robust male survival advantage throughout the study.
+![Sex-specific survival differences in the Mossy Pond Rana sierrae population, 2014-2018.
+Points represent the mean difference in survival probability between males and females (M-F) across 15 primary period transitions, colored by season (red: summer; blue: winter).
+Vertical lines show 95% Bayesian credible intervals.
+All credible intervals contain only positive values, indicating consistently higher male survival across all primary period transitions.](images/survival_sex_difference_plot-01.png)
 
-![**Figure 2.
-Sex-specific survival differences in the Mossy Pond *Rana sierrae* population, 2014-2018.** Points represent the mean difference in monthly survival probability between males and females (M-F) across 15 primary period transitions.
-Vertical lines show 95% credible intervals.
-All intervals are entirely above zero, indicating consistently higher male survival across both summer and overwinter intervals throughout the study period.
-Overwinter periods occur during the 3-4, 6-7, 9-10, and 12-13 interval transitions.](images/survival_by_sex.png)
+### Winter severity and its effects on population dynamics
 
-*Detection*
+Our study period coincided with high variation in winter severity at the Mossy Pond study area, with April 1 SWE ranging from 3.5% of historical average in 2015 to 150% in 2017 (**figure X**).
 
-For each survey (secondary period; n=47) of the study, we calculated the probability that an individual frog that is alive and available for detection is actually detected.
-Detection probability was lowest (median 0.26, 95% CrI: 0.2-0.31) during the sixteenth survey (09 September, 2015) and highest (median 0.61, 95% CrI: 0.54-0.67) during the forty-third survey (07 August, 2018).
-The average detection probability over 47 individual surveys was 0.45.
+![Winter severity index for the Sierra Nevada region, 2010–2011 through 2017–2018 winters, calculated as the percentage of average April 1 snow water equivalent (SWE).
+Bars show the relative snowpack severity for each winter season, with values greater than 100% indicating above-average snowpack and values below 100% indicating below-average conditions.](images/winter_severity_plot.png)
 
-### Winter severity effects on population dynamics
-
-Our study period coincided with high variation in winter severity at the Mossy Pond study area, with April 1 SWE ranging from 3.5% of historical average in 2015 to 150% in 2017.
 Our analyses reveal a strong negative correlation between winter severity lagged by one year and both the number of frogs recruited into the adult population (median correlation coefficient: -0.92; 95% CrI: –0.79 to -0.98; Figure 3) and the abundance of the population (median correlation coefficient: -0.84, 95% CrI: -0.8 to -0.89).
 Winter severity showed no notable relationship with overwinter survival probabilities.
-
-![**Figure 3.
-Relationship between adult Rana sierrae recruitment and prior year winter severity at Mossy Pond, 2014-2018.** The solid line represents the number of frogs recruited into the adult population each year (left y-axis), while the dashed line shows winter severity as percent of average snow-water equivalent (SWE) measured on April 1 of the prior year (right y-axis).
-Note the inverse relationship, with high recruitment following mild winters (low SWE) and reduced recruitment following severe winters (high SWE).](images/WinterSeverity_Recruitment-02.png)
 
@@
 
-#### **Bd infection dynamics and evidence for resistance**
+#### **Patterns of Bd infection**
 
-The Mossy Pond frogs carried Bd at moderate prevalence (67%) but at infection intensities that were two to three orders of magnitude lower than levels associated with increased mortality in MYL frogs.
-Such low loads mirror those reported for other naturally recovering populations in Yosemite National Park and stand in stark contrast to the uniformly high loads (\>600,000 ITS1 copies) and near-certain mortality that typify collapsing populations [@vredenburg2010].
-Coupled with the population’s stable abundance over the study period, these patterns are consistent with the idea that some degree of Bd resistance has arisen in this population.
+The moderate prevalence and low intensity of Bd infection in the Mossy Pond population is consistent with patterns reported for other naturally recovering populations in Yosemite National Park.
+This contrasts with the uniformly high loads (\>600,000 ITS1 copies) and associated high mortality that typify epizootic populations [@vredenburg2010].
+Coupled with the population’s stable abundance over the study period, these patterns are consistent with the idea that Mossy Pond frogs have developed some degree of Bd resistance.
 While such dynamics could reflect an evolutionary rescue process in which genotypes conferring lower susceptibility have increased in frequency, additional genetic and experimental evidence would be required to confirm that mechanism.
@@
 
-The pronounced male survival advantage across all intervals (Δφ = 0.02–0.07) contrasts with Fellers et al.
-(2013), who reported nearly identical annual survival of male and female R.
+The pronounced male survival advantage across all intervals (Δφ = 0.02–0.07) contrasts with Fellers et al. (2013), who reported nearly identical annual survival of male and female R.
 sierrae at Summit Meadow, with females sometimes faring slightly better.
@@
 montanus* show that females travel farther than males to locate suitable oviposition sites, increasing exposure to predators and desiccating microhabitats [@honeycutt2019].
-Together, these reproductive and behavioural costs may magnify female summer mortality in the Mossy Pond *R.
-sierrae* population, driving the male-biased survival pattern we observed.
+Together, these reproductive and behavioural costs may magnify female summer mortality in the Mossy Pond *R. sierrae* population, driving the male-biased survival pattern we observed.
 
@@
 
-Our results place the Mossy Pond frogs squarely within the “resistant-donor” framework that @knapp2024 used to successfully facilitate landscape-scale *R.
-sierrae* recovery in Yosemite National Park.
+Our results place the Mossy Pond frogs squarely within the “resistant-donor” framework that @knapp2024 used to successfully facilitate landscape-scale *R. sierrae* recovery in Yosemite National Park.
 In that 15-year experiment, reintroductions sourced from naturally recovering, Bd-resistant *R.
